### 2017-04-01 paper/main.tex
Refactored InputReaders.
WARNING, everything but AlignedRaw hasn't been implemented.

\documentclass[times, utf8, seminar]{fer}
\usepackage{booktabs}

\usepackage{mathtools, amsmath,amsfonts,amssymb, amsthm}
\usepackage{acronym}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{float}
\usepackage{graphicx}
\graphicspath{ {Figures/} }

\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\begin{document}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

\voditelj{Mile Sikić}
\title{Sustav potpore odlučivanju za predviđanje i optimalnu nabavu
}
\engtitle{A decision support system for forecasting and optimal procurement}
\author{Neven Miculinić}

\maketitle
\tableofcontents

\begin{sazetak}
    Optimalna nabava u većini industrija uključuje predviđanje dviju veliničina: cijene sirovina i potražnje. Cilj ovoga rada je ujediniti predviđanje u model planiranja proizvodnje sa ciljem minimiziranja cjelokupne nabave, troškova skladištenja i proizvodnje uz zadovoljenje potražnje. Sustav potpore odlučivanju treba omogućiti donositelju odluke integraciju kvalitativne nestrukturiranje informacije kroz jednostavno sučelje, odabir scenarija i profinjenje rješenja.

\kljucnerijeci{planiranje nabave, stohastička potražnja, optimalna količina nabave, Model dinamičke ekonomske količine nabave}
\end{sazetak}

\begin{abstract}
Optimal procurement in most industries involves forecasting of two
quantities: prices of raw materials and customer's demand.  The aim of
this work is to integrate forecasts into production planning models,
with the aim of minimizing overall procurement, holding and production
costs under demand satisfaction constraints.  The decision support
system should allow the decision maker to integrate qualitative,
unstructured information through a simple interface, for scenario
selection and solution refinement.
\keywords{procurement planning, dynamic lot sizing model, stochastic demands}
\end{abstract}

\chapter{Introduction}
Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
\chapter{Related work}
Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
\chapter{Model description}
Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
\chapter{Experimental results}
Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
\chapter{Conclusion and further work}
Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.\cite{deepnano}
% \bibliographystyle{fer}
\bibliographystyle{abbrv}
\bibliography{refs}

\end{document}

### 2017-04-01 src/input_readers.py

@@ -4,12 +4,109 @@ from random import shuffle
 
 import numpy as np
 import util
+import h5py
 
 root_dir_map = {
     'karla': '/hgst8TB/fjurisic/ecoli',
     'protagonist': '/home/lpp/Downloads/minion'
 }
 root_dir_default = root_dir_map.get(socket.gethostname(), '/data')
+
+
+class InputReader():
+    def input_fn(self):
+        raise NotImplemented
+
+    def fn_args(self, model):
+        raise NotImplemented
+
+
+class AlignedRaw(InputReader):
+
+    @staticmethod
+    def preprocessSignal(signal):
+        return (signal - 646.11133) / 75.673653
+
+    def read_fast5_raw_ref(self, fast5_path, ref_path, block_size_x, block_size_y, num_blocks, warn_if_short=False):
+        num_blocks += 1
+        with h5py.File(fast5_path, 'r') as h5, open(ref_path, 'r') as ref_file:
+            reads = h5['Analyses/EventDetection_000/Reads']
+            target_read = list(reads.keys())[0]
+            events = np.array(reads[target_read + '/Events'])
+            start_time = events['start'][0]
+            start_pad = int(start_time - h5['Raw/Reads/' + target_read].attrs['start_time'])
+
+            basecalled_events = h5['/Analyses/Basecall_1D_000/BaseCalled_template/Events']
+            basecalled = np.array(basecalled_events.value[['mean', 'stdv', 'model_state', 'move', 'start', 'length']])
+
+            signal = h5['Raw/Reads/' + target_read]['Signal']
+            signal_len = h5['Raw/Reads/' + target_read].attrs['duration'] - start_pad
+
+            x = np.zeros([block_size_x * num_blocks, 1], dtype=np.float32)
+            x_len = min(signal_len, block_size_x * num_blocks)
+            x[:x_len, 0] = signal[start_pad:start_pad + x_len]
+
+            np.testing.assert_allclose(len(signal), start_pad + np.sum(events['length']))
+
+            events_len = np.zeros([num_blocks], dtype=np.int32)
+
+            bcall_idx = 0
+            prev, curr_sec = "N", 0
+            for e in events:
+                if (e['start'] - start_time) // block_size_x > curr_sec:
+                    prev, curr_sec = "N", (e['start'] - start_time) // block_size_x
+                if curr_sec >= num_blocks:
+                    break
+
+                if bcall_idx < basecalled.shape[0]:
+                    b = basecalled[bcall_idx]
+
[…]

### 2017-04-01 src/model_small.py

@@ -62,7 +62,6 @@ def model_setup_params(hyper):
         g=tf.Graph(),
         block_size_x=8 * 3 * 50 // 2,
         block_size_y=80,
-        in_data="ALIGNED_RAW",
         num_blocks=1,
         batch_size=32,
         max_reach=8 * 20,  # 160

### 2017-04-01 src/model_utils.py

@@ -38,7 +38,7 @@ def default_lr_fn(global_step):
 
 
 class Model():
-    def __init__(self, g, num_blocks, batch_size, max_reach, model_fn, block_size=None, block_size_x=None, block_size_y=None, log_dir=None, run_id=None, overwrite=False, reuse=False, queue_cap=None, shrink_factor=1, test_queue_cap=None, in_data="EVENTS", lr_fn=None, dtype=tf.float32, hyper={}):
+    def __init__(self, g, num_blocks, batch_size, max_reach, model_fn, block_size=None, block_size_x=None, block_size_y=None, log_dir=None, run_id=None, overwrite=False, reuse=False, queue_cap=None, shrink_factor=1, test_queue_cap=None, in_data=input_readers.AlignedRaw(), lr_fn=None, dtype=tf.float32, hyper={}):
         """
             Args:
                 max_reach: int, size of contextual window for convolutions etc.
@@ -46,11 +46,10 @@ class Model():
                 hyper: dictionary of hyperparameters passed to the model function as **hyper
         """
 
-        if in_data not in ["RAW", "EVENTS", "ALIGNED_RAW", "ALIGNED_EVENTS"]:
-            raise ValueError("in_data must be one of two types")
+        if isinstance(in_data, input_readers.InputReader.__class__):
+            raise ValueError("in_data must instance of InputReader")
 
         self.in_data = in_data
-        self.data_in_dim = 1 if in_data == "RAW" or in_data == "ALIGNED_RAW" else 3
         self.lr_fn = lr_fn or default_lr_fn
 
         if overwrite and reuse:
@@ -244,7 +243,7 @@ class Model():
         with tf.variable_scope("input"):
             input_vars = [
                 tf.get_variable("X",
-                                shape=[self.batch_size, self.block_size_x * self.num_blocks, self.data_in_dim],
+                                shape=[self.batch_size, self.block_size_x * self.num_blocks, self.in_data.in_dim],
                                 initializer=tf.zeros_initializer(),
                                 dtype=self.dtype,
                                 trainable=False
@@ -327,9 +326,9 @@ class Model():
                 ]
                 padding = tf.convert_to_tensor(padding)
                 net = tf.pad(net, padding)
-                net.set_shape([None, 2 * self.max_reach + self.block_size_x, self.data_in_dim])
+                net.set_shape([None, 2 * self.max_reach + self.block_size_x, self.in_data.in_dim])
                 print(net.get_shape())
[…]

### 2017-04-01 src/util.py

@@ -328,69 +328,6 @@ def read_fast5_ref(fast5_path, ref_path, block_size, num_blocks, warn_if_short=F
 
     return x, x_len, y, y_len
 
-
-def read_fast5_raw_ref(fast5_path, ref_path, block_size_x, block_size_y, num_blocks, warn_if_short=False):
-
-    num_blocks += 1
-    with h5py.File(fast5_path, 'r') as h5, open(ref_path, 'r') as ref_file:
-        reads = h5['Analyses/EventDetection_000/Reads']
-        target_read = list(reads.keys())[0]
-        events = np.array(reads[target_read + '/Events'])
-        start_time = events['start'][0]
-        start_pad = int(start_time - h5['Raw/Reads/' + target_read].attrs['start_time'])
-
-        basecalled_events = h5['/Analyses/Basecall_1D_000/BaseCalled_template/Events']
-        basecalled = np.array(basecalled_events.value[['mean', 'stdv', 'model_state', 'move', 'start', 'length']])
-
-        signal = h5['Raw/Reads/' + target_read]['Signal']
-        signal_len = h5['Raw/Reads/' + target_read].attrs['duration'] - start_pad
-
-        x = np.zeros([block_size_x * num_blocks, 1], dtype=np.float32)
-        x_len = min(signal_len, block_size_x * num_blocks)
-        x[:x_len, 0] = signal[start_pad:start_pad + x_len]
-
-        np.testing.assert_allclose(len(signal), start_pad + np.sum(events['length']))
-
-        events_len = np.zeros([num_blocks], dtype=np.int32)
-
-        bcall_idx = 0
-        prev, curr_sec = "N", 0
-        for e in events:
-            if (e['start'] - start_time) // block_size_x > curr_sec:
-                prev, curr_sec = "N", (e['start'] - start_time) // block_size_x
-            if curr_sec >= num_blocks:
-                break
-
-            if bcall_idx < basecalled.shape[0]:
-                b = basecalled[bcall_idx]
-
-                if b[0] == e[2] and b[1] == e[3]:  # mean == mean and stdv == stdv
-                    added_bases = 0
-
-                    if bcall_idx == 0:
-                        added_bases = 5
-                        assert len(list(b[2].decode("ASCII"))) == 5
-
-                    bcall_idx += 1
-                    assert 0 <= b[3] <= 2
-                    added_bases += b[3]
-                    events_len[curr_sec] += added_bases
-
[…]

### 2017-04-01 src/residual_deep.py
Changed to better default params (from testing)

@@ -86,9 +86,9 @@ params = [
 ]
 
 default_params = {
-    'initial_lr': 1e-4,
-    'decay_factor': 0.1,
-    'num_layers': 20,
+    'initial_lr': 0.0014691078293764034,
+    'decay_factor': 0.001341733349277896,
+    'num_layers': 17,
     'num_sub_layers': 2
 }
 

### 2017-04-01 src/model_utils.py
Small model_utils refactor

@@ -87,10 +87,13 @@ class Model():
         self.train_queue_cap = queue_cap or 5 * self.batch_size
         self.test_queue_cap = test_queue_cap or self.train_queue_cap
         self.dtype = dtype
+        self.g = g
+        self.__setup_graph(model_fn, hyper)
 
-        with g.as_default():
+    def __setup_graph(self, model_fn, hyper):
+        with self.g.as_default():
             self.training_mode = get_training_mode()
-            self.batch_size_var = tf.placeholder_with_default(tf.convert_to_tensor(batch_size, dtype=tf.int32), [])
+            self.batch_size_var = tf.placeholder_with_default(tf.convert_to_tensor(self.batch_size, dtype=tf.int32), [])
             net = self.__create_train_input_objects()
 
             self.block_size_x_tensor = tf.placeholder_with_default(self.block_size_x, [])
@@ -103,7 +106,7 @@ class Model():
                     block_size=self.block_size_x_tensor,
                     out_classes=9,
                     batch_size=self.batch_size_var,
-                    dtype=dtype,
+                    dtype=self.dtype,
                     **hyper
                 )
 
@@ -159,7 +162,7 @@ class Model():
 
             self.grad_summ = tf.summary.merge(add_gradients_summary(self.grads))
             self.activation_summ = tf.summary.merge(
-                add_activations_summary(g.get_collection("activations")))
+                add_activations_summary(self.g.get_collection("activations")))
 
             self.train_summ = tf.summary.merge([
                 self.train_queue_size,
@@ -170,7 +173,6 @@ class Model():
                 keep_checkpoint_every_n_hours=1,
             )
 
-        self.g = g
         self.trace_level = tf.RunOptions.NO_TRACE
 
         self.batch_time = 0
@@ -178,6 +180,7 @@ class Model():
 
         self.bbt = 0
         self.bbt_clock = perf_counter()
+
 
     def __dedup_output(self, sparse_tensor):
         sol_values = sparse_tensor.values - 4 * tf.to_int32(sparse_tensor.values >= 4)

### 2017-04-01 src/model_utils.py
Continuing refactoring model_utils

@@ -38,12 +38,13 @@ def default_lr_fn(global_step):
 
 
 class Model():
-    def __init__(self, g, num_blocks, batch_size, max_reach, model_fn, block_size=None, block_size_x=None, block_size_y=None, log_dir=None, run_id=None, overwrite=False, reuse=False, queue_cap=None, shrink_factor=1, test_queue_cap=None, in_data=input_readers.AlignedRaw(), lr_fn=None, dtype=tf.float32, hyper={}):
+    def __init__(self, g, num_blocks, batch_size, max_reach, model_fn, block_size=None, block_size_x=None, block_size_y=None, log_dir=None, run_id=None, overwrite=False, reuse=False, queue_cap=None, shrink_factor=1, test_queue_cap=None, in_data=input_readers.AlignedRaw(), lr_fn=None, dtype=tf.float32, hyper={}, clip_grad=2.0):
         """
             Args:
                 max_reach: int, size of contextual window for convolutions etc.
                 model_fn: function accepting (batch_size, 2*max_reach + block_size, 3) -> (block_size, batch_size, out_classes). Notice shift to time major as well as reduction in time dimension.
                 hyper: dictionary of hyperparameters passed to the model function as **hyper
+                clip_grad: Maximum gradient_clipping value, None to disable
         """
 
         if isinstance(in_data, input_readers.InputReader.__class__):
@@ -88,108 +89,124 @@ class Model():
         self.test_queue_cap = test_queue_cap or self.train_queue_cap
         self.dtype = dtype
         self.g = g
+        self.clip_grad = clip_grad
         self.__setup_graph(model_fn, hyper)
 
     def __setup_graph(self, model_fn, hyper):
+        self.__setup_graph_pre()
         with self.g.as_default():
-            self.training_mode = get_training_mode()
-            self.batch_size_var = tf.placeholder_with_default(tf.convert_to_tensor(self.batch_size, dtype=tf.int32), [])
-            net = self.__create_train_input_objects()
+            data = self.__setup_logits(model_fn, hyper)
+            self.logits = data['logits']
+            self.init_state = data['init_state']
+            self.final_state = data['final_state']
+            self.reg = data.get('reg', tf.constant(0))
+            self.loss = self.__setup_loss(self.logits)
 
-            self.block_size_x_tensor = tf.placeholder_with_default(self.block_size_x, [])
-
-            with tf.variable_scope("model"):
-                data = model_fn(
-                    net,
-                    self.X_batch_len,
[…]

### 2017-04-01 src/model_utils.py
Refactored model_utils
Removed block_size, use _x and _y variant
Fixed optional gradient_clipping

@@ -38,7 +38,7 @@ def default_lr_fn(global_step):
 
 
 class Model():
-    def __init__(self, g, num_blocks, batch_size, max_reach, model_fn, block_size=None, block_size_x=None, block_size_y=None, log_dir=None, run_id=None, overwrite=False, reuse=False, queue_cap=None, shrink_factor=1, test_queue_cap=None, in_data=input_readers.AlignedRaw(), lr_fn=None, dtype=tf.float32, hyper={}, clip_grad=2.0):
+    def __init__(self, g, num_blocks, batch_size, max_reach, model_fn, block_size_x, block_size_y, lr_fn=default_lr_fn, log_dir=None, run_id=None, overwrite=False, reuse=False, queue_cap=None, shrink_factor=1, test_queue_cap=None, in_data=input_readers.AlignedRaw(), dtype=tf.float32, hyper={}, clip_grad=2.0):
         """
             Args:
                 max_reach: int, size of contextual window for convolutions etc.
@@ -76,8 +76,8 @@ class Model():
         del values
         del valargs
 
-        self.block_size_y = block_size_y or block_size
-        self.block_size_x = block_size_x or block_size
+        self.block_size_y = block_size_y
+        self.block_size_x = block_size_x
         self.shrink_factor = shrink_factor
         if self.block_size_x % shrink_factor != 0:
             raise ValueError("shrink factor need to divide block_size_x")
@@ -194,8 +194,9 @@ class Model():
         """
         optimizer = tf.train.AdamOptimizer(self.lr)
         self.grads = optimizer.compute_gradients(self.loss)
-        with tf.name_scope("gradient_clipping"):
-            self.grads = [(tf.clip_by_value(grad, -2., 2.), var) for grad, var in self.grads]
+        if self.clip_grad is not None:
+            with tf.name_scope("gradient_clipping"):
+                self.grads = [(tf.clip_by_value(grad, -2., 2.), var) for grad, var in self.grads]
 
         self.train_op = optimizer.apply_gradients(self.grads)
 
@@ -759,4 +760,4 @@ class Model():
 
 class TeacherStudentModel(Model):
     def __init__(self, **kwargs):
-        pass
+        super().__init__(self, **kwargs)

### 2017-04-01 src/eval.py
Refactored load_model
Moved load_model_parms from eval to model_utils

@@ -1,25 +1,12 @@
 import os
 from dotenv import load_dotenv, find_dotenv
 import model_utils
-import json
 import argparse
-import importlib
 load_dotenv(find_dotenv())
 
 
 def load_model(module_name, model_dir):
-    model_module = importlib.import_module(module_name)
-    model_dir = os.path.abspath(model_dir)
-    with open(os.path.join(model_dir, 'model_hyperparams.json'), 'r') as f:
-        hyper = json.load(f)
-
-    params = model_module.model_setup_params(hyper)
-    print(params, type(params))
-    params['reuse'] = True
-    params['overwrite'] = False
-    params['log_dir'] = model_dir
-    params['run_id'] = model_dir.split('/')[-1]
-    return model_utils.Model(**params)
+    return model_utils.Model(**model_utils.load_model_parms(module_name, model_dir))
 
 
 def eval_model():

### 2017-04-01 src/model_utils.py

@@ -21,6 +21,7 @@ from slacker_log_handler import SlackerLogHandler
 from edlib import Edlib
 import inspect
 import json
+import importlib
 
 
 # UGLY UGLY HACK!
@@ -35,6 +36,21 @@ logging.basicConfig(level=logging.DEBUG, format=log_fmt)
 
 def default_lr_fn(global_step):
     return tf.train.exponential_decay(1e-3, global_step, 100000, 0.01)
+
+
+def load_model_parms(module_name, model_dir):
+    model_module = importlib.import_module(module_name)
+    model_dir = os.path.abspath(model_dir)
+    with open(os.path.join(model_dir, 'model_hyperparams.json'), 'r') as f:
+        hyper = json.load(f)
+
+    params = model_module.model_setup_params(hyper)
+    print(params, type(params))
+    params['reuse'] = True
+    params['overwrite'] = False
+    params['log_dir'] = model_dir
+    params['run_id'] = model_dir.split('/')[-1]
+    return params
 
 
 class Model():

### 2017-04-01 src/residual_deep.py
Adapted residual_deep to new API

@@ -1,7 +1,6 @@
 import tensorflow as tf
 from tflearn.layers.conv import max_pool_1d, conv_1d
-import os
-import tflearn
+import input_readers
 from tflearn.layers.normalization import batch_normalization
 from dotenv import load_dotenv, find_dotenv
 from tflearn.initializations import variance_scaling_initializer
@@ -62,7 +61,7 @@ def model_setup_params(hyper):
         g=tf.Graph(),
         block_size_x=8 * 3 * 600 // 2,
         block_size_y=630,
-        in_data="ALIGNED_RAW",
+        in_data=input_readers.AlignedRaw(),
         num_blocks=3,
         batch_size=16,
         max_reach=8 * 20,  # 240

### 2017-04-01 src/model_utils.py
Fixed inheritance bug

@@ -72,13 +72,11 @@ class Model():
         if overwrite and reuse:
             raise ValueError("Cannot overwrite and reuse logdit and checkpoints")
 
-        self.__handle_logdir(log_dir, run_id, overwrite, reuse)
+        self._handle_logdir(log_dir, run_id, overwrite, reuse)
 
         frame = inspect.currentframe()
         args, _, _, values = inspect.getargvalues(frame)
         valargs = {arg: values[arg] for arg in args}
-        del valargs['g']
-        del valargs['self']
 
         for k in sorted(valargs.keys()):
             self.logger.info("%-20s: %7s" % (k, str(valargs[k])))
@@ -106,41 +104,40 @@ class Model():
         self.dtype = dtype
         self.g = g
         self.clip_grad = clip_grad
-        self.__setup_graph(model_fn, hyper)
+        self._setup_graph(model_fn, hyper)
 
-    def __setup_graph(self, model_fn, hyper):
-        self.__setup_graph_pre()
+    def _setup_graph(self, model_fn, hyper):
+        self._setup_graph_pre()
         with self.g.as_default():
-            data = self.__setup_logits(model_fn, hyper)
+            data = self._setup_logits(model_fn, hyper)
             self.logits = data['logits']
             self.init_state = data['init_state']
             self.final_state = data['final_state']
             self.reg = data.get('reg', tf.constant(0))
-            self.loss = self.__setup_loss(self.logits)
+            self.loss = self._setup_loss(self.logits)
 
-            self.__setup_prediction()
-            self.__setup_train()
+            self._setup_prediction()
+            self._setup_train()
 
             self.saver = tf.train.Saver(
                 keep_checkpoint_every_n_hours=1,
             )
 
-    def __setup_graph_pre(self):
+    def _setup_graph_pre(self):
         """
             Sets up everything before requirnig model_fn parametirzing model.
         """
-
         with self.g.as_default():
             self.training_mode = get_training_mode()
             self.batch_size_var = tf.placeholder_with_default(tf.convert_to_tensor(self.batch_size, dtype=tf.int32), [])
-            self.__create_train_input_objects()
+            self._create_train_input_objects()
             self.block_size_x_tensor = tf.placeholder_with_default(self.block_size_x, [])
 
[…]

### 2017-04-01 src/model_utils.py
Restricted training only to model variables

@@ -207,7 +207,7 @@ class Model():
             Sets up rest of training. Assumes self.lr and self.loss are defined
         """
         optimizer = tf.train.AdamOptimizer(self.lr)
-        self.grads = optimizer.compute_gradients(self.loss)
+        self.grads = optimizer.compute_gradients(self.loss, self.g.get_collection(tf.GraphKeys.TRAINABLE_VARIABLES, scope='model'))
         if self.clip_grad is not None:
             with tf.name_scope("gradient_clipping"):
                 self.grads = [(tf.clip_by_value(grad, -2., 2.), var) for grad, var in self.grads]

### 2017-04-01 src/model_utils.py
Implemented Teacher/Student network

@@ -118,10 +118,12 @@ class Model():
 
             self._setup_prediction()
             self._setup_train()
+            self._setup_saver()
 
-            self.saver = tf.train.Saver(
-                keep_checkpoint_every_n_hours=1,
-            )
+    def _setup_saver(self):
+        self.saver = tf.train.Saver(
+            keep_checkpoint_every_n_hours=1,
+        )
 
     def _setup_graph_pre(self):
         """
@@ -748,7 +750,7 @@ class Model():
             self.save()
             self.logger.info("Running final validation run")
             return self.run_validation_full(final_val_samples)
-
+            # self.print_k()  # TODO REMOVE!!!
         except Exception as ex:
             if not isinstance(ex, KeyboardInterrupt):
                 self.logger.error("Error happened", exc_info=1)
@@ -787,3 +789,26 @@ class TeacherStudentModel(Model):
         with tf.name_scope("loss"):
             loss = tf.square(logits - alt_logits['logits'])
             return tf.reduce_mean(loss)
+
+    def _setup_saver(self):
+        super()._setup_saver()
+        mapping = {'/'.join(v.op.name.split('/')[1:]): v for v in self.g.get_collection(tf.GraphKeys.TRAINABLE_VARIABLES, 'teacher')
+        }
+        self.teacher_saver = tf.train.Saver(
+            mapping
+        )
+
+    def init_session(self, *args, **kwargs):
+        super().init_session(*args, **kwargs)
+        self.print_k()
+        checkpoint = tf.train.latest_checkpoint(self.teacher_param['log_dir'])
+        if checkpoint is None:
+            raise ValueError("Teacher checkpoint must exist!")
+        self.logger.info("Loading teacher weights from %s", checkpoint)
+        self.teacher_saver.restore(self.sess, checkpoint)
+        print("AFTER")
+        self.print_k()
+
+    def print_k(self):
+        for v in self.g.get_collection(tf.GraphKeys.TRAINABLE_VARIABLES, 'teacher'):
+            print(v.op.name, self.sess.run(v).ravel()[:5])

### 2017-04-01 src/model_small_student.py
Implemented teacher/student

@@ -1,13 +1,12 @@
 import tensorflow as tf
 from tflearn.layers.conv import max_pool_1d, conv_1d
-import os
-import tflearn
 from tflearn.layers.normalization import batch_normalization
 from dotenv import load_dotenv, find_dotenv
 from tflearn.initializations import variance_scaling_initializer
-from train import sigopt_runner
 from ops import central_cut
 from util import sigopt_double
+import model_utils
+import os
 load_dotenv(find_dotenv())
 
 
@@ -66,11 +65,12 @@ def model_setup_params(hyper):
         batch_size=32,
         max_reach=8 * 20,  # 160
         queue_cap=300,
-        overwrite=False,
+        overwrite=True,
         reuse=False,
         shrink_factor=8,
         dtype=tf.float32,
         model_fn=model_fn,
+        run_id="TEST",
         lr_fn=lambda global_step: tf.train.exponential_decay(
             hyper['initial_lr'], global_step, 100000, hyper['decay_factor']),
         hyper=hyper,
@@ -83,10 +83,17 @@ params = [
 ]
 
 default_params = {
-    'initial_lr': 1e-3,
+    'initial_lr': 1e-4,
     'decay_factor': 0.1,
 }
 
 
 if __name__ == "__main__":
-    sigopt_runner(__file__[:-3].split('/')[-1])
+    print(model_setup_params(default_params))
+    model = model_utils.TeacherStudentModel(
+        'model_small',
+        os.path.join(model_utils.repo_root, 'log', 'protagonist', 'model_small_18071_9943114'),
+        **model_setup_params(default_params)
+    )
+
+    model.simple_managed_train_model(3000)

### 2017-04-01 src/model_utils.py

@@ -420,7 +420,7 @@ class Model():
     def get_global_step(self):
         return self.sess.run(self.global_step)
 
-    def train_minibatch(self, log_every=50, trace_every=10000):
+    def train_minibatch(self, log_every=20, trace_every=10000):
         """
             Trains minibatch and performs all required operations
 
@@ -449,15 +449,15 @@ class Model():
         loss = np.sum(vals[1]).item()
         reg_loss = np.sum(vals[2]).item()
 
-        if iter_step % log_every == 0 or True:
-            self.train_writer.add_summary(tf.Summary(value=[
-                tf.Summary.Value(tag="train/loss", simple_value=loss),
-                tf.Summary.Value(tag="train/reg_loss", simple_value=reg_loss),
-                tf.Summary.Value(tag="input/batch_time", simple_value=self.batch_time),
-                tf.Summary.Value(tag="input/dequeue_time", simple_value=self.dequeue_time),
-                tf.Summary.Value(tag="input/between_batch_time", simple_value=self.bbt),
-            ]), global_step=iter_step)
+        self.train_writer.add_summary(tf.Summary(value=[
+            tf.Summary.Value(tag="train/loss", simple_value=loss),
+            tf.Summary.Value(tag="train/reg_loss", simple_value=reg_loss),
+            tf.Summary.Value(tag="input/batch_time", simple_value=self.batch_time),
+            tf.Summary.Value(tag="input/dequeue_time", simple_value=self.dequeue_time),
+            tf.Summary.Value(tag="input/between_batch_time", simple_value=self.bbt),
+        ]), global_step=iter_step)
 
+        if iter_step % log_every == 0:
             train_summ, y_len = self.sess.run([self.train_summ, self.Y_len])
             self.train_writer.add_summary(train_summ, global_step=iter_step)
 
@@ -750,7 +750,6 @@ class Model():
             self.save()
             self.logger.info("Running final validation run")
             return self.run_validation_full(final_val_samples)
-            # self.print_k()  # TODO REMOVE!!!
         except Exception as ex:
             if not isinstance(ex, KeyboardInterrupt):
                 self.logger.error("Error happened", exc_info=1)

### 2017-04-01 src/util.py
Fixed basecalling bug

@@ -349,7 +349,7 @@ def sigopt_double(name, min, max):
 
 def get_raw_signal(fast5_path):
     """
-        Return 1D raw signal from fast5_path
+        Return 1D raw signal from fast5_path. No preprocessing
     """
     with h5py.File(fast5_path, 'r') as h5:
         reads = h5['Analyses/EventDetection_000/Reads']
@@ -361,7 +361,4 @@ def get_raw_signal(fast5_path):
         signal = h5['Raw/Reads/' + target_read]['Signal'][start_pad:].astype(np.float32)
         signal_len = h5['Raw/Reads/' + target_read].attrs['duration'] - start_pad
         assert(len(signal) == signal_len)
-
-        signal -= 646.11133
-        signal /= 75.673653
         return signal

### 2017-04-01 src/model_small_student.py
Added Support for multiple teachers (ensamble)

@@ -85,14 +85,15 @@ params = [
 default_params = {
     'initial_lr': 1e-4,
     'decay_factor': 0.1,
+    'ctc_scale': 1e-4
 }
 
 
 if __name__ == "__main__":
     print(model_setup_params(default_params))
     model = model_utils.TeacherStudentModel(
-        'model_small',
-        os.path.join(model_utils.repo_root, 'log', 'protagonist', 'model_small_18071_9943114'),
+        ['model_small'],
+        [os.path.join(model_utils.repo_root, 'log', 'protagonist', 'model_small_18071_9943114')],
         **model_setup_params(default_params)
     )
 

### 2017-04-01 src/model_utils.py

@@ -775,39 +775,46 @@ class Model():
 
 #  RNN based teacher not supported
 class TeacherStudentModel(Model):
-    def __init__(self, teacher_name, teacher_dir, model_fn, hyper, **kwargs):
-        self.teacher_param = load_model_parms(teacher_name, teacher_dir)
-        print("dirself", dir(self))
+    def __init__(self, teacher_names, teacher_dirs, model_fn, hyper, **kwargs):
+        self.teacher_params = [load_model_parms(teacher_name, teacher_dir) for teacher_name, teacher_dir in zip(teacher_names, teacher_dirs)]
+        self.ctc_scale = hyper['ctc_scale']
         super().__init__(model_fn=model_fn, hyper=hyper, **kwargs)
 
     def _setup_loss(self, logits):
-        print("TeacherStudentModel _setup_loss")
-        with tf.variable_scope("teacher"):
-            alt_logits = self._setup_logits(self.teacher_param['model_fn'], self.teacher_param['hyper'])
+        ensamble = []
+        for i, teacher_params in enumerate(self.teacher_params):
+            with tf.variable_scope("teacher%d" % i):
+                alt_logits = self._setup_logits(teacher_params['model_fn'], teacher_params['hyper'])['logits']
+                ensamble.append(tf.nn.softmax(alt_logits))
+
+        ensamble = tf.add_n(ensamble) / len(self.teacher_params)
+        self.ctc_loss = super()._setup_loss(logits)
 
         with tf.name_scope("loss"):
-            loss = tf.square(logits - alt_logits['logits'])
-            return tf.reduce_mean(loss)
+            self.ce_loss = tf.reduce_mean(tf.nn.softmax_cross_entropy_with_logits(labels=ensamble, logits=logits))
+            return self.ce_loss + self.ctc_scale * self.ctc_loss
 
     def _setup_saver(self):
         super()._setup_saver()
-        mapping = {'/'.join(v.op.name.split('/')[1:]): v for v in self.g.get_collection(tf.GraphKeys.TRAINABLE_VARIABLES, 'teacher')
-        }
-        self.teacher_saver = tf.train.Saver(
-            mapping
-        )
+        self.teacher_savers = []
+        for i in range(len(self.teacher_params)):
+            mapping = {'/'.join(v.op.name.split('/')[1:]): v for v in self.g.get_collection(tf.GraphKeys.TRAINABLE_VARIABLES, 'teacher%d' % i)
+            }
[…]

### 2017-04-01 src/model_utils.py
Small prediction refactor

@@ -196,13 +196,17 @@ class Model():
             return sol
 
         with tf.name_scope("prediction"):
-            predicted, prdicted_logprob = tf.nn.ctc_beam_search_decoder(self.logits, tf.div(self.X_batch_len, self.shrink_factor), merge_repeated=True, top_paths=1)
+            predicted = self._predict_from_logits()
             self.pred = dedup_output(tf.cast(predicted[0], tf.int32))
             self.dense_pred = tf.sparse_tensor_to_dense(self.pred, default_value=-1)
             self.edit_distance = tf.edit_distance(
                 self.pred,
                 dedup_output(self.Y_batch)
             )
+
+    def _predict_from_logits(self):
+        predicted, _ = tf.nn.ctc_beam_search_decoder(self.logits, tf.div(self.X_batch_len, self.shrink_factor), merge_repeated=True, top_paths=1)
+        return predicted
 
     def _setup_train(self):
         """

### 2017-04-01 src/model_utils.py
Ansambles implemented

@@ -784,7 +784,7 @@ class TeacherStudentModel(Model):
         self.ctc_scale = hyper['ctc_scale']
         super().__init__(model_fn=model_fn, hyper=hyper, **kwargs)
 
-    def _setup_loss(self, logits):
+    def _get_ensamble(self):
         ensamble = []
         for i, teacher_params in enumerate(self.teacher_params):
             with tf.variable_scope("teacher%d" % i):
@@ -792,8 +792,11 @@ class TeacherStudentModel(Model):
                 ensamble.append(tf.nn.softmax(alt_logits))
 
         ensamble = tf.add_n(ensamble) / len(self.teacher_params)
-        self.ctc_loss = super()._setup_loss(logits)
+        return ensamble
 
+    def _setup_loss(self, logits):
+        self.ctc_loss = super()._setup_loss(logits)
+        ensamble = self._get_ensamble()
         with tf.name_scope("loss"):
             self.ce_loss = tf.reduce_mean(tf.nn.softmax_cross_entropy_with_logits(labels=ensamble, logits=logits))
             return self.ce_loss + self.ctc_scale * self.ctc_loss
@@ -813,7 +816,7 @@ class TeacherStudentModel(Model):
             checkpoint = tf.train.latest_checkpoint(teacher_param['log_dir'])
             if checkpoint is None:
                 raise ValueError("Teacher checkpoint must exist!")
-            self.logger.info("Loading teacher%d weights from %s", checkpoint)
+            self.logger.info("Loading teacher%d weights from %s", i, checkpoint)
             saver.restore(self.sess, checkpoint)
 
         self.print_k()
@@ -821,4 +824,26 @@ class TeacherStudentModel(Model):
     def print_k(self):
         for i in range(len(self.teacher_params)):
             for v in self.g.get_collection(tf.GraphKeys.TRAINABLE_VARIABLES, 'teacher%d' % i):
-                print(v.op.name, self.sess.run(v).ravel()[:5])
+                if '/k' in v.op.name:
+                    print(v.op.name, self.sess.run(v).ravel()[:5])
+
+
+class Ensamble(TeacherStudentModel):
+    def __init__(self, teacher_names, teacher_dirs):
+        from model_small import default_params, model_setup_params
+        params = model_setup_params(default_params)
+        params['hyper']['ctc_scale'] = 0.0
+        super().__init__(teacher_names, teacher_dirs, **params)
+
[…]

### 2017-04-01 src/karla_ensamble.py
Test files for ansambles

@@ -1,0 +1,15 @@
+from dotenv import load_dotenv, find_dotenv
+import model_utils
+import os
+load_dotenv(find_dotenv())
+
+
+if __name__ == "__main__":
+    model = model_utils.Ensamble(
+        ['residual_deep'],
+        [os.path.join(model_utils.repo_root, 'log', 'karla_docker', 'residual_deep_prod_17466_%d' % x) for x in [9939286]],
+    )
+
+    model.init_session()
+    model.run_validation_full(frac=500)
+    model.close_session()

### 2017-04-01 src/model_small_ensamble.py

@@ -1,0 +1,15 @@
+from dotenv import load_dotenv, find_dotenv
+import model_utils
+import os
+load_dotenv(find_dotenv())
+
+
+if __name__ == "__main__":
+    model = model_utils.Ensamble(
+        ['model_small'],
+        [os.path.join(model_utils.repo_root, 'log', 'protagonist', 'model_small_18071_9943114')],
+    )
+
+    model.init_session()
+    model.run_validation_full(frac=500)
+    model.close_session()

### 2017-04-02 src/input_readers.py
Added RollingMedianAlignedRaw InputReader

@@ -1,6 +1,7 @@
 import os
 import socket
 from random import shuffle
+from scipy.signal import medfilt
 
 import numpy as np
 import util
@@ -23,8 +24,7 @@ class InputReader():
 
 class AlignedRaw(InputReader):
 
-    @staticmethod
-    def preprocessSignal(signal):
+    def preprocessSignal(self, signal):
         return (signal - 646.11133) / 75.673653
 
     def read_fast5_raw_ref(self, fast5_path, ref_path, block_size_x, block_size_y, num_blocks, warn_if_short=False):
@@ -107,6 +107,11 @@ class AlignedRaw(InputReader):
         signal = util.get_raw_signal(fast5_path)
         signal = self.preprocessSignal(signal)
         return signal.reshape(1, -1, 1)
+
+
+class RollingMedianAlignedRaw(AlignedRaw):
+    def preprocessSignal(self, signal):
+        return super().preprocessSignal(medfilt(signal, kernel_size=5))
 
 
 def sanitize_input_line(fname):

### 2017-04-02 src/residual_deep.py
Setup current best params

@@ -85,9 +85,9 @@ params = [
 ]
 
 default_params = {
-    'initial_lr': 0.0014691078293764034,
-    'decay_factor': 0.001341733349277896,
-    'num_layers': 17,
+    'initial_lr': 0.000965352400196344,
+    'decay_factor': 0.0017387361908150767,
+    'num_layers': 20,
     'num_sub_layers': 2
 }
 

### 2017-04-02 src/model_utils.py
Iteration of Ensambles

@@ -784,14 +784,26 @@ class TeacherStudentModel(Model):
         self.ctc_scale = hyper['ctc_scale']
         super().__init__(model_fn=model_fn, hyper=hyper, **kwargs)
 
-    def _get_ensamble(self):
-        ensamble = []
+    def _setup_teachers(self):
+        if hasattr(self, 'teachers'):
+            self.logger.warn("_setup_teachers CALLED MORE THAN ONCE")
+        self.teachers = []
         for i, teacher_params in enumerate(self.teacher_params):
             with tf.variable_scope("teacher%d" % i):
-                alt_logits = self._setup_logits(teacher_params['model_fn'], teacher_params['hyper'])['logits']
-                ensamble.append(tf.nn.softmax(alt_logits))
+                teacher = super()._setup_logits(teacher_params['model_fn'], teacher_params['hyper'])
 
-        ensamble = tf.add_n(ensamble) / len(self.teacher_params)
+            mapping = {'/'.join(v.op.name.split('/')[1:]): v for v in self.g.get_collection(tf.GraphKeys.TRAINABLE_VARIABLES, 'teacher%d' % i)
+            }
+            print(mapping.keys())
+            teacher['saver'] = tf.train.Saver(mapping)
+            teacher['log_dir'] = teacher_params['log_dir']
+            self.teachers.append(teacher)
+
+    def _get_ensamble(self):
+        self._setup_teachers()
+        ensamble = [tf.nn.softmax(teacher['logits']) for teacher in self.teachers]
+        ensamble = tf.stack(ensamble)
+        ensamble = tf.reduce_mean(ensamble, axis=0)
         return ensamble
 
     def _setup_loss(self, logits):
@@ -801,25 +813,22 @@ class TeacherStudentModel(Model):
             self.ce_loss = tf.reduce_mean(tf.nn.softmax_cross_entropy_with_logits(labels=ensamble, logits=logits))
             return self.ce_loss + self.ctc_scale * self.ctc_loss
 
-    def _setup_saver(self):
-        super()._setup_saver()
-        self.teacher_savers = []
-        for i in range(len(self.teacher_params)):
-            mapping = {'/'.join(v.op.name.split('/')[1:]): v for v in self.g.get_collection(tf.GraphKeys.TRAINABLE_VARIABLES, 'teacher%d' % i)
-            }
-            self.teacher_savers.append(tf.train.Saver(mapping))
+    # def _setup_saver(self):
+        # super()._setup_saver()
 
[…]

### 2017-04-02 src/model_utils.py
Fixed ensamble bug
* Moving averages weren't restored like their variables counterpart.

@@ -508,6 +508,8 @@ class Model():
         return avg_loss, avg_edit_distance
 
     def basecall_sample(self, fast5_path):
+        with self.g.as_default():
+            is_training(False, session=self.sess)
         signal = self.in_data.get_signal(fast5_path)
         t = perf_counter()
         basecalled = self.sess.run(
@@ -522,6 +524,13 @@ class Model():
         self.logger.debug("Basecalled %s in %.3f", fast5_path, perf_counter() - t)
 
         return "".join(util.decode(basecalled))
+
+    def skim_trainables(self):
+        vars = self.g.get_collection(tf.GraphKeys.TRAINABLE_VARIABLES)
+        vals = self.sess.run(vars)
+        for var, val in zip(vars, vals):
+            print(var.op.name, val.ravel()[:5])
+
 
     def get_aligement(self, fast5_path, ref_path, verbose):
         t = perf_counter()
@@ -792,9 +801,11 @@ class TeacherStudentModel(Model):
             with tf.variable_scope("teacher%d" % i):
                 teacher = super()._setup_logits(teacher_params['model_fn'], teacher_params['hyper'])
 
-            mapping = {'/'.join(v.op.name.split('/')[1:]): v for v in self.g.get_collection(tf.GraphKeys.TRAINABLE_VARIABLES, 'teacher%d' % i)
+            mapping = {'/'.join(v.op.name.split('/')[1:]): v for v in self.g.get_collection(tf.GraphKeys.GLOBAL_VARIABLES, 'teacher%d' % i)
             }
-            print(mapping.keys())
+            self.logger.debug("Teacher %d uses following varibles", i)
+            for k in mapping.keys():
+                self.logger.debug(k)
             teacher['saver'] = tf.train.Saver(mapping)
             teacher['log_dir'] = teacher_params['log_dir']
             self.teachers.append(teacher)
@@ -812,9 +823,6 @@ class TeacherStudentModel(Model):
         with tf.name_scope("loss"):
             self.ce_loss = tf.reduce_mean(tf.nn.softmax_cross_entropy_with_logits(labels=ensamble, logits=logits))
             return self.ce_loss + self.ctc_scale * self.ctc_loss
-
-    # def _setup_saver(self):
-        # super()._setup_saver()
 
     def _restore_teachers(self):
         self.print_k()

### 2017-04-02 src/model_utils.py
Implemented Ensabmled...it seems to be working now

@@ -808,6 +808,7 @@ class TeacherStudentModel(Model):
                 self.logger.debug(k)
             teacher['saver'] = tf.train.Saver(mapping)
             teacher['log_dir'] = teacher_params['log_dir']
+            self.logger.info("Configured teacher%d, logdir %s", i, teacher['log_dir'])
             self.teachers.append(teacher)
 
     def _get_ensamble(self):
@@ -862,9 +863,13 @@ class Ensamble(TeacherStudentModel):
         pass
 
     def _setup_logits(self, *args):
-        self._setup_teachers()
-        return self.teachers[0]
+        return {
+            'logits': None,
+            'init_state': tf.constant(0.0),
+            'final_state': tf.constant(0.0)
+        }
 
-    # def _predict_from_logits(self):
-        # predicted, _ = tf.nn.ctc_beam_search_decoder(self._get_ensamble(), tf.div(self.X_batch_len, self.shrink_factor), merge_repeated=True)
-        # return predicted
+    def _predict_from_logits(self):
+        ens = self._get_ensamble()
+        self.logits = tf.log(ens)
+        return super()._predict_from_logits()

### 2017-04-02 src/model_utils.py
Optimized print_k

@@ -841,9 +841,10 @@ class TeacherStudentModel(Model):
 
     def print_k(self):
         for i in range(len(self.teacher_params)):
-            for v in self.g.get_collection(tf.GraphKeys.TRAINABLE_VARIABLES, 'teacher%d' % i):
-                if '/k' in v.op.name:
-                    print(v.op.name, self.sess.run(v).ravel()[:5])
+            vars = [v for v in self.g.get_collection(tf.GraphKeys.TRAINABLE_VARIABLES, 'teacher%d' % i) if '/k' in v.op.name]
+            vals = self.sess.run(vars)
+            for var, val in zip(vars, vals):
+                print(var.op.name, val.ravel()[:5])
 
 
 class Ensamble(TeacherStudentModel):

### 2017-04-02 src/tt.py
Useful debug script

@@ -1,0 +1,20 @@
+import model_utils
+import os
+model = model_utils.Ensamble(
+    ['model_small'],
+    [os.path.join(model_utils.repo_root, 'log', 'protagonist', 'model_small_18071_9943114')],
+)
+
+model.init_session()
+m1 = model.basecall_sample('/home/lpp/Downloads/minion/pass/26075_ch161_read656_strand.fast5')
+model.close_session()
+m1_ld = model.log_dir
+
+import eval
+model = eval.load_model('model_small', os.path.join(model_utils.repo_root, 'log', 'protagonist', 'model_small_18071_9943114'))
+model.init_session()
+model.restore()
+m2 = model.basecall_sample('/home/lpp/Downloads/minion/pass/26075_ch161_read656_strand.fast5')
+model.close_session()
+print(m1, m1_ld)
+print(m2)

### 2017-04-08 paper/main.tex
Pre-alpha draft

@@
 \voditelj{Mile Sikić}
-\title{Sustav potpore odlučivanju za predviđanje i optimalnu nabavu
-}
-\engtitle{A decision support system for forecasting and optimal procurement}
+\title{CNNNano --- MinION  basecaller pogonjen dubokim ucenjem}
+\engtitle{CNNNano --- MinION deep learning powered basecaller}
 \author{Neven Miculinić}
@@
 \begin{sazetak}
-    Optimalna nabava u većini industrija uključuje predviđanje dviju veliničina: cijene sirovina i potražnje.
-Cilj ovoga rada je ujediniti predviđanje u model planiranja proizvodnje sa ciljem minimiziranja cjelokupne nabave, troškova skladištenja i proizvodnje uz zadovoljenje potražnje.
-Sustav potpore odlučivanju treba omogućiti donositelju odluke integraciju kvalitativne nestrukturiranje informacije kroz jednostavno sučelje, odabir scenarija i profinjenje rješenja.
+TODO: ovo je pre-alpha draft...bude se vec osvjezilo
 
@@
 \begin{abstract}
-Optimal procurement in most industries involves forecasting of two
-quantities: prices of raw materials and customer's demand.  The aim of
-this work is to integrate forecasts into production planning models,
-with the aim of minimizing overall procurement, holding and production
-costs under demand satisfaction constraints.  The decision support
-system should allow the decision maker to integrate qualitative,
-unstructured information through a simple interface, for scenario
-selection and solution refinement.
+    The MinION device by Oxford Nanopore is the first portable sequencing device.
+MinION is able to produce very long reads (reads over 100 kBp were reported), however it suffers from high sequencing error rate.
+In this paper, we show that the error rate can be reduced by improving the base calling process using residual convolutional neural networks(CNN) combined with Connectionist Temporal Classification (CTC).
 \keywords{procurement planning, dynamic lot sizing model, stochastic demands}
@@
 \chapter{Introduction}
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+In this paper, we introduce the base caller for the MinION nanopore sequencing platform (Mikheyev and Tin, 2014).
+The MinION device by Oxford Nanopore, weighing only 90 grams, is currently the smallest high-throughput DNA sequencer.
+Thanks to its low capital costs, small size and the possibility of analyzing the data in real time as they are produced, MinION is very promising for clinical applications, such as monitoring infectious disease outbreaks (Judge et al., 2015; Quick et al., 2015, 2016), and characterizing structural variants in cancer (Norris et al., 2016).
+Although MinION is able to produce long reads, they have a high sequencing error rate.
+In this paper, we show that this error rate can be reduced by replacing the default base caller provided by the manufacturer with a properly trained neural network model.
+
+In the MinION device, single-stranded DNA fragments move through nanopores, which causes drops in the electric current.
+The electric current is measured at each pore several thousand times per second.
+The electric current depends mostly on the context of several DNA bases passing through the pore at the time of measurement.
+As the DNA moves through the pore, the context shifts and the electric current changes.
+
+A MinION device typically yields reads several thousand bases long; reads as long as 100,000 bp have been reported.
+1D read have usually error rate about   30\%.
+After parameter training, base calling can be performed by running the Viterbi algorithm, which will result in the sequence of states with the highest likelihood.
+
 \chapter{Related work}
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+Official MinION basecaller is
+
+Note that neural networks were previously used for base calling Sanger sequencing reads (Tibbetts et al., 1994; Mohammed et al., 2013), though the nature of MinION data is rather different.
+Several tools for processing nanopore sequencing data were already published, including read mappers (Jain et al., 2015; Sovic et al., 2015), and error correction tools using short Illumina reads (Goodwin et al., 2015).
+Most closes related to our work are Nanopolish (Loman et al., 2015b) and PoreSeq (Szalay and Golovchenko, 2015).
+Both of these tools create a consensus sequence by combining information from multiple overlapping reads, considering not only the final base calls from Metrichor, but also the sequence of events.
+They analyze the events by hidden Markov models with emission probabilities provided by Metrichor.
+In contrast, our base caller does not require read overlaps, it processes reads individually and
+
 \chapter{Model description}
@@
 Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.\cite{deepnano}
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+% \cite{deepnano}
 % \bibliographystyle{fer}
-\bibliographystyle{abbrv}
-\bibliography{refs}
+% \bibliographystyle{abbrv}
+% \bibliography{refs}
 

### 2017-04-27 src/preprocess_dataset.py
Minor bug fixes

@@ -4,13 +4,14 @@ import os
 import shutil
 import subprocess
 import tempfile
-import sys
 import time
+from argparse import ArgumentParser
+from configparser import ConfigParser
+
 import h5py
 import pysam
 from tqdm import tqdm
-from configparser import ConfigParser
-from argparse import ArgumentParser
+
 from bioinf_utils import reverse_complement, cigar_pairs_to_str, read_fasta
 
 
@@ -24,8 +25,12 @@ def get_target_sequences(sam_out):
             if x.is_unmapped:
                 logging.warning("%s unmapped" % name)
                 continue
+            try:
+                target = x.get_reference_sequence()
+            except ValueError:
+                logging.error("%s Mapped but reference len equals 0" % name)
+                continue
 
-            target = x.get_reference_sequence()
             ref_name = x.reference_name
             length = x.reference_length
             start_pos = x.reference_start
@@ -40,7 +45,7 @@ def get_target_sequences(sam_out):
     return result_dict
 
 
-def _preprocess_batch(files_in_batch, generate_sam_f, ref_starts, out_root):
+def _align_for_reference_batch(files_in_batch, generate_sam_f, ref_starts, out_root):
     name_to_file = {}
     tmp_work_dir = tempfile.mkdtemp()
     tmp_sam_path = os.path.join(tmp_work_dir, 'tmp.sam')
@@ -57,15 +62,15 @@ def _preprocess_batch(files_in_batch, generate_sam_f, ref_starts, out_root):
 
                 fq_file.write(fastq)
                 fq_file.write(b'\n')
-        print("load fq", time.time() - s_load)
+        logging.debug("load fq", time.time() - s_load)
         s_sam = time.time()
         generate_sam_f(tmp_fastq_path, tmp_sam_path)
-        print("generate sam", time.time() - s_sam)
+        logging.debug("generate sam", time.time() - s_sam)
         s_dict = time.time()
         result_dict = get_target_sequences(tmp_sam_path)
-        print("result_dict", time.time() - s_dict)
+        logging.debug("result_dict", time.time() - s_dict)
 
-          # cleanup
+    # cleanup
     shutil.rmtree(tmp_work_dir)
     s_dump_files = time.time()
     for name, (target, ref_name, start_position, length, cigar) in result_dict.items():
@@ -84,11 +89,7 @@ def _preprocess_batch(files_in_batch, generate_sam_f, ref_starts, out_root):
     print("total", time.time() - s_load)
 
 
-
-
-
-
[…]

### 2017-04-28 src/bioinf_utils.py
Added official edlib support,  different cigar conversion util methods, error_rates calculation....

@@ -1,5 +1,6 @@
 import itertools
 import logging
+from collections import Counter
 
 CIGAR_TO_BYTE = {
     'M': 0,
@@ -12,6 +13,13 @@ CIGAR_TO_BYTE = {
     '=': 7,
     'X': 8
 }
+
+CIGAR_OPERATIONS = 'MIDNSHP=X'
+CIGAR_MATCH_MISSMATCH = 'M=X'
+CIGAR_MATCH = ['=']
+CIGAR_MISSMATCH = ['X']
+CIGAR_INSERTION = 'ISH'
+CIGAR_DELETION = 'DNP'
 
 BYTE_TO_CIGAR = {v: k for k, v in CIGAR_TO_BYTE.items()}
 
@@ -41,9 +49,37 @@ def cigar_pairs_to_str(cigar_pairs):
     return ''.join(cigar)
 
 
-def decompress_cigar(cigar_pairs):
-    extended = (cnt * [cigar_int_to_c(b)] for b, cnt in cigar_pairs)
+def cigar_str_to_pairs(cigar):
+    split_locations = []
+    for i, c in enumerate(cigar):
+        if c in CIGAR_OPERATIONS:
+            split_locations.append(i)
+
+    cigar_pairs = []
+    for i, end in enumerate(split_locations):
+        start = split_locations[i-1] + 1 if i > 0 else 0
+
+        cnt = int(cigar[start:end])
+        char = cigar[end].upper()
+        cigar_pairs.append((char, cnt))
+    return cigar_pairs
+
+
+def decompress_cigar_pairs(cigar_pairs, mode='ints'):
+    if mode == 'ints':
+        convert = cigar_int_to_c
+    elif mode == 'chars':
+        convert = lambda x: x
+    else:
+        raise Exception('Invalid mode argument. Expected ints or chars')
+
+    extended = (cnt * [convert(b)] for b, cnt in cigar_pairs)
     return ''.join(itertools.chain.from_iterable(extended))
+
+
+def decompress_cigar(cigar_str):
+    cigar_pairs = cigar_str_to_pairs(cigar_str)
+    return decompress_cigar_pairs(cigar_pairs, mode='chars')
 
 
 def get_ref_len_from_cigar(cigar_pairs):
@@ -56,7 +92,7 @@ def get_ref_len_from_cigar(cigar_pairs):
     return ref_len
 
 
-def reference_align_string(ref, cigar_pairs):
+def reference_align_string(ref, cigar_int_pairs):
     """
[…]

### 2017-04-28 src/model_utils.py

@@ -18,12 +18,13 @@ from tflearn.summaries import add_gradients_summary, add_activations_summary
 import logging
 from tflearn.config import is_training, get_training_mode
 from slacker_log_handler import SlackerLogHandler
-from edlib import Edlib
+import edlib
 import inspect
 import json
 import importlib
 import subprocess
 
+import bioinf_utils
 
 # UGLY UGLY HACK!
 for name, logger in logging.root.manager.loggerDict.items():
@@ -573,13 +574,15 @@ class Model():
         self.logger.debug("Basecalled \n%s", basecalled)
         self.logger.debug("Target \n%s", target)
 
-        result = Edlib().align(basecalled, target)
-        self.logger.debug("Aligment %s", "".join(map(str, result.alignment)))
+        result = edlib.align(basecalled, target)
+        cigar = bioinf_utils.decompress_cigar(result['cigar'])
+
+        self.logger.debug("Aligment %s", cigar)
         self.logger.debug("Whole time %.3f", perf_counter() - t)
 
-        acc = np.sum(np.array(result.alignment) == Edlib().EDLIB_EDOP_MATCH) / len(result.alignment)
-        nedit = result.edit_distance / len(target)
-        return nedit, acc
+        error_rate, match_rate, missmatch_rate, ins_rate, del_rate = bioinf_utils.error_rates_from_cigar(cigar)
+        nedit = result["editDistance"] / len(target)
+        return nedit, match_rate
 
     def run_validation_full(self, frac, verbose=False, fasta_out_dir=None, ref=None):
         """

### 2017-04-28 src/preprocess_dataset.py

@@ -12,7 +12,7 @@ import h5py
 import pysam
 from tqdm import tqdm
 
-from bioinf_utils import reverse_complement, cigar_pairs_to_str, read_fasta
+from bioinf_utils import reverse_complement, decompress_cigar_pairs, read_fasta
 
 
 def get_target_sequences(sam_out):
@@ -40,7 +40,7 @@ def get_target_sequences(sam_out):
                 target = reverse_complement(target)
                 cigar_pairs = list(reversed(cigar_pairs))
 
-            cigar_str = cigar_pairs_to_str(cigar_pairs)
+            cigar_str = decompress_cigar_pairs(cigar_pairs, mode='ints')
             result_dict[name] = [target, ref_name, start_pos, length, cigar_str]
     return result_dict
 
@@ -192,6 +192,6 @@ if __name__ == '__main__':
     args = parser.parse_args()
 
     preprocess_all_ref(args.dataset_conf_path, args.split, generate_sam_graphmap)
-    
+
 
 

### 2017-04-28 src/util.py

@@ -1,7 +1,7 @@
-from edlib import Edlib
+import edlib
 import h5py
 import numpy as np
-
+import bioinf_utils
 
 def next_num(prev, symbol):
     val = {
@@ -202,22 +202,21 @@ def get_basecalled_sequence(basecalled_events):
         seq.extend(add_chr)
     return ''.join(seq)
 
-
-def init_aligment_end_struct(ref_seq, called_seq, aligment_seq):
+def init_aligment_end_struct(ref_seq, called_seq, full_cigar):
     # called_seq[i] alignes up to aligned_upto[i] index in ref_seq
 
     aligned_upto = []
     ref_idx = 0
 
-    for c in aligment_seq:
-        if c == Edlib.EDLIB_EDOP_MATCH or c == Edlib.EDLIB_EDOP_MISMATCH:
+    for c in full_cigar:
+        if c in bioinf_utils.CIGAR_MATCH_MISSMATCH:
             aligned_upto.append(ref_idx)
             ref_idx += 1
 
-        elif c == Edlib.EDLIB_EDOP_INSERT:
+        elif c in bioinf_utils.CIGAR_INSERTION:
             aligned_upto.append(max(0, ref_idx - 1))
 
-        elif c == Edlib.EDLIB_EDOP_DELETE:
+        elif c in bioinf_utils.CIGAR_DELETION:
             ref_idx += 1
 
     aligned_upto.append(aligned_upto[-1])
@@ -237,11 +236,13 @@ class AligmentError(Exception):
     pass
 
 
-def extract_blocks(ref_seq, called_seq, events_len, block_size, num_blocks, skip_first=True):
-    aligner = Edlib()
-    result = aligner.align(called_seq, ref_seq)
-    aligment_seq = result.alignment
-    aligment_end = init_aligment_end_struct(ref_seq, called_seq, aligment_seq)
+def extract_blocks(ref_seq, called_seq, events_len, block_size, num_blocks, cigar=None, skip_first=True):
+    if cigar is None:
+        result = edlib.align(called_seq, ref_seq, mode="NW", task="path")
+        aligment_cigar = result['cigar']
+        cigar = bioinf_utils.decompress_cigar(aligment_cigar)
+
+    aligment_end = init_aligment_end_struct(ref_seq, called_seq, cigar)
     ref_seq = np.fromstring(ref_seq, np.int8)
 
     y = np.zeros([num_blocks * block_size], dtype=np.uint8)
[…]

### 2017-04-28 paper/main.tex
Added nanocall into paper
Changed abstract

@@
 \begin{sazetak}
-TODO: ovo je pre-alpha draft...bude se vec osvjezilo
+TODO
 
-\kljucnerijeci{planiranje nabave, stohastička potražnja, optimalna količina nabave, Model dinamičke ekonomske količine nabave}
+\kljucnerijeci{TODO}
 \end{sazetak}
@@
 MinION is able to produce very long reads (reads over 100 kBp were reported), however it suffers from high sequencing error rate.
-In this paper, we show that the error rate can be reduced by improving the base calling process using residual convolutional neural networks(CNN) combined with Connectionist Temporal Classification (CTC).
+In this paper, we show novel model reducing the error rate can be reduced by improving the base calling process using gated residual convolutional neural networks (CNN) combined with Connectionist Temporal Classification (CTC) learning approach.
 \keywords{procurement planning, dynamic lot sizing model, stochastic demands}

### 2017-04-28 paper/main.tex
Changed paper structure

@@
 
-\chapter{Related work}
+\chapter{Background}
 
@@
 
-\chapter{Model description}
+\chapter{Method}
+
 Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
@@
 Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-\chapter{Experimental results}
+\chapter{Results}
+
 Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.

### 2017-04-28 paper/main.tex
Small chenges

@@
 MinION is able to produce very long reads (reads over 100 kBp were reported), however it suffers from high sequencing error rate.
-In this paper, we show novel model reducing the error rate can be reduced by improving the base calling process using gated residual convolutional neural networks (CNN) combined with Connectionist Temporal Classification (CTC) learning approach.
+In this paper, we show novel model reducing the error rate can be reduced by improving the base calling process on 1D reads using gated residual convolutional neural networks (CNN) combined with Connectionist Temporal Classification (CTC) learning approach.
 \keywords{procurement planning, dynamic lot sizing model, stochastic demands}
@@
 
-In this paper, we introduce the base caller for the MinION nanopore sequencing platform (Mikheyev and Tin, 2014).
-The MinION device by Oxford Nanopore, weighing only 90 grams, is currently the smallest high-throughput DNA sequencer.
+In this paper, we introduce the base caller for the MinION nanopore sequencing platform~\cite{mikheyev2014first}.The MinION device by Oxford Nanopore, weighing only 90 grams, is currently the smallest high-throughput DNA sequencer.
 Thanks to its low capital costs, small size and the possibility of analyzing the data in real time as they are produced, MinION is very promising for clinical applications, such as monitoring infectious disease outbreaks (Judge et al., 2015; Quick et al., 2015, 2016), and characterizing structural variants in cancer (Norris et al., 2016).

### 2017-04-28 paper/main.tex
Abstract changed
Added some citations.

@@
 In this paper, we show novel model reducing the error rate can be reduced by improving the base calling process on 1D reads using gated residual convolutional neural networks (CNN) combined with Connectionist Temporal Classification (CTC) learning approach.
-\keywords{procurement planning, dynamic lot sizing model, stochastic demands}
+
+    Whole process is trainned end-to-end from raw signal on R9 sequecing pore and once trainned it's independent from Metrichor.
+\keywords{}
 \end{abstract}
@@
 In this paper, we introduce the base caller for the MinION nanopore sequencing platform~\cite{mikheyev2014first}.The MinION device by Oxford Nanopore, weighing only 90 grams, is currently the smallest high-throughput DNA sequencer.
-Thanks to its low capital costs, small size and the possibility of analyzing the data in real time as they are produced, MinION is very promising for clinical applications, such as monitoring infectious disease outbreaks (Judge et al., 2015; Quick et al., 2015, 2016), and characterizing structural variants in cancer (Norris et al., 2016).
+Thanks to its low capital costs, small size and the possibility of analyzing the data in real time as they are produced, MinION is very promising for clinical applications, such as monitoring infectious disease outbreaks~\cite{judge2015early}\cite{quick2016real}, characterizing structural variants in cancer\cite{norris2016nanopore} and even full human genome assembly~\cite{jain2017nanopore}.
+
 Although MinION is able to produce long reads, they have a high sequencing error rate.
+This has been somewhat aleviated with new R9 pore model, replacing older R7 ones.
 In this paper, we show that this error rate can be reduced by replacing the default base caller provided by the manufacturer with a properly trained neural network model.

### 2017-04-28 paper/main.tex
ch

@@
 
+\section{Sequencing overview}
+Informally, the MinION sequencer works as follows.
+First, DNA is sheared into fragments of 8–20 kbp and adapters are ligated to either end of the fragments.
+The resulting DNA fragments pass through a protein embedded in a membrane via a nanometre-sized channel (this protein is the `nanopore').
+A single strand of DNA passes through the pore; the optional use of a hairpin adapter at one end of the fragment allows the two strands of DNA to serially pass through nanopore, allowing two measurements of the fragment.
+With hairpin it's called 2D read, which isn't topic of this paper.
+In here we focus only on 1D reads.
+
+ In ONT terminology, the first strand going through the nanopore is the template, and the second is the complement.
+As a DNA strand passes through the pore it partially blocks the flow of electric current through the pore.
+The flow of current is sampled over time which is the observable output of the system.
+The central idea is that the single-stranded DNA product present in the nanopore affects the current in a way that is strong enough to enable decoding the electric signal data into a DNA sequence.
+This process, called basecalling, takes as input a list of current measurements, and produces as output a list of DNA bases most likely to have generated those currents.
+
 Official MinION basecaller is

### 2017-04-29 paper/main.tex
Small changes
Added sections

@@
 
-Official MinION basecaller is
+The nanopore is 6 nucleoides wide, and eariler HMM based models model it as such.
 
-Note that neural networks were previously used for base calling Sanger sequencing reads (Tibbetts et al., 1994; Mohammed et al., 2013), though the nature of MinION data is rather different.
-Several tools for processing nanopore sequencing data were already published, including read mappers (Jain et al., 2015; Sovic et al., 2015), and error correction tools using short Illumina reads (Goodwin et al., 2015).
-Most closes related to our work are Nanopolish (Loman et al., 2015b) and PoreSeq (Szalay and Golovchenko, 2015).
-Both of these tools create a consensus sequence by combining information from multiple overlapping reads, considering not only the final base calls from Metrichor, but also the sequence of events.
-They analyze the events by hidden Markov models with emission probabilities provided by Metrichor.
-In contrast, our base caller does not require read overlaps, it processes reads individually and
+\section{Basecalling}
+
+The core of the decoding process is the basecalling step. Official basecaller is Metrichor, previously perfomed in the cloud before being opensourced under name Nanonet.
+
+Earlier models were (Hidden markov model) HMM based where hidden state modeled DNA sequence of length 6 (6-mer) in the nanopore.
+Pore models were used in computing emission probabilities.~\cite{loman2015complete,schreiber2015analysis,szalay2015novo,timp2012dna} and the recest open source HMM based basecaller Nanocall~\cite{david2016nanocall}.
+
+Recent models opted using (Recurrent neural network) RNN, notably DeepNano~\cite{deepnano} and recently open sourced official Nanonets from ONT.
 

### 2017-04-29 paper/main.tex
Added method beginning

@@
 
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+Instead of opting for traditional path using HMM or RNN we tried using CNN~(Convolutional neural networks)~\cite{lecun-98}, that is their residual version~\cite{he2016deep}.
+We opted out for gated residuan network variant~\cite{savarese2016learning}.
+For loss we used CTC (Connectionist temporal classification)~\cite{graves2006connectionist} between basecallled and target sequence.
+The implementation used is open source warp-ctc~\cite{warpctc}.
+Main computation framework is tensorflow~\cite{tensorflow2015-whitepaper}
+
+\section{Data preprocessing}
+
 \chapter{Results}

### 2017-04-29 src/input_readers.py
Moved to RNN nanocall bootstrapping
Rewritten aligning code for blocks.
Moved to edlib de facto bindings.

@@ -27,52 +27,38 @@ class AlignedRaw(InputReader):
     def preprocessSignal(self, signal):
         return (signal - 646.11133) / 75.673653
 
-    def read_fast5_raw_ref(self, fast5_path, ref_path, block_size_x, block_size_y, num_blocks, warn_if_short=False):
+    def read_fast5_raw_ref(self, fast5_path, ref_path, block_size_x, block_size_y, num_blocks, warn_if_short=False, verify_file=True):
         num_blocks += 1
         ref_ext = os.path.splitext(ref_path)[1]
         with h5py.File(fast5_path, 'r') as h5, open(ref_path, 'r') as ref_file:
             reads = h5['Analyses/EventDetection_000/Reads']
             target_read = list(reads.keys())[0]
-            events = np.array(reads[target_read + '/Events'])
-            start_time = events['start'][0]
-            start_pad = int(start_time - h5['Raw/Reads/' + target_read].attrs['start_time'])
+            sampling_rate = h5['UniqueGlobalKey/channel_id'].attrs['sampling_rate']
 
-            basecalled_events = h5['/Analyses/Basecall_1D_000/BaseCalled_template/Events']
-            basecalled = np.array(basecalled_events.value[['mean', 'stdv', 'model_state', 'move', 'start', 'length']])
+            try:
+                basecalled_events = h5['/Analyses/Basecall_RNN_1D_000/BaseCalled_template/Events']
+            except:
+                print(fast5_path, "Not found RNN component")
+                raise util.AligmentError()
+
+            basecalled = np.array(basecalled_events.value[['start', 'length', 'model_state', 'move']])
 
             signal = h5['Raw/Reads/' + target_read]['Signal']
-            signal_len = h5['Raw/Reads/' + target_read].attrs['duration'] - start_pad
+            start_time = basecalled[0]['start']
+            start_pad = int(sampling_rate * basecalled[0]['start'])
+            signal_len = int(sampling_rate * (basecalled[-1]['start'] + basecalled[-1]['length'] - basecalled[0]['start']))
 
-            x = np.zeros([block_size_x * num_blocks, 1], dtype=np.float32)
-            x_len = min(signal_len, block_size_x * num_blocks)
-            x[:x_len, 0] = signal[start_pad:start_pad + x_len]
+            np.testing.assert_allclose(len(signal), start_pad + signal_len, rtol=1e-2)  # Within 1% relative tolerance
 
[…]

### 2017-04-29 src/util.py

@@ -1,6 +1,8 @@
-from edlib import Edlib
+import edlib
 import h5py
 import numpy as np
+import re
+import unittest
 
 
 def next_num(prev, symbol):
@@ -25,126 +27,6 @@ def read_fasta(fp):
             return rr(f)
     else:
         return rr(fp)
-
-def read_fast5(filename, block_size, num_blocks, warn_if_short=False):
-    'Read fast5 file.'
-    with h5py.File(filename, 'r') as h5:
-        reads = h5['Analyses/EventDetection_000/Reads']
-        events = np.array(reads[list(reads.keys())[0] + '/Events'])
-
-        basecalled_events = h5['/Analyses/Basecall_1D_000/BaseCalled_template/Events']
-        basecalled = np.array(basecalled_events.value[['mean', 'stdv', 'model_state', 'move', 'start', 'length']])
-
-        length = block_size * num_blocks
-        if warn_if_short and events.shape[0] < length:
-            print("WARNING...less then truncate events", filename)
-
-        # x[i] is feat values for event #i
-        x = np.zeros([length, 3], dtype=np.float32)
-        x_len = min(length, events.shape[0])
-
-        # y[2*i] and y[2*i + 1] are bases for event #i
-        y = np.zeros([length], dtype=np.uint8)
-        y_len = np.zeros([num_blocks], dtype=np.int32)
-
-        bcall_idx = 0
-        prev, curr_sec = "N", 0
-        for i, e in enumerate(events[:length]):
-            if i // block_size > curr_sec:
-                prev, curr_sec = "N", i // block_size
-
-            if bcall_idx < basecalled.shape[0]:
-                b = basecalled[bcall_idx]
-                if b[0] == e[2] and b[1] == e[3]:  # mean == mean and stdv == stdv
-                    add_chr = []
-                    if bcall_idx == 0:
-                        add_chr.extend(list(b[2].decode("ASCII")))  # initial model state
-                    bcall_idx += 1
-                    if b[3] == 1:
-                        add_chr.append(chr(b[2][-1]))
-                    if b[3] == 2:
-                        add_chr.append(chr(b[2][-2]))
-                        add_chr.append(chr(b[2][-1]))
-                    # print(add_chr)
-                    for c in add_chr:
-                        prev, sym = next_num(prev, c)
[…]

### 2017-04-29 src/util.py
Moved to RNN nanocall bootstrapping

@@ -26,67 +26,6 @@ def read_fasta(fp):
     else:
         return rr(fp)
 
-def read_fast5(filename, block_size, num_blocks, warn_if_short=False):
-    'Read fast5 file.'
-    with h5py.File(filename, 'r') as h5:
-        reads = h5['Analyses/EventDetection_000/Reads']
-        events = np.array(reads[list(reads.keys())[0] + '/Events'])
-
-        basecalled_events = h5['/Analyses/Basecall_1D_000/BaseCalled_template/Events']
-        basecalled = np.array(basecalled_events.value[['mean', 'stdv', 'model_state', 'move', 'start', 'length']])
-
-        length = block_size * num_blocks
-        if warn_if_short and events.shape[0] < length:
-            print("WARNING...less then truncate events", filename)
-
-        # x[i] is feat values for event #i
-        x = np.zeros([length, 3], dtype=np.float32)
-        x_len = min(length, events.shape[0])
-
-        # y[2*i] and y[2*i + 1] are bases for event #i
-        y = np.zeros([length], dtype=np.uint8)
-        y_len = np.zeros([num_blocks], dtype=np.int32)
-
-        bcall_idx = 0
-        prev, curr_sec = "N", 0
-        for i, e in enumerate(events[:length]):
-            if i // block_size > curr_sec:
-                prev, curr_sec = "N", i // block_size
-
-            if bcall_idx < basecalled.shape[0]:
-                b = basecalled[bcall_idx]
-                if b[0] == e[2] and b[1] == e[3]:  # mean == mean and stdv == stdv
-                    add_chr = []
-                    if bcall_idx == 0:
-                        add_chr.extend(list(b[2].decode("ASCII")))  # initial model state
-                    bcall_idx += 1
-                    if b[3] == 1:
-                        add_chr.append(chr(b[2][-1]))
-                    if b[3] == 2:
-                        add_chr.append(chr(b[2][-2]))
-                        add_chr.append(chr(b[2][-1]))
-                    # print(add_chr)
-                    for c in add_chr:
-                        prev, sym = next_num(prev, c)
-                        y[curr_sec * block_size + y_len[curr_sec]] = sym
-                        y_len[curr_sec] += 1
-                    if y_len[curr_sec] > block_size:
-                        print("Too many events in block!")
-                        return None
-
[…]

### 2017-04-30 src/errors.py
Almost finalized migrating input_readers

@@ -1,0 +1,18 @@
+class MinIONBasecallerException(Exception):
+    pass
+
+
+class TooLargeEditDistance(MinIONBasecallerException):
+    pass
+
+
+class MissingRNN1DBasecall(MinIONBasecallerException):
+    pass
+
+
+class BlockSizeYTooSmall(MinIONBasecallerException):
+    pass
+
+
+class InsufficientDataBlocks(MinIONBasecallerException):
+    pass

### 2017-04-30 src/input_readers.py

@@ -1,8 +1,7 @@
 import os
 import socket
 from random import shuffle
-from scipy.signal import medfilt
-
+from errors import MissingRNN1DBasecall, InsufficientDataBlocks, MinIONBasecallerException
 import numpy as np
 import util
 import h5py
@@ -27,10 +26,8 @@ class AlignedRaw(InputReader):
     def preprocessSignal(self, signal):
         return (signal - 646.11133) / 75.673653
 
-    def read_fast5_raw_ref(self, fast5_path, ref_path, block_size_x, block_size_y, num_blocks, warn_if_short=False, verify_file=True):
-        num_blocks += 1
-        ref_ext = os.path.splitext(ref_path)[1]
-        with h5py.File(fast5_path, 'r') as h5, open(ref_path, 'r') as ref_file:
+    def read_fast5(self, fast5_path):
+        with h5py.File(fast5_path, 'r') as h5:
             reads = h5['Analyses/EventDetection_000/Reads']
             target_read = list(reads.keys())[0]
             sampling_rate = h5['UniqueGlobalKey/channel_id'].attrs['sampling_rate']
@@ -38,8 +35,7 @@ class AlignedRaw(InputReader):
             try:
                 basecalled_events = h5['/Analyses/Basecall_RNN_1D_000/BaseCalled_template/Events']
             except:
-                print(fast5_path, "Not found RNN component")
-                raise util.AligmentError()
+                MissingRNN1DBasecall("{} Not found RNN component".format(fast5_path))
 
             basecalled = np.array(basecalled_events.value[['start', 'length', 'model_state', 'move']])
 
@@ -50,15 +46,40 @@ class AlignedRaw(InputReader):
 
             np.testing.assert_allclose(len(signal), start_pad + signal_len, rtol=1e-2)  # Within 1% relative tolerance
 
+            basecalled['start'] -= start_time
+            signal = signal[start_pad:start_pad + signal_len]
+            signal = self.preprocessSignal(signal)
+            fastq = h5['/Analyses/Basecall_RNN_1D_000/BaseCalled_template/Fastq'][()].decode().split('\n')
+
+            return {
+                'signal': signal,
+                'basecalled': basecalled,
+                'sampling_rate': sampling_rate,
+                'fastq': fastq
+            }
+
+    def read_fast5_raw_ref(self, fast5_path, ref_path, block_size_x, block_size_y, num_blocks, verify_file=True):
+        # num_blocks += 1
+        ref_ext = os.path.splitext(ref_path)[1]
[…]

### 2017-04-30 src/util.py

@@ -1,8 +1,8 @@
 import edlib
-import h5py
 import numpy as np
 import re
 import unittest
+from errors import TooLargeEditDistance, BlockSizeYTooSmall
 
 
 def next_num(prev, symbol):
@@ -65,10 +65,6 @@ def decode_example(Y, Y_len, num_blocks, block_size_y, pad=None):
     return list(map(lambda x: x.ljust(pad, ' '), gg))
 
 
-class AligmentError(Exception):
-    pass
-
-
 def correct_basecalled(bucketed_basecall, reference, nedit_tol=0.2):
     basecalled = "".join(bucketed_basecall)
     origin = np.zeros(len(basecalled), dtype=np.int32)
@@ -80,7 +76,7 @@ def correct_basecalled(bucketed_basecall, reference, nedit_tol=0.2):
     result_set = edlib.align(basecalled, reference, task="path")
     nedit = result_set['editDistance'] / len(reference)
     if nedit > nedit_tol:
-        raise RuntimeWarning("Normalized edit distance is large...%.3f" % nedit)
+        raise TooLargeEditDistance("Normalized edit distance is large...%.3f" % nedit)
 
     result = ["" for _ in bucketed_basecall]
     idx_ref = 0
@@ -102,6 +98,23 @@ def correct_basecalled(bucketed_basecall, reference, nedit_tol=0.2):
                 idx_ref += 1
     assert "".join(result) == reference
     return result
+
+
+def prepare_y(bucketed_basecall, block_size_y):
+    num_blocks = len(bucketed_basecall)
+    y = np.zeros([num_blocks * block_size_y], dtype=np.uint8)
+    y_len = np.zeros([num_blocks], dtype=np.int32)
+
+    prev = "N"
+    for i, seq in enumerate(bucketed_basecall):
+        y_len[i] = len(seq)
+        if y_len[i] > block_size_y:
+            raise BlockSizeYTooSmall("On block {}, got {}".format(i, y_len[i]))
+
+        for j, c in enumerate(seq):
+            prev, y[i * block_size_y + j] = next_num(prev, c)
+
+    return y, y_len
 
 
 class TestCorrectedBasecalled(unittest.TestCase):
@@ -135,23 +148,6 @@ def sigopt_int(name, min, max):
 
 def sigopt_double(name, min, max):
     return sigopt_numeric('double', name, min, max)
-
-
-def get_raw_signal(fast5_path):
-    """
-        Return 1D raw signal from fast5_path. No preprocessing
-    """
[…]

### 2017-04-30 src/errors.py
Finished migrationg, on to testing

@@ -16,3 +16,7 @@ class BlockSizeYTooSmall(MinIONBasecallerException):
 
 class InsufficientDataBlocks(MinIONBasecallerException):
     pass
+
+
+class ZeroLenY(MinIONBasecallerException):
+    pass

### 2017-05-04 mincall/util.py
Performance issue addressed

@@ -107,7 +107,6 @@ def correct_basecalled(bucketed_basecall, reference, nedit_tol=0.2):
             elif op in CIGAR_DELETION:
                 result[origin[idx_bcalled]] += reference[idx_ref]
                 idx_ref += 1
-    assert "".join(result) == reference
     return result
 
 

### 2017-05-04 mincall/align_utils.py
Demo notebook for evaluation, refacotred some utils

@@ -1,0 +1,48 @@
+import logging
+import subprocess
+
+import pysam
+
+from mincall.bioinf_utils import reverse_complement, decompress_cigar_pairs
+
+
+def get_target_sequences(sam_out):
+    result_dict = {}
+
+    with pysam.AlignmentFile(sam_out, "r") as samfile:
+        for x in samfile.fetch():
+            name = x.query_name
+
+            if x.is_unmapped:
+                logging.warning("%s unmapped" % name)
+                continue
+            try:
+                target = x.get_reference_sequence()
+            except ValueError:
+                logging.error("%s Mapped but reference len equals 0" % name)
+                continue
+
+            ref_name = x.reference_name
+            length = x.reference_length
+            start_pos = x.reference_start
+            cigar_pairs = x.cigartuples
+
+            if x.is_reverse:
+                target = reverse_complement(target)
+                cigar_pairs = list(reversed(cigar_pairs))
+
+            cigar_str = decompress_cigar_pairs(cigar_pairs, mode='ints')
+            result_dict[name] = [target, ref_name, start_pos, length, cigar_str]
+    return result_dict
+
+
+def align_with_graphmap(reads_path, ref_path, is_circular, out_sam):
+    args = ["graphmap", "align", "-r", ref_path, "-d", reads_path, "-o", out_sam, "-v", "0", "--extcigar"]
+    if is_circular:
+        args.append("-C")
+
+    exit_status = subprocess.call(args)
+    if exit_status != 0:
+        logging.warning("Graphmap exit status %d" % exit_status)
+
+

### 2017-05-05 mincall/align_utils.py
Consensus from isovic added, demo notebook for evaluation

@@ -1,6 +1,6 @@
 import logging
 import subprocess
-
+import math
 import pysam
 
 from mincall.bioinf_utils import reverse_complement, decompress_cigar_pairs
@@ -45,4 +45,26 @@ def align_with_graphmap(reads_path, ref_path, is_circular, out_sam):
     if exit_status != 0:
         logging.warning("Graphmap exit status %d" % exit_status)
 
+
+def filter_aligments_in_sam(sam_path, out_path, filters=[]):
+    n_reads = 0
+    n_kept = 0
+
+    with pysam.AlignmentFile(sam_path, "r") as in_sam:
+        with pysam.AlignmentFile(out_path, "w", template=in_sam) as out_sam:
+
+            for x in in_sam.fetch():
+                n_reads += 1
+
+                if all([f(x) for f in filters]):
+                    n_kept += 1
+                    out_sam.write(x)
+    return n_kept, n_reads - n_kept
+
+
+def read_len_filter(min_len=-1, max_len=math.inf):
+    def _filter(aligment):
+        return min_len < aligment.query_length < max_len
+
+    return _filter
 

### 2017-05-05 mincall/bioinf_utils.py

@@ -92,7 +92,7 @@ def get_ref_len_from_cigar(cigar_pairs):
     ref_len = 0
 
     for b, cnt in cigar_pairs:
-        sym = cigar_int_to_c(b)
+        sym = cigar_int_to_c(b) if isinstance(b, int) else b
         if sym in CIGAR_MATCH_MISSMATCH or sym in CIGAR_DELETION:
             ref_len += cnt
     return ref_len
@@ -221,7 +221,6 @@ def error_positions_report(sam_path, n_buckets=1000):
     df.sort_values(by='relative_position', inplace=True)
     df.reset_index(inplace=True, drop=True)
     return df
-
 
 
 def read_fasta(fp):

### 2017-05-05 mincall/align_utils.py
Updated dependencies, added merge_sam util method

@@ -2,7 +2,8 @@ import logging
 import subprocess
 import math
 import pysam
-
+import os
+import glob
 from mincall.bioinf_utils import reverse_complement, decompress_cigar_pairs
 
 
@@ -67,4 +68,9 @@ def read_len_filter(min_len=-1, max_len=math.inf):
         return min_len < aligment.query_length < max_len
 
     return _filter
+
+def merge_sam_files(sam_dir_path, out_sam_path):
+    sam_files = glob.glob(os.path.join(sam_dir_path, '*.sam'))
+    pysam.merge('-f', out_sam_path, *sam_files, catch_stdout=False)
+
 

### 2017-05-05 mincall/preprocess_dataset.py

@@ -12,7 +12,6 @@ import pysam
 from tqdm import tqdm
 from mincall import align_utils, bioinf_utils
 
-
 load_dotenv(find_dotenv())
 
 

### 2017-05-05 paper/main.tex
Small incremental update

@@
 
+Dataset were obtained from (Ask Marko!).
+In this research models were trainined on ecoli K-12 strand.
+(Should I cite someone for strand).
+Data was split into train and test subset, such that aligned reads map to different reference genome parts.
+For initial model data bootstrapping metrichorn was used, version (Ask Someone?).
+
+The fast5 input training files were further split into smaller training blocks, constiting of fixed block size on raw signal. For each block target sequence basecalled data is used in following way.
+
+We're using basecalled knowledge which tells us on each raw read part which 6-mer were currently in the nanopore.
+Using this data, we get the intermediate target sequence for each block.
+To correc for model errors, we use aligned information, that is reference genome and alignment data(cigar string) to correct those information, yielding finalized target sequence.
+For further performace we skip first and last traning block, since most error aggregate on edges.
+
+Adjacent matching bases are spearated with surogate one, for example: AAA -> AA'A for reasong described in following section.
+
+\section{Residual arhitecture}
+
+\section{CTC}
+
+This section gives brief overview over CTC, for further detail and in depth explaintion we recommend original paper~\cite{graves2006connectionist} or searhcing for contemporary blog posts.
+
+In CTC we have target sequecne $\mathbf{l}$ consisting from symbols of alphabet $\Sigma$. Our models outputs $\mathbf{x}$, and CTC maximizes $p(\mathbf{l}|\mathbf{x})$.
+
+$\mathbf{x}$ is consistend of $n$ independent discrete random variables, $X_i$  over domain $\Sigma \cup \text{Blank}$.
+Path $p$ through $\mathbf{x}$ is assigning for each $X_i$ single value.
+It's log probabity is $\log P(p) = \sum_i {\log P(X_i=x_i)}$ where $x_i$s define the path.
+
+We further have merging operator on path, $\text{merge}(x_1, x_2, \ldots, x_n)$ which merges together adjacent equivalent elements.
+Then $p(\mathbf{l}|\mathbf{x}) = \sum_{p\in\text{paths}(\mathbf{x}, \text{merged}(p) = \mathbf{l}}{P(p)}$ or less formally, $P(\mathbf{l}|\mathbf{x})$ is sum of all path probabilites which when merged give $\mathbf{l}$
+
+In out concrete $x$ aplication $\Sigma = \{\text{A G T C A' G' T' A'}\}$
+
+For decoding we use beam search decoder, with beam width 100.
+
 \chapter{Results}
@@
 Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+
 \chapter{Conclusion and further work}

### 2017-05-05 paper/main.tex
Update

@@
 
+We use gated residual network, that is layer function is like this: $f(x) = k g(x) + x$ Where $x$ is input layer and $k$ constant learnt during training.
+Specifically $g$ used is multiple, 1--3 times in our experiments, Relu-BatchNorm-CNN layers.
+
 \section{CTC}
@@
 
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+TODO.
 

### 2017-05-05 paper/main.tex
Update with conclusion.

@@
 
-In CTC we have target sequecne $\mathbf{l}$ consisting from symbols of alphabet $\Sigma$. Our models outputs $\mathbf{x}$, and CTC maximizes $p(\mathbf{l}|\mathbf{x})$.
+In CTC we have target sequecne $\mathbf{l}$ consisting from symbols of alphabet $\Sigma$. Our models outputs $\mathbf{x}$, and CTC loss maximizes $p(\mathbf{l}|\mathbf{x})$.
 
@@
 
-We further have merging operator on path, $\text{merge}(x_1, x_2, \ldots, x_n)$ which merges together adjacent equivalent elements.
+We further have merging operator on path, $\text{merge}(x_1, x_2, \ldots, x_n)$ which merges together adjacent equivalent elements. For example $\text{merge}(AAAGC) = AGC$.
+
 Then $p(\mathbf{l}|\mathbf{x}) = \sum_{p\in\text{paths}(\mathbf{x}, \text{merged}(p) = \mathbf{l}}{P(p)}$ or less formally, $P(\mathbf{l}|\mathbf{x})$ is sum of all path probabilites which when merged give $\mathbf{l}$
@@
 \chapter{Conclusion and further work}
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
-% \cite{deepnano}
+
+This model used advance state-of-the-art gated residual convolutional neural network, with top models having 270 layers and over 3M parameters, yet improvements over Metrichorn baseline are mariginal.
+As conclusion it might be that we've reached Bayesian error rate for R9 chemistry.
+Furthermore R9.5 and 1D^2 reads are under development which shall yield this paper's result obsolite quite soon, yet underlying code developed could easily be adjusted, and trained on new data.
+
+Unlike Nanonet which uses custom OpenCL kernels, or Albacore -- a novel ONT basecaller as of May 2017 lacking GPU support, this work used world class computational framework tensorflow with highly optimized kernels and large development community.
+Therefore resulting paper's effect is showcasing Residual CNN approach or pure CNN approach with CTC loss is marginally better than already established basecaller, and providing code in contemporary framework.
+
 % \bibliographystyle{fer}

### 2017-05-05 paper/main.tex
Small changes...now it build the paper

@@
-\documentclass[times, utf8, seminar]{fer}
+\documentclass[times, utf8, seminar, numeric]{fer}
 \usepackage{booktabs}
-
 \usepackage{mathtools, amsmath,amsfonts,amssymb, amsthm}
@@
 \voditelj{Mile Sikić}
-\title{CNNNano --- MinION  basecaller pogonjen dubokim ucenjem}
-\engtitle{CNNNano --- MinION deep learning powered basecaller}
+\title{MinCall --- MinION  basecaller pogonjen dubokim ucenjem}
+\engtitle{MinCall --- MinION deep learning powered basecaller}
 \author{Neven Miculinić}
@@
 As conclusion it might be that we've reached Bayesian error rate for R9 chemistry.
-Furthermore R9.5 and 1D^2 reads are under development which shall yield this paper's result obsolite quite soon, yet underlying code developed could easily be adjusted, and trained on new data.
+Furthermore R9.5 and 1D\^2 reads are under development which shall yield this paper's result obsolite quite soon, yet underlying code developed could easily be adjusted, and trained on new data.
 
@@
 
-% \bibliographystyle{fer}
+\bibliographystyle{fer}
+\bibliography{refs}
 % \bibliographystyle{abbrv}
-% \bibliography{refs}
 

### 2017-05-05 paper/main.tex
Whatever

@@
 \voditelj{Mile Sikić}
-\title{MinCall --- MinION  basecaller pogonjen dubokim ucenjem}
-\engtitle{MinCall --- MinION deep learning powered basecaller}
+\title{MinCall --- MinION od raw signala do genoma}
+\engtitle{MinCall --- MinION end2end deep learning basecaller}
 \author{Neven Miculinić}
@@
 \begin{sazetak}
-TODO
 
-\kljucnerijeci{TODO}
+\kljucnerijeci{Basecaller, MinION, sekvenciniranje sljedece generacije}
 \end{sazetak}
@@
     Whole process is trainned end-to-end from raw signal on R9 sequecing pore and once trainned it's independent from Metrichor.
-\keywords{}
+\keywords{Basecaller, MinION, R9, CNN, CTC, Next generation sequecing}
 \end{abstract}

### 2017-05-05 paper/main.tex
Added space after colon.

@@
 
-In this paper, we introduce the base caller for the MinION nanopore sequencing platform~\cite{mikheyev2014first}.The MinION device by Oxford Nanopore, weighing only 90 grams, is currently the smallest high-throughput DNA sequencer.
+In this paper, we introduce the base caller for the MinION nanopore sequencing platform~\cite{mikheyev2014first}.
+The MinION device by Oxford Nanopore, weighing only 90 grams, is currently the smallest high-throughput DNA sequencer.
 Thanks to its low capital costs, small size and the possibility of analyzing the data in real time as they are produced, MinION is very promising for clinical applications, such as monitoring infectious disease outbreaks~\cite{judge2015early}\cite{quick2016real}, characterizing structural variants in cancer\cite{norris2016nanopore} and even full human genome assembly~\cite{jain2017nanopore}.

### 2017-05-05 mincall/controller.py
Added try/catch around basecaller

@@ -102,19 +102,22 @@ def basecall(create_test_model, **kwargs):
         total_bases = 0
         pbar = tqdm(file_list, unit='reads', unit_scale=True, dynamic_ncols=True)
         for f in pbar:
-            if args.out_dir is not None:
-                out = os.path.splitext(f)[0].split('/')[-1] + ".fasta"
-                out = os.path.join(args.out_dir, out)
-            else:
-                out = None
-            t0 = perf_counter()
-            basecalled = model.basecall_sample(f, fasta_out=out, write_logits=args.write_logits)
-            total_time += perf_counter() - t0
-            total_bases += len(basecalled)
-            pbar.set_postfix(speed="{:.3f} b/s".format(total_bases / total_time))
+            try:
+                if args.out_dir is not None:
+                    out = os.path.splitext(f)[0].split('/')[-1] + ".fasta"
+                    out = os.path.join(args.out_dir, out)
+                else:
+                    out = None
+                t0 = perf_counter()
+                basecalled = model.basecall_sample(f, fasta_out=out, write_logits=args.write_logits)
+                total_time += perf_counter() - t0
+                total_bases += len(basecalled)
+                pbar.set_postfix(speed="{:.3f} b/s".format(total_bases / total_time))
 
-            if out is None:
-                util.dump_fasta(os.path.splitext(f)[0].split(os.sep)[-1], basecalled, sys.stdout)
+                if out is None:
+                    util.dump_fasta(os.path.splitext(f)[0].split(os.sep)[-1], basecalled, sys.stdout)
+            except Exception as ex:
+                model.logger.error("Error happened in %s", f, exc_info=True)
 
     finally:
         model.close_session()

### 2017-05-05 mincall/align_utils.py
util methods

@@ -43,6 +43,8 @@ def align_with_graphmap(reads_path, ref_path, is_circular, out_sam):
         args.append("-C")
 
     exit_status = subprocess.call(args)
+    logging.info("Graphmap exit status %d" % exit_status)
+
     if exit_status != 0:
         logging.warning("Graphmap exit status %d" % exit_status)
 
@@ -69,8 +71,22 @@ def read_len_filter(min_len=-1, max_len=math.inf):
 
     return _filter
 
+
 def merge_sam_files(sam_dir_path, out_sam_path):
     sam_files = glob.glob(os.path.join(sam_dir_path, '*.sam'))
     pysam.merge('-f', out_sam_path, *sam_files, catch_stdout=False)
 
+
+def merge_reads(reads_fastx_root, out_fastx_path):
+    dir_content = [os.path.join(reads_fastx_root, f) for f in os.listdir(reads_fastx_root)]
+    files = filter(os.path.isfile, dir_content)
+
+    def _copy_to(outfp, input_file_path):
+        with open(input_file_path, 'r') as fin:
+            for line in fin:
+                outfp.write(line)
+
+    with open(out_fastx_path, 'w') as fout:
+        for in_path in files:
+            _copy_to(fout, in_path)
 

### 2017-05-06 mincall/align_utils.py
Util methods, circular aligment - 2 samlines bug fixed

@@ -4,13 +4,13 @@ import math
 import pysam
 import os
 import glob
-from mincall.bioinf_utils import reverse_complement, decompress_cigar_pairs
+from mincall.bioinf_utils import reverse_complement, decompress_cigar_pairs, rtrim_cigar, ltrim_cigar
 
 
-def get_target_sequences(sam_out):
+def get_target_sequences(sam_path):
     result_dict = {}
 
-    with pysam.AlignmentFile(sam_out, "r") as samfile:
+    with pysam.AlignmentFile(sam_path, "r") as samfile:
         for x in samfile.fetch():
             name = x.query_name
 
@@ -33,11 +33,56 @@ def get_target_sequences(sam_out):
                 cigar_pairs = list(reversed(cigar_pairs))
 
             cigar_str = decompress_cigar_pairs(cigar_pairs, mode='ints')
+
+            if name in result_dict:
+                prev_target, _, prev_start_pos, _, prev_cigar_str = result_dict[name]
+                merged = _merge_circular_aligment(prev_target, prev_start_pos, prev_cigar_str,
+                                                  target, start_pos, cigar_str, x.is_reverse, x.query_name)
+                if not merged:
+                    continue
+
+                target, start_pos, cigar_str = merged
+                length = len(target)
+
             result_dict[name] = [target, ref_name, start_pos, length, cigar_str]
     return result_dict
 
 
+def _merge_circular_aligment(target_1, start_pos_1, cigar_str_1,
+                             target_2, start_pos_2, cigar_str_2, is_reversed, qname):
+
+    if is_reversed:
+        # reverse back both
+        cigar_str_1 = ''.join(reversed(cigar_str_1))
+        target_1 = reverse_complement(target_1)
+
+        cigar_str_2 = ''.join(reversed(cigar_str_2))
+        target_2 = reverse_complement(target_2)
+
+    if start_pos_1 == 0:
+        start = start_pos_2
+        cigar = rtrim_cigar(cigar_str_2) + ltrim_cigar(cigar_str_1)
+        target = target_2 + target_1
+
+    elif start_pos_2 == 0:
+        start = start_pos_1
+        cigar = rtrim_cigar(cigar_str_1) + ltrim_cigar(cigar_str_2)
+        target = target_1 + target_2
+
+    else:
+        # not circular, duplicate
[…]

### 2017-05-06 mincall/bioinf_utils.py

@@ -1,6 +1,7 @@
 import itertools
 import logging
 import pysam
+import re
 import csv
 import numpy as np
 from collections import Counter, defaultdict
@@ -26,6 +27,7 @@ CIGAR_INSERTION = 'ISH'
 CIGAR_DELETION = 'DNP'
 CIGAR_SOFTCLIP = ['S']
 CIGAR_HARDCLIP = ['H']
+CIGAR_CLIP = 'SH'
 
 BYTE_TO_CIGAR = {v: k for k, v in CIGAR_TO_BYTE.items()}
 
@@ -86,6 +88,32 @@ def decompress_cigar_pairs(cigar_pairs, mode='ints'):
 def decompress_cigar(cigar_str):
     cigar_pairs = cigar_str_to_pairs(cigar_str)
     return decompress_cigar_pairs(cigar_pairs, mode='chars')
+
+
+def compress_cigar(cigar_str):
+    pairs = []
+    count = 1
+
+    for i in range(1, len(cigar_str)):
+        if cigar_str[i-1] == cigar_str[i]:
+            count += 1
+        else:
+            pairs.append((cigar_str[i-1], count))
+            count = 1
+    pairs.append((cigar_str[-1], count))
+    return pairs
+
+
+def ltrim_cigar(cigar):
+    for op in CIGAR_CLIP:
+        cigar = re.sub(r'^%s*' % op, r'', cigar)
+    return cigar
+
+
+def rtrim_cigar(cigar):
+    for op in CIGAR_CLIP:
+        cigar = re.sub(r'%s*$' % op, r'', cigar)
+    return cigar
 
 
 def get_ref_len_from_cigar(cigar_pairs):

### 2017-05-06 mincall/align_utils.py
VCF stats analysis started

@@ -123,6 +123,12 @@ def secondary_aligments_filter():
     return _filter
 
 
+def only_mapped_filter():
+    def _filter(x):
+        return not x.is_unmapped
+    return _filter
+
+
 def merge_sam_files(sam_dir_path, out_sam_path):
     sam_files = glob.glob(os.path.join(sam_dir_path, '*.sam'))
     os.makedirs(os.path.dirname(out_sam_path), exist_ok=True)

### 2017-05-06 mincall/bioinf_utils.py

@@ -208,11 +208,12 @@ def error_rates_for_sam(sam_path):
     all_errors = []
     with pysam.AlignmentFile(sam_path, "r") as samfile:
         for x in samfile.fetch():
+            if x.is_unmapped:
+                logging.error("%s is unmapped", x.query_name)
+                continue
+
             qname = x.query_name
             cigar_pairs = x.cigartuples
-            if not cigar_pairs:
-                logging.error("%s No cigar string found", x.query_name)
-                continue
 
             full_cigar = decompress_cigar_pairs(cigar_pairs, mode='ints')
             all_errors.append([qname] + list(error_rates_from_cigar(full_cigar)))

### 2017-05-07 mincall/tools/analyse_bc.py
Initial commit for analysie_bc script

@@ -1,0 +1,101 @@
+import os
+import pandas as pd
+import logging
+import subprocess
+import argparse
+import mincall.align_utils as align_utils
+from mincall.align_utils import filter_aligments_in_sam, read_len_filter
+from mincall.bioinf_utils import error_rates_for_sam, error_positions_report, CIGAR_OPERATIONS
+import seaborn as sns
+import matplotlib.pyplot as plt
+from mincall.consensus import get_consensus_report
+
+
+log_fmt = '\r[%(levelname)s] %(name)s: %(message)s'
+logging.basicConfig(level=logging.INFO, format=log_fmt)
+
+args = argparse.ArgumentParser()
+args.add_argument("input_folder", help="Input fast5 folder to basecall")
+args.add_argument("out_folder", help="Output folder for all analysis")
+args.add_argument("--ref", help="Reference genome")
+args.add_argument("--min_length", type=int, default=500, help="Minimum read lenght for further analysis")
+args.add_argument("-c", "--circular", help="Is genome circular", action="store_true")
+args.add_argument("--coverage_threshold", help="Minimal coverage threshold for consensus", type=float, default=0.0)
+args = args.parse_args()
+print(args, dir(args))
+
+basecallers = {
+    # "mincall_m270": ["nvidia-docker", "run", "--rm", "-v", "%s:/data" % args.input_folder, "-u=%d" % os.getuid(), "nmiculinic/mincall:9947283"],
+    # "nanonet": ["nanonetcall", args.input_folder, "--chemistry", "r9", "--platforms", "nvidia:0:20", "--exc_opencl"]
+    "nanonet": ["nanonetcall", args.input_folder, "--chemistry", "r9", "--jobs", "8"]
+}
+
+os.makedirs(args.out_folder, exist_ok=True)
+dfs = {}
+for name, cmd in basecallers.items():
+    fasta_path = os.path.join(args.out_folder, name + ".fa")
+    if os.path.isfile(fasta_path):
+        logging.info("%s exists, skipping", fasta_path)
+    else:
+        logging.info("Basecalling %s with %s", args.input_folder, name)
+        logging.info("Full command: %s", " ".join(cmd))
+        with open(fasta_path, 'w') as f:
+            subprocess.check_call(cmd, stdout=f)
+
+    sam_path = os.path.join(args.out_folder, name + ".sam")
+    if os.path.isfile(sam_path):
[…]

### 2017-05-07 mincall/tools/analyse_bc.py
Fixed some save paths

@@ -22,7 +22,6 @@ args.add_argument("--min_length", type=int, default=500, help="Minimum read leng
 args.add_argument("-c", "--circular", help="Is genome circular", action="store_true")
 args.add_argument("--coverage_threshold", help="Minimal coverage threshold for consensus", type=float, default=0.0)
 args = args.parse_args()
-print(args, dir(args))
 
 basecallers = {
     # "mincall_m270": ["nvidia-docker", "run", "--rm", "-v", "%s:/data" % args.input_folder, "-u=%d" % os.getuid(), "nmiculinic/mincall:9947283"],
@@ -38,6 +37,7 @@ for name, cmd in basecallers.items():
         logging.info("%s exists, skipping", fasta_path)
     else:
         logging.info("Basecalling %s with %s", args.input_folder, name)
+        logging.info("Output file", fasta_path)
         logging.info("Full command: %s", " ".join(cmd))
         with open(fasta_path, 'w') as f:
             subprocess.check_call(cmd, stdout=f)
@@ -63,12 +63,13 @@ for name, cmd in basecallers.items():
         logging.info("%s file exists, loading", reads_csv)
         df = pd.read_csv(reads_csv)
     else:
-        df = error_rates_for_sam(reads_csv)
+        df = error_rates_for_sam(filtered_sam)
         df.to_csv(reads_csv)
         desc = df.describe()
-        with open(os.path.join(args.out_folder, name + "_read_data.tex"), 'w') as f:
+        with open(os.path.join(args.out_folder, name + "_read_summary.tex"), 'w') as f:
             desc.to_latex(f)
-        desc.to_csv(os.path.join(args.out_folder, name + "_read_summary.csv"))
+        with open(os.path.join(args.out_folder, name + "_read_summary.txt"), 'w') as f:
+            desc.to_string(f)
     logging.info("%s\n%s", name, df.describe())
     dfs[name] = df
 
@@ -78,10 +79,12 @@ for name, cmd in basecallers.items():
     else:
         consensus_report = get_consensus_report(filtered_sam, args.ref, args.coverage_threshold)
         consensus_report.to_csv(consensus_report_path)
+        with open(os.path.splitext(consensus_report_path)[0] + ".txt", 'w') as f:
+            consensus_report.to_string(f)
 
 
-fig_kde_path = os.path.join(args.out_dir, name + "reads_kde.png")
[…]

### 2017-05-07 mincall/tools/analyse_bc.py
Fixed script bugs

@@ -32,35 +32,36 @@ basecallers = {
 os.makedirs(args.out_folder, exist_ok=True)
 dfs = {}
 for name, cmd in basecallers.items():
+    logger = logging.getLogger(name)
     fasta_path = os.path.join(args.out_folder, name + ".fa")
     if os.path.isfile(fasta_path):
-        logging.info("%s exists, skipping", fasta_path)
+        logger.info("%s exists, skipping", fasta_path)
     else:
-        logging.info("Basecalling %s with %s", args.input_folder, name)
-        logging.info("Output file", fasta_path)
-        logging.info("Full command: %s", " ".join(cmd))
+        logger.info("Basecalling %s with %s", args.input_folder, name)
+        logger.info("Output file", fasta_path)
+        logger.info("Full command: %s", " ".join(cmd))
         with open(fasta_path, 'w') as f:
             subprocess.check_call(cmd, stdout=f)
 
     sam_path = os.path.join(args.out_folder, name + ".sam")
     if os.path.isfile(sam_path):
-        logging.info("%s exists, skipping", sam_path)
+        logger.info("%s exists, skipping", sam_path)
     else:
-        logging.info("Aligning %s to reference %s with graphmap", fasta_path, args.ref)
+        logger.info("Aligning %s to reference %s with graphmap", fasta_path, args.ref)
         align_utils.align_with_graphmap(fasta_path, args.ref, args.circular, sam_path)
 
     filtered_sam = os.path.join(args.out_folder, name + "_filtered.sam")
     if os.path.isfile(filtered_sam):
-        logging.info("%s exists, skipping", filtered_sam)
+        logger.info("%s exists, skipping", filtered_sam)
     else:
         filters = [read_len_filter(min_len=args.min_length)]
         n_kept, n_discarded = filter_aligments_in_sam(sam_path, filtered_sam, filters)
-        logging.info("Outputed filtered sam to %s\n%d kept, %d discarded",
+        logger.info("Outputed filtered sam to %s\n%d kept, %d discarded",
                  filtered_sam, n_kept, n_discarded)
 
     reads_csv = os.path.join(args.out_folder, name + "_read_data.csv")
     if os.path.isfile(reads_csv):
-        logging.info("%s file exists, loading", reads_csv)
[…]

### 2017-05-07 paper/main.tex
Update

@@
+%!TeX spellcheck = en-US
 \documentclass[times, utf8, seminar, numeric]{fer}
@@
 \begin{sazetak}
+MinION je citac DNA najnovije generacije.
+Oxford Nanopore Technologies ga je u svrhu prvog portabilnog DNA sekvencora.
+Proizvodi vrlo duga ocitanja (duljine i preko 100 kBp), doduse pod cijelu preciznosti.
+U ovom radu prezentiramo altrenativnu metodu ocitanja DNA koristenjem convolucijskim neuronskih mreza (CNN) umjesto dosadasnjeg pristupa s HMM ili RNN.
+Takoder cijeli postupak je istreniran od izvornog signala do ocitanih nukleoida, bez ovisnosti o umjetnoj segregaciji na dogadaje.
 
-\kljucnerijeci{Basecaller, MinION, sekvenciniranje sljedece generacije}
+\kljucnerijeci{Basecaller, MinION, R9, CNN, CTC, sekvenciniranje sljedece generacije}
 \end{sazetak}
@@
 MinION is able to produce very long reads (reads over 100 kBp were reported), however it suffers from high sequencing error rate.
-In this paper, we show novel model reducing the error rate can be reduced by improving the base calling process on 1D reads using gated residual convolutional neural networks (CNN) combined with Connectionist Temporal Classification (CTC) learning approach.
+In this paper, we show novel model basecaller which basecalles 1D reads using gated residual convolutional neural networks (CNN) combined with Connectionist Temporal Classification (CTC) learning approach.
+Its performance is on par with state of the art and offers possible future improvement.
 
@@
 A MinION device typically yields reads several thousand bases long; reads as long as 100,000 bp have been reported.
-1D read have usually error rate about   30\%.
-After parameter training, base calling can be performed by running the Viterbi algorithm, which will result in the sequence of states with the highest likelihood.
+1D read have usually error rate about   10\%.
+(Check details!!, What is the error rate, how it's defined???) After parameter training, base calling can be performed by running the Viterbi algorithm, which will result in the sequence of states with the highest likelihood.
 
@@
 Earlier models were (Hidden markov model) HMM based where hidden state modeled DNA sequence of length 6 (6-mer) in the nanopore.
-Pore models were used in computing emission probabilities.~\cite{loman2015complete,schreiber2015analysis,szalay2015novo,timp2012dna} and the recest open source HMM based basecaller Nanocall~\cite{david2016nanocall}.
+Pore models were used in computing emission probabilities.~\cite{loman2015complete,schreiber2015analysis,szalay2015novo,timp2012dna} and the recent open source HMM based basecaller Nanocall~\cite{david2016nanocall}.
 
@@
 
-Dataset were obtained from (Ask Marko!).
+Dataset were obtained from (Ask somebody, Marko doesn't know!).
 In this research models were trainined on ecoli K-12 strand.
@@
 Using this data, we get the intermediate target sequence for each block.
-To correc for model errors, we use aligned information, that is reference genome and alignment data(cigar string) to correct those information, yielding finalized target sequence.
+To correct for model errors, we use aligned information, that is reference genome and alignment data(cigar string) to correct those information, yielding finalized target sequence.
 For further performace we skip first and last traning block, since most error aggregate on edges.
@@
 Adjacent matching bases are spearated with surogate one, for example: AAA -> AA'A for reasong described in following section.
-
 \section{Residual arhitecture}
@@
 
-TODO.
+TODO
+
+See \url{https://github.com/nmiculinic/minion-basecaller/blob/master/notebook/Lambda_nanonet_vs_m270.ipynb for preliminary results.}
 
@@
 
-Unlike Nanonet which uses custom OpenCL kernels, or Albacore -- a novel ONT basecaller as of May 2017 lacking GPU support, this work used world class computational framework tensorflow with highly optimized kernels and large development community.
+Unlike Nanonet which uses custom OpenCL kernels, or Albacore --- a novel ONT basecaller as of May 2017 lacking GPU support, this work used world class computational framework tensorflow with highly optimized kernels and large development community.
 Therefore resulting paper's effect is showcasing Residual CNN approach or pure CNN approach with CTC loss is marginally better than already established basecaller, and providing code in contemporary framework.
@@
 \bibliography{refs}
-% \bibliographystyle{abbrv}
 

### 2017-05-07 paper/main.tex
Spelling checked.

@@
-%!TeX spellcheck = en-US
 \documentclass[times, utf8, seminar, numeric]{fer}
@@
 
-\voditelj{Mile Sikić}
+\voditelj{izv.~prof.~dr.~sc.~Mile Sikić}
 \title{MinCall --- MinION od raw signala do genoma}
@@
     The MinION device by Oxford Nanopore is the first portable sequencing device.
-MinION is able to produce very long reads (reads over 100 kBp were reported), however it suffers from high sequencing error rate.
+MinION is able to produce very long reads (reads over 100 kBp were reported), however, it suffers from high sequencing error rate.
 In this paper, we show novel model basecaller which basecalles 1D reads using gated residual convolutional neural networks (CNN) combined with Connectionist Temporal Classification (CTC) learning approach.
@@
 
-    Whole process is trainned end-to-end from raw signal on R9 sequecing pore and once trainned it's independent from Metrichor.
+    The whole process is trained end-to-end from the raw signal on R9 sequencing pore and once trained it's independent of Metrichor.
 \keywords{Basecaller, MinION, R9, CNN, CTC, Next generation sequecing}
@@
 Although MinION is able to produce long reads, they have a high sequencing error rate.
-This has been somewhat aleviated with new R9 pore model, replacing older R7 ones.
+This has been somewhat alleviated with new R9 pore model, replacing older R7 ones.
 In this paper, we show that this error rate can be reduced by replacing the default base caller provided by the manufacturer with a properly trained neural network model.
@@
 The electric current depends mostly on the context of several DNA bases passing through the pore at the time of measurement.
-As the DNA moves through the pore, the context shifts and the electric current changes.
+As the DNA moves through the pore, the context shifts, and the electric current changes.
 
 A MinION device typically yields reads several thousand bases long; reads as long as 100,000 bp have been reported.
-1D read have usually error rate about   10\%.
+1D read has usually error rate about   10\%.
 (Check details!!, What is the error rate, how it's defined???) After parameter training, base calling can be performed by running the Viterbi algorithm, which will result in the sequence of states with the highest likelihood.
@@
 The resulting DNA fragments pass through a protein embedded in a membrane via a nanometre-sized channel (this protein is the `nanopore').
-A single strand of DNA passes through the pore; the optional use of a hairpin adapter at one end of the fragment allows the two strands of DNA to serially pass through nanopore, allowing two measurements of the fragment.
-With hairpin it's called 2D read, which isn't topic of this paper.
+A single strand of DNA passes through the pore; the optional use of a hairpin adapter at one end of the fragment allows the two strands of DNA to serially pass through the nanopore, allowing two measurements of the fragment.
+With hairpin, it's called 2D read, which isn't the topic of this paper.
 In here we focus only on 1D reads.
@@
 The central idea is that the single-stranded DNA product present in the nanopore affects the current in a way that is strong enough to enable decoding the electric signal data into a DNA sequence.
-This process, called basecalling, takes as input a list of current measurements, and produces as output a list of DNA bases most likely to have generated those currents.
+This process, called basecalling, takes as input a list of current measurements and produces as output a list of DNA bases most likely to have generated those currents.
 
-The nanopore is 6 nucleoides wide, and eariler HMM based models model it as such.
+The nanopore is 6 nucleotides wide, and eariler HMM-based models model it as such.
 
@@
 
-The core of the decoding process is the basecalling step. Official basecaller is Metrichor, previously perfomed in the cloud before being opensourced under name Nanonet.
+The core of the decoding process is the basecalling step. Official basecaller is Metrichor, previously performed in the cloud before being open-sourced under name Nanonet.
 
-Earlier models were (Hidden markov model) HMM based where hidden state modeled DNA sequence of length 6 (6-mer) in the nanopore.
-Pore models were used in computing emission probabilities.~\cite{loman2015complete,schreiber2015analysis,szalay2015novo,timp2012dna} and the recent open source HMM based basecaller Nanocall~\cite{david2016nanocall}.
+Earlier models were (Hidden Markov model) HMM-based where hidden state modeled DNA sequence of length 6 (6-mer) in the nanopore.
+Pore models were used in computing emission probabilities.~\cite{loman2015complete,schreiber2015analysis,szalay2015novo,timp2012dna} and the recent open source HMM-based basecaller Nanocall~\cite{david2016nanocall}.
 
@@
 
-Instead of opting for traditional path using HMM or RNN we tried using CNN~(Convolutional neural networks)~\cite{lecun-98}, that is their residual version~\cite{he2016deep}.
-We opted out for gated residuan network variant~\cite{savarese2016learning}.
-For loss we used CTC (Connectionist temporal classification)~\cite{graves2006connectionist} between basecallled and target sequence.
+Instead of opting for the traditional path using HMM or RNN we tried using CNN~(Convolutional neural networks)~\cite{lecun-98}, that is their residual version~\cite{he2016deep}.
+We opted out for gated residual network variant~\cite{savarese2016learning}.
+For loss we used CTC (Connectionist temporal classification)~\cite{graves2006connectionist} between basecallled and the target sequence.
 The implementation used is open source warp-ctc~\cite{warpctc}.
@@
 
-Dataset were obtained from (Ask somebody, Marko doesn't know!).
-In this research models were trainined on ecoli K-12 strand.
+Dataset was obtained from (Ask somebody, Marko doesn't know!).
+In this research, models were trained on the ecoli K-12 strand.
 (Should I cite someone for strand).
@@
 
-The fast5 input training files were further split into smaller training blocks, constiting of fixed block size on raw signal. For each block target sequence basecalled data is used in following way.
+The fast5 input training files were further split into smaller training blocks, consisting of fixed block size on raw signal. For each block target sequence, basecalled data is used in following way.
 
@@
 Using this data, we get the intermediate target sequence for each block.
-To correct for model errors, we use aligned information, that is reference genome and alignment data(cigar string) to correct those information, yielding finalized target sequence.
-For further performace we skip first and last traning block, since most error aggregate on edges.
+To correct for model errors, we use aligned information, that is reference genome and alignment data(cigar string) to correct that information, yielding finalized target sequence.
+For further performance we skip first and last training block, since most error aggregate on edges.
 
-Adjacent matching bases are spearated with surogate one, for example: AAA -> AA'A for reasong described in following section.
+Adjacent matching bases are separated with surrogate one, for example AAA -> AA'A for reasons described in the following section.
 \section{Residual arhitecture}
@@
 We use gated residual network, that is layer function is like this: $f(x) = k g(x) + x$ Where $x$ is input layer and $k$ constant learnt during training.
-Specifically $g$ used is multiple, 1--3 times in our experiments, Relu-BatchNorm-CNN layers.
+Specifically, $g$ used is multiple, 1--3 times in our experiments, Relu-BatchNorm-CNN layers.
 
@@
 
-This section gives brief overview over CTC, for further detail and in depth explaintion we recommend original paper~\cite{graves2006connectionist} or searhcing for contemporary blog posts.
+This section gives a brief overview over CTC, for further detail and in depth explanation we recommend original paper~\cite{graves2006connectionist} or searching for contemporary blog posts.
 
@@
 
-For decoding we use beam search decoder, with beam width 100.
+For decoding, we use beam search decoder, with beam width 100.
 
@@
 
-This model used advance state-of-the-art gated residual convolutional neural network, with top models having 270 layers and over 3M parameters, yet improvements over Metrichorn baseline are mariginal.
-As conclusion it might be that we've reached Bayesian error rate for R9 chemistry.
-Furthermore R9.5 and 1D\^2 reads are under development which shall yield this paper's result obsolite quite soon, yet underlying code developed could easily be adjusted, and trained on new data.
+This model used advance state-of-the-art gated residual convolutional neural network, with top models having 270 layers and over 3M parameters, yet improvements over Metrichorn baseline are marginal.
+As the conclusion, it might be that we've reached Bayesian error rate for R9 chemistry.
+Furthermore, R9.5 and 1D\^2 reads are under development which shall yield this paper's result obsolete quite soon, yet underlying code developed could easily be adjusted and trained on new data.
 
-Unlike Nanonet which uses custom OpenCL kernels, or Albacore --- a novel ONT basecaller as of May 2017 lacking GPU support, this work used world class computational framework tensorflow with highly optimized kernels and large development community.
-Therefore resulting paper's effect is showcasing Residual CNN approach or pure CNN approach with CTC loss is marginally better than already established basecaller, and providing code in contemporary framework.
+Unlike Nanonet which uses custom OpenCL kernels or Albacore --- a novel ONT basecaller as of May 2017 lacking GPU support, this work used world-class computational framework tensorflow with highly optimized kernels and large development community.
+Therefore resulting paper's effect is showcasing Residual CNN approach or pure CNN approach with CTC loss is marginally better than already established basecaller and providing code in the contemporary framework.
 
@@
 \end{document}
+end{document}

### 2017-05-07 mincall/tools/analyse_bc.py
Operational settings for analysis_bc

@@ -24,9 +24,9 @@ args.add_argument("--coverage_threshold", help="Minimal coverage threshold for c
 args = args.parse_args()
 
 basecallers = {
-    # "mincall_m270": ["nvidia-docker", "run", "--rm", "-v", "%s:/data" % args.input_folder, "-u=%d" % os.getuid(), "nmiculinic/mincall:9947283"],
-    # "nanonet": ["nanonetcall", args.input_folder, "--chemistry", "r9", "--platforms", "nvidia:0:20", "--exc_opencl"]
-    "nanonet": ["nanonetcall", args.input_folder, "--chemistry", "r9", "--jobs", "8"]
+    "mincall_m270": ["nvidia-docker", "run", "--rm", "-v", "%s:/data" % args.input_folder, "-u=%d" % os.getuid(), "nmiculinic/mincall:9947283"],
+    "nanonet": ["nanonetcall", args.input_folder, "--chemistry", "r9", "--platforms", "nvidia:0:20", "--exc_opencl"]
+    # "nanonet": ["nanonetcall", args.input_folder, "--chemistry", "r9", "--jobs", "8"]
 }
 
 os.makedirs(args.out_folder, exist_ok=True)

### 2017-05-07 mincall/consensus.py
Migrated input readers to new api (train test dirs), consensus with percentages

@@ -89,13 +89,19 @@ def process_mpileup(alignments_path, reference_path, mpileup_path, coverage_thre
         # transfrorm coverage sum to average coverage
         counts[-1] /= (i+1)
 
-        columns = ['field', 'value']
         fields = ['alignments_file', 'mpileup_file', 'coverage_threshold', 'snp_count', 'insertion_count',
                   'deletion_count', 'num_undercovered_bases', 'num_called_bases',
                   'num_correct_bases', 'average_coverage']
         values = [alignments_path, mpileup_path, coverage_threshold] + counts.tolist()
-        data = list(zip(fields, values))
-        report = pd.DataFrame(data, columns=columns)
+        report = pd.DataFrame([values], columns=fields)
+
+        for col in filter(lambda c: c.endswith('_count'), report.columns):
+            new_col = col.replace('count', 'rate')
+            report[new_col] = 100 * report[col] / report.num_called_bases
+
+        report['correct_rate'] = 100 * report.num_correct_bases / report.num_called_bases
+        report = report.transpose()
+
         if output_prefix:
             summary_file = os.path.join(output_prefix, 'cov_%d.sum.vcf' % coverage_threshold)
             report.to_csv(summary_file, sep=';', index=False)

### 2017-05-07 mincall/input_readers.py

@@ -3,6 +3,7 @@ import socket
 from random import shuffle
 import numpy as np
 import h5py
+from glob import glob
 from collections import defaultdict
 
 from . import util
@@ -11,10 +12,23 @@ from .errors import MissingRNN1DBasecall, InsufficientDataBlocks, MinIONBasecall
 
 root_dir_map = {
     'karla': '/hgst8TB/fjurisic/ecoli',
-    'protagonist': '/home/lpp/Downloads/minion'
+    'protagonist': '/home/lpp/Downloads/minion',
+    'inspiron5520': '/data'
 }
 root_dir_default = root_dir_map.get(socket.gethostname(), '/data')
 
+
+def find_ref(fast5_path):
+    dirname = os.path.dirname(fast5_path)
+    basename = os.path.basename(fast5_path)
+    name, ext = os.path.splitext(basename)
+    ref_path = os.path.join(dirname, name + '.ref')
+
+    if not os.path.exists(ref_path):
+        raise errors.RefFileNotFound(
+            "{}-> search query:{}.ref in following directory {}".format(basename, name, dirname))
+
+    return ref_path
 
 class InputReader():
     def input_fn(self):
@@ -90,6 +104,7 @@ class AlignedRawAbstract(InputReader):
             else:
                 raise ValueError("extension not recognized %s" % ref_ext)
 
+            ref_seq = ref_seq.upper()
             corrected_basecalled = util.correct_basecalled(bucketed_basecall, ref_seq, nedit_tol=self.nedit_tol)
 
             if verify_file:
@@ -105,20 +120,26 @@ class AlignedRawAbstract(InputReader):
             y, y_len = util.prepare_y(corrected_basecalled[1:1 + num_blocks], block_size_y)
             return x, x_len, y, y_len
 
-    def find_ref(self, fast5, dirs):
-        fname = os.path.splitext(fast5)[0].split(os.sep)[-1]
-        for d in dirs:
-            pick = os.path.join(d, fname + ".ref")
-            if os.path.exists(pick):
-                return pick
+    def find_ref(self, fast5_path):
+        dirname = os.path.dirname(fast5_path)
+        basename = os.path.basename(fast5_path)
+        name, ext = os.path.splitext(basename)
+        ref_path = os.path.join(dirname, name + '.ref')
+        if not os.path.exists(ref_path):
+            raise errors.RefFileNotFound(
+                "{}-> search query:{}.ref in following directory {}".format(basename, name, dirname))
 
[…]

### 2017-05-07 mincall/models/residual_deep.py

@@ -56,6 +56,7 @@ def model_fn(net, X_len, max_reach, block_size, out_classes, batch_size, dtype, 
 def model_setup_params(hyper):
     return dict(
         g=tf.Graph(),
+        per_process_gpu_memory_fraction=0.6,
         block_size_x=8 * 3 * 600 // 2,
         block_size_y=630,
         in_data=input_readers.HMMAlignedRaw(),
@@ -75,15 +76,11 @@ def model_setup_params(hyper):
 
 
 sigopt_params = [
-    sigopt_double('initial_lr', 1e-5, 1e-3),
-    sigopt_double('decay_factor', 1e-3, 0.5),
-    sigopt_int('num_layers', 10, 20),
-    sigopt_int('num_sub_layers', 1, 2),
 ]
 
 default_params = {
-    'initial_lr': 0.000965352400196344,
-    'decay_factor': 0.0017387361908150767,
+    'initial_lr': 0.0009,
+    'decay_factor': 0.002,
     'num_layers': 20,
     'num_sub_layers': 2
 }
@@ -99,7 +96,7 @@ def create_test_model(hyper, **kwargs):
     return create_train_model(hyper, **kwargs)
 
 
-default_name = "resdeep"
+default_name = "resdeep_test"
 
 
 if __name__ == "__main__":

### 2017-05-08 mincall/models/cnn.py
Simple cnn model

@@ -1,0 +1,92 @@
+import tensorflow as tf
+from tflearn.layers.conv import max_pool_1d, conv_1d
+from tflearn.layers.normalization import batch_normalization
+from dotenv import load_dotenv, find_dotenv
+from tflearn.initializations import variance_scaling_initializer
+
+from mincall.ops import central_cut
+from mincall import input_readers
+from mincall.util import sigopt_int, sigopt_double
+from mincall.controller import control
+from mincall.model_utils import Model
+load_dotenv(find_dotenv())
+
+
+def model_fn(net, X_len, max_reach, block_size, out_classes, batch_size, dtype, **kwargs):
+    """
+        Args:
+        net -> Input tensor shaped (batch_size, max_reach + block_size + max_reach, 3)
+        Returns:
+        logits -> Unscaled logits tensor in time_major form, (block_size, batch_size, out_classes)
+    """
+
+    for block in range(1, 4):
+        with tf.variable_scope("block%d" % block):
+            for layer in range(kwargs['num_layers']):
+                with tf.variable_scope('layer_%d' % layer):
+                    net = batch_normalization(net)
+                    net = tf.nn.relu(net)
+                    net = conv_1d(net, 64, 3,
+                                  scope="conv_1d_%d" % layer,
+                                  weights_init=variance_scaling_initializer(dtype=dtype))
+
+            net = max_pool_1d(net, 2)
+        net = tf.nn.relu(net)
+
+    net = central_cut(net, block_size, 8)
+    net = tf.transpose(net, [1, 0, 2], name="Shift_to_time_major")
+    net = conv_1d(net, 9, 1, scope='logits')
+    return {
+        'logits': net,
+        'init_state': tf.constant(0),
+        'final_state': tf.constant(0),
+    }
+
+
+def model_setup_params(hyper):
+    return dict(
+        g=tf.Graph(),
+        per_process_gpu_memory_fraction=0.65,
+        block_size_x=8 * 3 * 600 // 2,
+        block_size_y=630,
+        in_data=input_readers.HMMAlignedRaw(),
+        num_blocks=3,
+        batch_size=16,
+        max_reach=8 * 20,  # 240
+        queue_cap=300,
+        overwrite=False,
+        reuse=False,
+        shrink_factor=8,
+        dtype=tf.float32,
+        model_fn=model_fn,
[…]

### 2017-05-08 mincall/bioinf_utils.py
Input reader minor randomisation

@@ -79,7 +79,7 @@ def decompress_cigar_pairs(cigar_pairs, mode='ints'):
     elif mode == 'chars':
         convert = lambda x: x
     else:
-        raise Exception('Invalid mode argument. Expected ints or chars')
+        raise ValueError('Invalid mode argument. Expected ints or chars')
 
     extended = (cnt * [convert(b)] for b, cnt in cigar_pairs)
     return ''.join(itertools.chain.from_iterable(extended))
@@ -187,8 +187,8 @@ def error_rates_from_cigar(cigar_full_str):
     read_len = n_insertions + n_missmatches + n_matches
     noncliped_len = n_all_insertions + n_missmatches + n_matches
     if cigar_len != n_deletions + n_all_insertions + n_missmatches + n_matches:
-        raise Exception("cigar_len != n_deletions + n_all_insertions + n_missmatches + n_matches "
-                        "- Expected extended cigar format")
+        raise ValueError("cigar_len != n_deletions + n_all_insertions + n_missmatches + n_matches"
+                         ";Expected extended cigar format")
 
     n_errors = n_missmatches + n_insertions + n_deletions
 

### 2017-05-08 mincall/input_readers.py

@@ -1,6 +1,6 @@
 import os
 import socket
-from random import shuffle
+from random import shuffle, randint
 import numpy as np
 import h5py
 from glob import glob
@@ -93,8 +93,11 @@ class AlignedRawAbstract(InputReader):
                     bucketed_basecall[target_bucket] += \
                         b['model_state'][-b['move']:].decode("ASCII")
 
-            if len(bucketed_basecall) < num_blocks + 2:
-                raise InsufficientDataBlocks("Has only {} blocks, while requesting {} + first and last".format(len(bucketed_basecall), num_blocks))
+            # -2 for first and last that are always skipped
+            num_blocks_max = len(bucketed_basecall) - 2
+            if num_blocks_max < num_blocks:
+                raise InsufficientDataBlocks("Has only {} blocks, while requesting {} + first and last"
+                                             .format(len(bucketed_basecall), num_blocks))
 
             ref_ext = os.path.splitext(ref_path)[1]
             if ref_ext == ".ref":
@@ -112,12 +115,15 @@ class AlignedRawAbstract(InputReader):
                 np.testing.assert_string_equal("".join(bucketed_basecall), fastq[1])
                 np.testing.assert_string_equal("".join(corrected_basecalled), ref_seq)
 
+            # Skipping first n random blocks
+            skip_first_n = randint(1, num_blocks_max-num_blocks+1)
+
             x = np.zeros([block_size_x * num_blocks, 1], dtype=np.float32)
             x_len = min(len(signal), block_size_x * num_blocks)
 
-            # Skipping first block
-            x[:x_len, 0] = signal[block_size_x:block_size_x + x_len]
-            y, y_len = util.prepare_y(corrected_basecalled[1:1 + num_blocks], block_size_y)
+            x_offset = skip_first_n*block_size_x
+            x[:x_len, 0] = signal[x_offset:x_offset + x_len]
+            y, y_len = util.prepare_y(corrected_basecalled[skip_first_n:skip_first_n + num_blocks], block_size_y)
             return x, x_len, y, y_len
 
     def find_ref(self, fast5_path):
@@ -160,6 +166,7 @@ class AlignedRawAbstract(InputReader):
                         num_blocks=model.num_blocks,
                         verify_file=False
                     )
+                    # sol = (x, x_len, y, y_len)
                     np.testing.assert_array_less(0, sol[3], err_msg='y_len must be > 0')
 
                     yield {

### 2017-05-08 mincall/models/cnn.py

@@ -50,7 +50,7 @@ def model_setup_params(hyper):
         block_size_x=8 * 3 * 600 // 2,
         block_size_y=630,
         in_data=input_readers.HMMAlignedRaw(),
-        num_blocks=3,
+        num_blocks=4,
         batch_size=16,
         max_reach=8 * 20,  # 240
         queue_cap=300,
@@ -71,7 +71,7 @@ sigopt_params = [
 default_params = {
     'initial_lr': 0.0009,
     'decay_factor': 0.002,
-    'num_layers': 20,
+    'num_layers': 5,
 }
 
 
@@ -85,7 +85,7 @@ def create_test_model(hyper, **kwargs):
     return create_train_model(hyper, **kwargs)
 
 
-default_name = "cnn_20l_64c_3"
+default_name = "cnn_5l_64c_3"
 
 
 if __name__ == "__main__":

### 2017-05-08 mincall/input_readers.py
Input reader minor change

@@ -116,7 +116,8 @@ class AlignedRawAbstract(InputReader):
                 np.testing.assert_string_equal("".join(corrected_basecalled), ref_seq)
 
             # Skipping first n random blocks
-            skip_first_n = randint(1, num_blocks_max-num_blocks+1)
+            #skip_first_n = randint(1, num_blocks_max-num_blocks+1)
+            skip_first_n = 1
 
             x = np.zeros([block_size_x * num_blocks, 1], dtype=np.float32)
             x_len = min(len(signal), block_size_x * num_blocks)

### 2017-05-08 mincall/models/residual_deep.py
per_process_gpu_memory_fraction example usage in test model

@@ -76,11 +76,15 @@ def model_setup_params(hyper):
 
 
 sigopt_params = [
+    sigopt_double('initial_lr', 1e-5, 1e-3),
+    sigopt_double('decay_factor', 1e-3, 0.5),
+    sigopt_int('num_layers', 10, 20),
+    sigopt_int('num_sub_layers', 1, 2),
 ]
 
 default_params = {
-    'initial_lr': 0.0009,
-    'decay_factor': 0.002,
+    'initial_lr': 0.000965352400196344,
+    'decay_factor': 0.0017387361908150767,
     'num_layers': 20,
     'num_sub_layers': 2
 }
@@ -96,7 +100,7 @@ def create_test_model(hyper, **kwargs):
     return create_train_model(hyper, **kwargs)
 
 
-default_name = "resdeep_test"
+default_name = "resdeep"
 
 
 if __name__ == "__main__":

### 2017-05-08 mincall/tools/analyse_bc.py

@@ -38,7 +38,7 @@ for name, cmd in basecallers.items():
         logger.info("%s exists, skipping", fasta_path)
     else:
         logger.info("Basecalling %s with %s", args.input_folder, name)
-        logger.info("Output file", fasta_path)
+        logger.info("Output file %s", fasta_path)
         logger.info("Full command: %s", " ".join(cmd))
         with open(fasta_path, 'w') as f:
             subprocess.check_call(cmd, stdout=f)

### 2017-05-08 mincall/align_utils.py
Count of mapped reads in sam

@@ -123,6 +123,12 @@ def secondary_aligments_filter():
     return _filter
 
 
+def supplementary_aligments_filter():
+    def _filter(aligment):
+        return not aligment.is_supplementary
+    return _filter
+
+
 def only_mapped_filter():
     def _filter(x):
         return not x.is_unmapped
@@ -169,3 +175,6 @@ def cnt_reads_in_sam(sam_path):
     with pysam.AlignmentFile(sam_path, "r") as samfile:
         total = sum(1 for _ in samfile.fetch())
     return total
+
+def extend_cigar_in_sam(sam_path_in, sam_path_out, reference, reads_fastx):
+    pass

### 2017-05-11 paper/main.tex
Update
Updated abstract

@@
 \begin{sazetak}
-MinION je citac DNA najnovije generacije.
-Oxford Nanopore Technologies ga je u svrhu prvog portabilnog DNA sekvencora.
-Proizvodi vrlo duga ocitanja (duljine i preko 100 kBp), doduse pod cijelu preciznosti.
-U ovom radu prezentiramo altrenativnu metodu ocitanja DNA koristenjem convolucijskim neuronskih mreza (CNN) umjesto dosadasnjeg pristupa s HMM ili RNN.
-Takoder cijeli postupak je istreniran od izvornog signala do ocitanih nukleoida, bez ovisnosti o umjetnoj segregaciji na dogadaje.
+    U ovom znanstvenom clanku istenirali smo model dubokog ucenja za Oxford Nanopore Tehnology's MinION portabilni DNK sekvencor.
+MinION potencijal je ogroman, i zbog njegove male velicina i portabilnosti, kao i dugih ocitanja senzora (ocitanja > 100kbP su primjecena).
+Nazalost pati od vecih gresaka u usporedbi s drugim modelima.
+
+    Kreirali smo nezavisni model pocevsi od izvornih podataka i koristili konvolucijske neuronske mreze (CNN) s CTC gubitkom te postigli mjerljive perfomanse.
 
@@
 \begin{abstract}
-    The MinION device by Oxford Nanopore is the first portable sequencing device.
-MinION is able to produce very long reads (reads over 100 kBp were reported), however, it suffers from high sequencing error rate.
-In this paper, we show novel model basecaller which basecalles 1D reads using gated residual convolutional neural networks (CNN) combined with Connectionist Temporal Classification (CTC) learning approach.
-Its performance is on par with state of the art and offers possible future improvement.
+    In this study we create end2end deep learning model for Oxford Nanopore Technology's MinION portable DNA sequencing device.
+The device's potential is enormous, both due to it's small size and portability and long read (over 100kBp reported).
+However is suffers with higher error rate compared to other methods.
 
-    The whole process is trained end-to-end from the raw signal on R9 sequencing pore and once trained it's independent of Metrichor.
+    We've created independant model starting with raw data and used Convolutional Neural Network (CNN) combined with Connectionist Temporal Classification (CTC) loss for measurable performance.
 \keywords{Basecaller, MinION, R9, CNN, CTC, Next generation sequecing}
@@
 (Check details!!, What is the error rate, how it's defined???) After parameter training, base calling can be performed by running the Viterbi algorithm, which will result in the sequence of states with the highest likelihood.
+
+The exact error rate metric is unreliable since multiple pipline tools could be the issue.
+First the sample is prepared, hopefully uncontaminated and matching reference genome as close as possile, then sequenced using the MinION device obtaining raw data.
+Next our model (or other groups ones) come along, basecall the sequence.
+To evalute error rate metric basecalled read is aligned to reference using aligners with their own erros/biases, mostly commonly used BWA-MEM and Graphmap.
 
@@
 
-In out concrete $x$ aplication $\Sigma = \{\text{A G T C A' G' T' A'}\}$
+In out concrete $x$ application $\Sigma = \{\text{A G T C A' G' T' A'}\}$
 
@@
 
+\chapter{Acknowledgments}
+
+TODO.
+
 \bibliographystyle{fer}

### 2017-05-12 mincall/tools/analyse_bc.py
Added Blacnk albacore

@@ -25,7 +25,8 @@ args = args.parse_args()
 
 basecallers = {
     "mincall_m270": ["nvidia-docker", "run", "--rm", "-v", "%s:/data" % args.input_folder, "-u=%d" % os.getuid(), "nmiculinic/mincall:9947283"],
-    "nanonet": ["nanonetcall", args.input_folder, "--chemistry", "r9", "--platforms", "nvidia:0:20", "--exc_opencl"]
+    "nanonet": ["nanonetcall", args.input_folder, "--chemistry", "r9", "--platforms", "nvidia:0:20", "--exc_opencl"],
+    "albacore": None
     # "nanonet": ["nanonetcall", args.input_folder, "--chemistry", "r9", "--jobs", "8"]
 }
 

### 2017-05-12 mincall/tools/analyse_bc.py
Fixed .

@@ -87,17 +87,20 @@ for name, cmd in basecallers.items():
     logger.info("%s consensus_report:\n%s", name, consensus_report)
 
 
-fig_kde_path = os.path.join(args.out_folder, name + "reads_kde.png")
-fig_hist_path = os.path.join(args.out_folder, name + "reads_hist.png")
+fig_kde_path = os.path.join(args.out_folder, "reads_kde.png")
+fig_hist_path = os.path.join(args.out_folder, "reads_hist.png")
 
 fig_kde, axes_kde = plt.subplots(3, 2)
 fig_hist, axes_hist = plt.subplots(3, 2)
 fig_kde.set_size_inches(12, 20)
 fig_hist.set_size_inches(12, 20)
 for col, ax_kde, ax_hist in zip(next(iter(dfs.values()))._get_numeric_data(), axes_kde.ravel(), axes_hist.ravel()):
+    fig, ax = plt.subplots()
     for k in dfs.keys():
+        sns.kdeplot(dfs[k][col], shade=True, label=k, alpha=0.5, ax=ax)
         sns.kdeplot(dfs[k][col], shade=True, label=k, alpha=0.5, ax=ax_kde)
         ax_hist.hist(dfs[k][col], label=k, alpha=0.5)
+    fig.savefig(os.path.join(args.out_folder, col + ".png"))
     for ax in [ax_kde, ax_hist]:
         ax.legend()
         ax.set_title(col)

### 2017-05-12 paper/main.tex
Implemented feedback

@@
 \tableofcontents
+% - razvio si novi base caller za 1D nanopore reads koji je temeljen na konvolucijskim neuronskim mrežama
+% - reci koliko slojeva ima model
+% - kakve si rezultate dobio te da li su bolji od postojećih.
+% - koju si razinu greške dobio
+% - daj neko ime base calleru
 
 \begin{sazetak}
-    U ovom znanstvenom clanku istenirali smo model dubokog ucenja za Oxford Nanopore Tehnology's MinION portabilni DNK sekvencor.
-MinION potencijal je ogroman, i zbog njegove male velicina i portabilnosti, kao i dugih ocitanja senzora (ocitanja > 100kbP su primjecena).
-Nazalost pati od vecih gresaka u usporedbi s drugim modelima.
-
-    Kreirali smo nezavisni model pocevsi od izvornih podataka i koristili konvolucijske neuronske mreze (CNN) s CTC gubitkom te postigli mjerljive perfomanse.
+    U ovoj studiji kreirali smo MinCall, cjelokupni model očitanja nukleotida za MinION, portabilni DNK sekvencer firme Oxford Nanopore Technology.
+Model je baziran na dubokom učenju i koristi konvolucijske neuronske mreže u svojoj implementaciji.
+Najperformantniji 270 slojeva dubok model postiže najbolje dosad performanse od 90.5\% preciznosti na E.Coli skupu za učenje koristeći R9 kemiju i 1D čitanja.
 
@@
 \begin{abstract}
-    In this study we create end2end deep learning model for Oxford Nanopore Technology's MinION portable DNA sequencing device.
-The device's potential is enormous, both due to it's small size and portability and long read (over 100kBp reported).
-However is suffers with higher error rate compared to other methods.
+    In this study, we created MinCall, an end2end basecaller model for the MinION, an  Oxford Nanopore Technology's portable DNA sequencing device.
+The model is based on deep learning and uses convolutional neural networks (CNN) in its implementation.
+The best performing 270 layers deep model achieves state-of-the-art 90.5\% match rate on E.Coli dataset using R9 pore chemistry and 1D reads.
 
-    We've created independant model starting with raw data and used Convolutional Neural Network (CNN) combined with Connectionist Temporal Classification (CTC) loss for measurable performance.
 \keywords{Basecaller, MinION, R9, CNN, CTC, Next generation sequecing}

### 2017-05-13 mincall/align_utils.py
Bugfix in validation in model utils (i not referenced), added n_threads in preprocess

@@ -85,11 +85,14 @@ def _merge_circular_aligment(target_1, start_pos_1, cigar_str_1,
     return [target, start, cigar]
 
 
-def align_with_graphmap(reads_path, ref_path, is_circular, out_sam):
+def align_with_graphmap(reads_path, ref_path, is_circular, out_sam, n_threads=None):
     os.makedirs(os.path.dirname(out_sam), exist_ok=True)
     args = ["graphmap", "align", "-r", ref_path, "-d", reads_path, "-o", out_sam, "-v", "0", "--extcigar"]
     if is_circular:
         args.append("-C")
+
+    if n_threads is not None:
+        args.extend(["-t", str(n_threads)])
 
     exit_status = subprocess.call(args)
     logging.info("Graphmap exit status %d" % exit_status)
@@ -98,13 +101,15 @@ def align_with_graphmap(reads_path, ref_path, is_circular, out_sam):
         logging.warning("Graphmap exit status %d" % exit_status)
 
 
-def align_with_bwa_mem(reads_path, ref_path, is_circular, out_sam, extended_cigar=True):
+def align_with_bwa_mem(reads_path, ref_path, is_circular, out_sam, n_threads=None):
     os.makedirs(os.path.dirname(out_sam), exist_ok=True)
     if os.path.exists(out_sam):
         logging.info("Removing %s", out_sam)
         os.remove(out_sam)
 
-    n_threads = multiprocessing.cpu_count()
+    if n_threads is None:
+        n_threads = multiprocessing.cpu_count()
+
     args = ["bwa", "mem", "-x", "ont2d", "-t", str(n_threads),
             ref_path, reads_path]
     args_index = ["bwa", "index", ref_path]

### 2017-05-13 mincall/model_utils.py

@@ -667,7 +667,7 @@ class Model():
 
         mu_edit, mu_acc = np.mean(nedit), np.mean(acc)
         std_edit, std_acc = np.std(nedit), np.std(acc)
-        se_edit, se_acc = std_edit / np.sqrt(i + 1), std_acc / np.sqrt(i + 1)
+        se_edit, se_acc = std_edit / np.sqrt(len(nedit) + 1), std_acc / np.sqrt(len(nedit) + 1)
 
         self.logger.info("step: %d [samples %d] avg edit %.4f s %.4f CI <%.4f, %.4f> avg_acc %.4f s %.4f CI <%.4f, %.4f> %.3f bps",
                          self.get_global_step(), n, mu_edit, std_edit, mu_edit - 2*se_edit, mu_edit + 2*se_edit,

### 2017-05-13 mincall/preprocess_dataset.py

@@ -153,7 +153,7 @@ def preprocess_all_ref(dataset_conf_path, only_split, align_function):
     train_root = dataset_config['DEFAULT']['train_root']
     test_root = dataset_config['DEFAULT']['test_root']
     per_genome_split_root = dataset_config['DEFAULT']['per_genome_split_root']
-
+    n_threads = int(dataset_config['DEFAULT']['n_threads'])
     for section in dataset_config.sections():
         config = dataset_config[section]
 
@@ -167,7 +167,7 @@ def preprocess_all_ref(dataset_conf_path, only_split, align_function):
             logging.info("Started %s" % section)
 
             def generate_sam_f(reads, sam_out):
-                align_function(reads, ref_path, is_circular, sam_out)
+                align_function(reads, ref_path, is_circular, sam_out, n_threads)
 
             logging.info("Started aligning")
             align_for_reference(fast5_root, ref_root, generate_sam_f, ref_path, batch_size)

### 2017-05-13 paper/main.tex
Update

@@
 
-\voditelj{izv.~prof.~dr.~sc.~Mile Sikić}
-\title{MinCall --- MinION od raw signala do genoma}
-\engtitle{MinCall --- MinION end2end deep learning basecaller}
-\author{Neven Miculinić}
-
-\maketitle
-\tableofcontents
 % - razvio si novi base caller za 1D nanopore reads koji je temeljen na konvolucijskim neuronskim mrežama
@@
 
+\voditelj{izv.~prof.~dr.~sc.~Mile Sikić}
+\title{MinCall --- MinION basecaller model pogonjen dubokim učenjem}
+\engtitle{MinCall --- MinION end2end deep learning basecaller}
+\author{Neven Miculinić}
+
+\maketitle
+\tableofcontents
+
 \begin{sazetak}
-    U ovoj studiji kreirali smo MinCall, cjelokupni model očitanja nukleotida za MinION, portabilni DNK sekvencer firme Oxford Nanopore Technology.
+    MinION uredaj firme Oxford Nanopore Technology je prvi portabilni DNK sekvencer.
+Sposoban je pročitati značajno duga čitanja, preko 100 kBp, no nažalost pati od značajno vise pogreške nego ostale metode.
+
+    U ovoj studiji kreirali smo MinCall, cjelokupni model očitanja nukleotida za MinION.
 Model je baziran na dubokom učenju i koristi konvolucijske neuronske mreže u svojoj implementaciji.
+Za najbolje performanse koristi najnovije tehnike dubokog učenja i arhitekture, specifično rezidualne neuronske mreže, normalizaciju nad grupom i CTC gubitak.
+
     Najperformantniji 270 slojeva dubok model postiže najbolje dosad performanse od 90.5\% preciznosti na E.Coli skupu za učenje koristeći R9 kemiju i 1D čitanja.
-
 \kljucnerijeci{Basecaller, MinION, R9, CNN, CTC, sekvenciniranje sljedece generacije}
@@
 \begin{abstract}
-    In this study, we created MinCall, an end2end basecaller model for the MinION, an  Oxford Nanopore Technology's portable DNA sequencing device.
+    The Oxford Nanopore Technologies's MinION is the first portable DNA sequencing device.
+It's capable of producing long reads, over 100 kBp were reported, however, is has significantly higher error rate than other methods.
+
+    In this study, we created MinCall, an end2end basecaller model for the MinION.
 The model is based on deep learning and uses convolutional neural networks (CNN) in its implementation.
+For extra perfomnaces is uses cutting edge deep learning techniques and architectures, notably gated residual network, batch normalization and Connectionist Temporal Classification (CTC) loss.
+
     The best performing 270 layers deep model achieves state-of-the-art 90.5\% match rate on E.Coli dataset using R9 pore chemistry and 1D reads.
-
 \keywords{Basecaller, MinION, R9, CNN, CTC, Next generation sequecing}

### 2017-05-14 paper/main.tex
Small update

@@
 
-In this paper, we introduce the base caller for the MinION nanopore sequencing platform~\cite{mikheyev2014first}.
-The MinION device by Oxford Nanopore, weighing only 90 grams, is currently the smallest high-throughput DNA sequencer.
-Thanks to its low capital costs, small size and the possibility of analyzing the data in real time as they are produced, MinION is very promising for clinical applications, such as monitoring infectious disease outbreaks~\cite{judge2015early}\cite{quick2016real}, characterizing structural variants in cancer\cite{norris2016nanopore} and even full human genome assembly~\cite{jain2017nanopore}.
+Oxford Nanopore Technology's MinION nanopore sequecing platform~\cite{mikheyev2014first} is the first portable DNA sequecing device.
+It's small weight, of only 90 grms, low capial cost, and long read length combided with high-throughtput, real time data analysis, and decent accuracy yield promissing resultins in various applications.
+From clinical application such as monitoring infectious desease outbreaks~\cite{judge2015early}\cite{quick2016real}, characterizing structural variants in cancer\cite{norris2016nanopore} and even full human genome assembly~\cite{jain2017nanopore}.
 
@@
 In this paper, we show that this error rate can be reduced by replacing the default base caller provided by the manufacturer with a properly trained neural network model.
+In the future new R9.5 chemistry and 1D^2 reads supersede current models.
 
 In the MinION device, single-stranded DNA fragments move through nanopores, which causes drops in the electric current.
-The electric current is measured at each pore several thousand times per second.
+The electric current is measured at each pore several thousand times per second, 4000 times exactly in our dataset.
 The electric current depends mostly on the context of several DNA bases passing through the pore at the time of measurement.
@@
 A MinION device typically yields reads several thousand bases long; reads as long as 100,000 bp have been reported.
-1D read has usually error rate about   10\%.
+1D read has usually error rate about   10\%, though it varies across diferent genomes, human one being lower accuracy than E.Coli one.
 (Check details!!, What is the error rate, how it's defined???) After parameter training, base calling can be performed by running the Viterbi algorithm, which will result in the sequence of states with the highest likelihood.
 
-The exact error rate metric is unreliable since multiple pipline tools could be the issue.
-First the sample is prepared, hopefully uncontaminated and matching reference genome as close as possile, then sequenced using the MinION device obtaining raw data.
+The exact error rate metric is unreliable since multiple pipeline tools could be the issue.
+First the sample is prepared, hopefully uncontaminated and matching reference genome as close as possibe, then sequenced using the MinION device obtaining raw data.
 Next our model (or other groups ones) come along, basecall the sequence.
-To evalute error rate metric basecalled read is aligned to reference using aligners with their own erros/biases, mostly commonly used BWA-MEM and Graphmap.
+To evalute error rate metric basecalled read is aligned to the reference genome using aligners with their own erros/biases, mostly commonly used BWA-MEM~\cite{li2013aligning} and Graphmap~\cite{sovic2016fast}.
 
@@
 \section{Sequencing overview}
-Informally, the MinION sequencer works as follows.
+Conceptually, the MinION sequencer works as follows.
 First, DNA is sheared into fragments of 8–20 kbp and adapters are ligated to either end of the fragments.
-The resulting DNA fragments pass through a protein embedded in a membrane via a nanometre-sized channel (this protein is the `nanopore').
-A single strand of DNA passes through the pore; the optional use of a hairpin adapter at one end of the fragment allows the two strands of DNA to serially pass through the nanopore, allowing two measurements of the fragment.
-With hairpin, it's called 2D read, which isn't the topic of this paper.
-In here we focus only on 1D reads.
+The resulting DNA fragments pass through a protein embedded in a membrane via a nanometre-sized channel, a nanopore.
+A single DNA strand passes through the pore.
+Optionally, hairpin protein adapter can merge two DNA strands, allowing both template and complemente read passing through the nanopore sequentially for more accurate reads.
+This tehnique is refered as 2D reads, while we focus on 1D reads containing only template DNA and no hairpin adapter.
 
-In ONT terminology, the first strand going through the nanopore is the template, and the second is the complement.
-As a DNA strand passes through the pore it partially blocks the flow of electric current through the pore.
-The flow of current is sampled over time which is the observable output of the system.
-The central idea is that the single-stranded DNA product present in the nanopore affects the current in a way that is strong enough to enable decoding the electric signal data into a DNA sequence.
-This process, called basecalling, takes as input a list of current measurements and produces as output a list of DNA bases most likely to have generated those currents.
-
-The nanopore is 6 nucleotides wide, and eariler HMM-based models model it as such.
+Electrical current runs through the nanopore, and exact nucelotides context within influences the nanopore's resistance.
+This resistive effect is our sensor data, that is the current fluctuations as DNA passes though the pore.
+The nanopore is 6 nucleotides wide, and many models use this information in their approaches, while we're opted out of this technicality.
 
@@
 
-The core of the decoding process is the basecalling step. Official basecaller is Metrichor, previously performed in the cloud before being open-sourced under name Nanonet.
+The core of the decoding process is the basecalling step. Official basecaller is Metrichor, previously performed in the cloud before being open-sourced under name Nanonet. (Citation, whom?)
 
@@
 
-Recent models opted using (Recurrent neural network) RNN, notably DeepNano~\cite{deepnano} and recently open sourced official Nanonets from ONT.
+Recent models opted using (Recurrent neural network) RNN, notably DeepNano~\cite{deepnano} and recently open sourced official Nanonet from ONT.
 
@@
 Using this data, we get the intermediate target sequence for each block.
-To correct for model errors, we use aligned information, that is reference genome and alignment data(cigar string) to correct that information, yielding finalized target sequence.
+To correct for model errors, we use aligned information, that is reference genome and alignment data (cigar string) to correct that information, yielding finalized target sequence.
 For further performance we skip first and last training block, since most error aggregate on edges.
@@
 Adjacent matching bases are separated with surrogate one, for example AAA -> AA'A for reasons described in the following section.
+
 \section{Residual arhitecture}
@@
 
+\subsection{Final model}
+
+Best performing model used has 270 total layers, divided into 3 90-layer blocks.
+Between each 90 layers blocks MaxPool layer is inserted with receptive width of 2 and stride 2, to ease computation effort and add precision.
+
+Each block consists of 30 gated residual layers, each residual layer composed of 3 sequential Relu-Batch Normalization-Conv1D layers.
+Each convolutional layer uses receptive width of 3 with 64 chanells as output throughout the model.
+
+\subsection{E.Coli perfomnace}
+
 TODO

### 2017-05-14 paper/main.tex
Update

@@
 In this paper, we show that this error rate can be reduced by replacing the default base caller provided by the manufacturer with a properly trained neural network model.
-In the future new R9.5 chemistry and 1D^2 reads supersede current models.
+In the future new R9.5 chemistry and 1D\^2 reads supersede current models.
 
@@
 
-\subsection{E.Coli perfomnace}
+\subsection{Performance tables}
 
-TODO
+As decribed in previous section, model was tested on E.Coli test set and compared to open source Nanonet and Albacore.
+Albacore doesn't have specific R9 chemistry mode, thus R9.4 was used instead which explains its lower perfomances on this task.
+Results are in \ref{table:1}
 
-See \url{https://github.com/nmiculinic/minion-basecaller/blob/master/notebook/Lambda_nanonet_vs_m270.ipynb for preliminary results.}
+\begin{table}[h!]
+\centering
+\begin{tabular}{lrrrrrrr}
+\toprule
+{} &  Deletion rate &  Error rate &  Insertion rate &  Match rate &  Mismatch rate &  Read length\\
+\midrule
+albacore     &       0.060 &    0.194 &        0.070 &    0.867 &       0.063 &  9843 \\
+nanonet      &       0.088 &    0.190 &        0.040 &    0.897 &       0.062 &  5029 \\
+mincall\_m270 &       0.077 &    0.172 &        0.040 &   0.905 &       0.056 &  9378 \\
+\bottomrule
+\end{tabular}
+
+\caption{Mean performance metrics on E.Coli dataset, 5k sample}
+\label{table:1}
+\end{table}
+
+We also tested and compared models on Human genome~\cite{jain2017nanopore} chromosone Y with R9.4 chemistry.
+Our model was only trained on R9 chemisty which explain drop in perfomance compared to Albacore.
+In our tests Nanonet produces significantly shorted read then data length and generally unusable perfomance on this dataset despite supporting R9.4 chemisty used in dataset.
 

### 2017-05-14 mincall/tools/analyse_bc.py

@@ -18,6 +18,7 @@ args = argparse.ArgumentParser()
 args.add_argument("input_folder", help="Input fast5 folder to basecall")
 args.add_argument("out_folder", help="Output folder for all analysis")
 args.add_argument("--ref", help="Reference genome")
+args.add_argument("--name", help="Run name")
 args.add_argument("--min_length", type=int, default=500, help="Minimum read lenght for further analysis")
 args.add_argument("-c", "--circular", help="Is genome circular", action="store_true")
 args.add_argument("--coverage_threshold", help="Minimal coverage threshold for consensus", type=float, default=0.0)
@@ -87,14 +88,28 @@ for name, cmd in basecallers.items():
     logger.info("%s consensus_report:\n%s", name, consensus_report)
 
 
-fig_kde_path = os.path.join(args.out_folder, name + "reads_kde.png")
-fig_hist_path = os.path.join(args.out_folder, name + "reads_hist.png")
+columns = list(iter(next(iter(dfs.values()))._get_numeric_data()))
+df_prep = []
+names = []
+for name, df in dfs.items():
+    names.append(name)
+    df_prep.append({col: df[col].mean() for col in columns})
+df_summary = pd.DataFrame(df_prep, index=names)
+df_summary.to_csv(os.path.join(args.out_folder, args.name + "_summary.csv"))
+df_summary.to_latex(os.path.join(args.out_folder, args.name + "_summary.tex"))
+del names
+del df_prep
+
+fig_kde_path = os.path.join(args.out_folder, args.name + "reads_kde.png")
+fig_hist_path = os.path.join(args.out_folder, args.name + "reads_hist.png")
 
 fig_kde, axes_kde = plt.subplots(3, 2)
 fig_hist, axes_hist = plt.subplots(3, 2)
 fig_kde.set_size_inches(12, 20)
 fig_hist.set_size_inches(12, 20)
-for col, ax_kde, ax_hist in zip(next(iter(dfs.values()))._get_numeric_data(), axes_kde.ravel(), axes_hist.ravel()):
+
+for col, ax_kde, ax_hist in zip(columns, axes_kde.ravel(), axes_hist.ravel()):
+    logging.info("Plotting column %s", col)
     for k in dfs.keys():
         sns.kdeplot(dfs[k][col], shade=True, label=k, alpha=0.5, ax=ax_kde)
         ax_hist.hist(dfs[k][col], label=k, alpha=0.5)

### 2017-05-14 paper/main.tex
Pre grammar check commit

@@
 
-As decribed in previous section, model was tested on E.Coli test set and compared to open source Nanonet and Albacore.
+As described in previous section, model was tested on E.Coli test set and compared to open source Nanonet and Albacore.
 Albacore doesn't have specific R9 chemistry mode, thus R9.4 was used instead which explains its lower perfomances on this task.
-Results are in \ref{table:1}
+Mean CIGAR operation are in table\ref{table:1} and KDE Match rate plot is in figure~\ref{fig:ecoli_match}.
 
@@
 
-We also tested and compared models on Human genome~\cite{jain2017nanopore} chromosone Y with R9.4 chemistry.
+\begin{figure}[t]
+\centering
+\includegraphics[width=\textwidth]{ecoli_match}
+\label{fig:ecoli_match}
+\end{figure}
+
+% We also tested and compared models on Human genome~\cite{jain2017nanopore} chromosone Y with R9.4 chemistry.
 Our model was only trained on R9 chemisty which explain drop in perfomance compared to Albacore.
@@
 
-TODO.
+First and foremost I'd like to thank my mentor izv.~prof.~dr.~sc.~Mile Sikić for setting up the problem, guiding us through its beginnings and providing helpful advice for its completion.
+Next Marko Ratkovic for its not only his insightful conversations, but also meaningful code contributions.
+
+Finally I'm thanking various other people whose code, tools and advice I've used in completing this paper: Fran Jurišić, Ana Marija Selak and Ivan Sović.
 
@@
 \bibliography{refs}
-
 \end{document}
-end{document}

### 2017-05-14 mincall/tools/analyse_bc.py

@@ -115,6 +115,7 @@ for col, ax_kde, ax_hist in zip(columns, axes_kde.ravel(), axes_hist.ravel()):
         sns.kdeplot(dfs[k][col], shade=True, label=k, alpha=0.5, ax=ax)
         sns.kdeplot(dfs[k][col], shade=True, label=k, alpha=0.5, ax=ax_kde)
         ax_hist.hist(dfs[k][col], label=k, alpha=0.5)
+    ax.set_title(args.name + " " + col)
     fig.savefig(os.path.join(args.out_folder, col + ".png"))
     for ax in [ax_kde, ax_hist]:
         ax.legend()

### 2017-05-14 paper/main.tex
Eng spelling passed

@@
 The model is based on deep learning and uses convolutional neural networks (CNN) in its implementation.
-For extra perfomnaces is uses cutting edge deep learning techniques and architectures, notably gated residual network, batch normalization and Connectionist Temporal Classification (CTC) loss.
+For extra performances is uses cutting edge deep learning techniques and architectures, notably gated residual network, batch normalization and Connectionist Temporal Classification (CTC) loss.
 
@@
 
-Oxford Nanopore Technology's MinION nanopore sequecing platform~\cite{mikheyev2014first} is the first portable DNA sequecing device.
-It's small weight, of only 90 grms, low capial cost, and long read length combided with high-throughtput, real time data analysis, and decent accuracy yield promissing resultins in various applications.
-From clinical application such as monitoring infectious desease outbreaks~\cite{judge2015early}\cite{quick2016real}, characterizing structural variants in cancer\cite{norris2016nanopore} and even full human genome assembly~\cite{jain2017nanopore}.
+Oxford Nanopore Technology's MinION nanopore sequencing platform~\cite{mikheyev2014first} is the first portable DNA sequencing device.
+It's small weight, of only 90 grams, low capital cost, and long read length combined with high-throughput, real-time data analysis, and decent accuracy yield promising results in various applications.
+From clinical application such as monitoring infectious disease outbreaks~\cite{judge2015early}\cite{quick2016real}, characterizing structural variants in cancer\cite{norris2016nanopore} and even full human genome assembly~\cite{jain2017nanopore}.
 
@@
 A MinION device typically yields reads several thousand bases long; reads as long as 100,000 bp have been reported.
-1D read has usually error rate about   10\%, though it varies across diferent genomes, human one being lower accuracy than E.Coli one.
+1D read has usually error rate about   10\%, though it varies across different genomes, human one being lower accuracy than E.Coli one.
 (Check details!!, What is the error rate, how it's defined???) After parameter training, base calling can be performed by running the Viterbi algorithm, which will result in the sequence of states with the highest likelihood.
@@
 The exact error rate metric is unreliable since multiple pipeline tools could be the issue.
-First the sample is prepared, hopefully uncontaminated and matching reference genome as close as possibe, then sequenced using the MinION device obtaining raw data.
-Next our model (or other groups ones) come along, basecall the sequence.
-To evalute error rate metric basecalled read is aligned to the reference genome using aligners with their own erros/biases, mostly commonly used BWA-MEM~\cite{li2013aligning} and Graphmap~\cite{sovic2016fast}.
+First the sample is prepared, hopefully, uncontaminated and matching reference genome as close as possible, then sequenced using the MinION device obtaining raw data.
+Next, our model (or other groups ones) come along, basecall the sequence.
+To evaluate error rate metric basecalled read is aligned to the reference genome using aligners with their own errors/biases, mostly commonly used BWA-MEM~\cite{li2013aligning} and Graphmap~\cite{sovic2016fast}.
 
@@
 A single DNA strand passes through the pore.
-Optionally, hairpin protein adapter can merge two DNA strands, allowing both template and complemente read passing through the nanopore sequentially for more accurate reads.
-This tehnique is refered as 2D reads, while we focus on 1D reads containing only template DNA and no hairpin adapter.
+Optionally, hairpin protein adapter can merge two DNA strands, allowing both template and complement read passing through the nanopore sequentially for more accurate reads.
+This technique is referred as 2D reads, while we focus on 1D reads containing only template DNA and no hairpin adapter.
 
-Electrical current runs through the nanopore, and exact nucelotides context within influences the nanopore's resistance.
+Electrical current runs through the nanopore and exact nucleotides context within influences the nanopore's resistance.
 This resistive effect is our sensor data, that is the current fluctuations as DNA passes though the pore.
@@
 
-Recent models opted using (Recurrent neural network) RNN, notably DeepNano~\cite{deepnano} and recently open sourced official Nanonet from ONT.
+Recent models opted using (Recurrent neural network) RNN, notably, DeepNano~\cite{deepnano} and recently open sourced official Nanonet from ONT.
 
@@
 We opted out for gated residual network variant~\cite{savarese2016learning}.
-For loss we used CTC (Connectionist temporal classification)~\cite{graves2006connectionist} between basecallled and the target sequence.
+For loss, we used CTC (Connectionist temporal classification)~\cite{graves2006connectionist} between basecalled and the target sequence.
 The implementation used is open source warp-ctc~\cite{warpctc}.
@@
 Dataset was obtained from (Ask somebody, Marko doesn't know!).
-In this research, models were trained on the ecoli K-12 strand.
+In this research, models were trained on the E.
+Coli K-12 strand.
 (Should I cite someone for strand).
@@
 To correct for model errors, we use aligned information, that is reference genome and alignment data (cigar string) to correct that information, yielding finalized target sequence.
-For further performance we skip first and last training block, since most error aggregate on edges.
+For further performance, we skip first and last training block, since most error aggregate on edges.
 
-Adjacent matching bases are separated with surrogate one, for example AAA -> AA'A for reasons described in the following section.
+Adjacent matching bases are separated with surrogate one, for example, AAA -> AA'A for reasons described in the following section.
 
@@
 Best performing model used has 270 total layers, divided into 3 90-layer blocks.
-Between each 90 layers blocks MaxPool layer is inserted with receptive width of 2 and stride 2, to ease computation effort and add precision.
+Between each 90 layers blocks, MaxPool layer is inserted with the receptive width of 2 and stride 2, to ease computation effort and add precision.
 
 Each block consists of 30 gated residual layers, each residual layer composed of 3 sequential Relu-Batch Normalization-Conv1D layers.
-Each convolutional layer uses receptive width of 3 with 64 chanells as output throughout the model.
+Each convolutional layer uses the receptive width of 3 with 64 channels as output throughout the model.
 
@@
 
-As described in previous section, model was tested on E.Coli test set and compared to open source Nanonet and Albacore.
-Albacore doesn't have specific R9 chemistry mode, thus R9.4 was used instead which explains its lower perfomances on this task.
+As described in the previous section, the model was tested on E.Coli test set and compared to open source Nanonet and Albacore.
+Albacore doesn't have specific R9 chemistry mode, thus R9.4 was used instead which explains its lower performance on this task.
 Mean CIGAR operation are in table\ref{table:1} and KDE Match rate plot is in figure~\ref{fig:ecoli_match}.
@@
 
-% We also tested and compared models on Human genome~\cite{jain2017nanopore} chromosone Y with R9.4 chemistry.
-Our model was only trained on R9 chemisty which explain drop in perfomance compared to Albacore.
-In our tests Nanonet produces significantly shorted read then data length and generally unusable perfomance on this dataset despite supporting R9.4 chemisty used in dataset.
+% We also tested and compared models on Human genome~\cite{jain2017nanopore} chromosome Y with R9.4 chemistry.
+Our model was only trained on R9 chemistry which explain the drop in performance compared to Albacore.
+In our tests, Nanonet produces significantly shorter read then data length and generally unusable performance on this dataset despite supporting R9.4 chemistry used in the dataset.
 
@@
 First and foremost I'd like to thank my mentor izv.~prof.~dr.~sc.~Mile Sikić for setting up the problem, guiding us through its beginnings and providing helpful advice for its completion.
-Next Marko Ratkovic for its not only his insightful conversations, but also meaningful code contributions.
+Next Marko Ratkovic for his not only insightful conversations, but also meaningful code contributions.
 
-Finally I'm thanking various other people whose code, tools and advice I've used in completing this paper: Fran Jurišić, Ana Marija Selak and Ivan Sović.
+Finally, I'm thanking various other people whose code, tools and advice I've used in completing this paper: Fran Jurišić, Ana Marija Selak, and Ivan Sović.
 

### 2017-05-14 paper/main.tex
Grammar, hr

@@
 \begin{sazetak}
-    MinION uredaj firme Oxford Nanopore Technology je prvi portabilni DNK sekvencer.
-Sposoban je pročitati značajno duga čitanja, preko 100 kBp, no nažalost pati od značajno vise pogreške nego ostale metode.
+Uređaj MinION firme Oxford Nanopore Technology prvi je prenosivi sekvencer DNA.
+Sposoban je pročitati nizove značajno velike duljine, preko 100 kBp, no nažalost pati od značajno vise pogrešaka nego ostale metode sekvenciranja.
 
 U ovoj studiji kreirali smo MinCall, cjelokupni model očitanja nukleotida za MinION.
-Model je baziran na dubokom učenju i koristi konvolucijske neuronske mreže u svojoj implementaciji.
-Za najbolje performanse koristi najnovije tehnike dubokog učenja i arhitekture, specifično rezidualne neuronske mreže, normalizaciju nad grupom i CTC gubitak.
+Model je baziran na metodama dubokog učenja i koristi konvolucijske neuronske mreže u svojoj implementaciji.
+Za najbolje performanse koristi najnovije tehnike dubokog učenja i arhitekture, specifično rezidualne neuronske mreže, normalizaciju nad grupom i gubitak vezane vremenske klasifikacije.
 
-    Najperformantniji 270 slojeva dubok model postiže najbolje dosad performanse od 90.5\% preciznosti na E.Coli skupu za učenje koristeći R9 kemiju i 1D čitanja.
+Najuspješniji model, u koji je uključeno 270 slojeva neurona, postiže najbolje dosad zabilježene performan- se od 90.5\% preciznosti na skupu za učenje iz E.
+coli koristeći R9-kemiju nanopore i 1D-čitanja.
 \kljucnerijeci{Basecaller, MinION, R9, CNN, CTC, sekvenciniranje sljedece generacije}

### 2017-05-14 paper/main.tex
Fix and first draft is kinda ready

@@
 
-Finally, I'm thanking various other people whose code, tools and advice I've used in completing this paper: Fran Jurišić, Ana Marija Selak, and Ivan Sović.
+Finally, I'm thanking various other people whose code, tools and advice I've used in completing this paper: Fran Jurišić, Ana Marija Selak, Ivan Sović and Martin Šošić.
 

### 2017-05-15 mincall/tools/analyse_bc.py
Added metrochorn

@@ -27,8 +27,9 @@ args = args.parse_args()
 basecallers = {
     "mincall_m270": ["nvidia-docker", "run", "--rm", "-v", "%s:/data" % args.input_folder, "-u=%d" % os.getuid(), "nmiculinic/mincall:9947283"],
     "nanonet": ["nanonetcall", args.input_folder, "--chemistry", "r9", "--platforms", "nvidia:0:20", "--exc_opencl"],
-    "albacore": None
-    # "nanonet": ["nanonetcall", args.input_folder, "--chemistry", "r9", "--jobs", "8"]
+    "albacore": None,
+    # "nanonet": ["nanonetcall", args.input_folder, "--chemistry", "r9", "--jobs", "8"],
+    'metrichorn': ["poretools", "fastq", "--type fwd", args.input_folder]
 }
 
 os.makedirs(args.out_folder, exist_ok=True)
@@ -36,7 +37,8 @@ dfs = {}
 for name, cmd in basecallers.items():
     logger = logging.getLogger(name)
     fasta_path = os.path.join(args.out_folder, name + ".fa")
-    if os.path.isfile(fasta_path):
+    fastq_path = os.path.join(args.out_folder, name + ".fastq")
+    if os.path.isfile(fasta_path) or os.path.isfile(fastq_path):
         logger.info("%s exists, skipping", fasta_path)
     else:
         logger.info("Basecalling %s with %s", args.input_folder, name)
@@ -49,8 +51,11 @@ for name, cmd in basecallers.items():
     if os.path.isfile(sam_path):
         logger.info("%s exists, skipping", sam_path)
     else:
-        logger.info("Aligning %s to reference %s with graphmap", fasta_path, args.ref)
-        align_utils.align_with_graphmap(fasta_path, args.ref, args.circular, sam_path)
+        path = fasta_path
+        if os.path.exists(fastq_path):
+            path = fastq_path
+        logger.info("Aligning %s to reference %s with graphmap", path, args.ref)
+        align_utils.align_with_graphmap(path, args.ref, args.circular, sam_path)
 
     filtered_sam = os.path.join(args.out_folder, name + "_filtered.sam")
     if os.path.isfile(filtered_sam):

### 2017-05-15 mincall/tools/analyse_bc.py
Update

@@ -69,15 +69,15 @@ for name, cmd in basecallers.items():
     reads_csv = os.path.join(args.out_folder, name + "_read_data.csv")
     if os.path.isfile(reads_csv):
         logger.info("%s file exists, loading", reads_csv)
-        df = pd.read_csv(reads_csv)
+        df = pd.read_pickle(os.path.join(args.out_folder, name + "_read_summary.pkl"))
     else:
         df = error_rates_for_sam(filtered_sam)
         df.to_csv(reads_csv)
+        df.to_pickle(os.path.join(args.out_folder, name + "_read_summary.pkl"))
         desc = df.describe()
-        with open(os.path.join(args.out_folder, name + "_read_summary.tex"), 'w') as f:
-            desc.to_latex(f)
-        with open(os.path.join(args.out_folder, name + "_read_summary.txt"), 'w') as f:
-            desc.to_string(f)
+        desc.to_latex(os.path.join(args.out_folder, name + "_read_summary.tex"))
+        desc.to_string(os.path.join(args.out_folder, name + "_read_summary.txt"))
+    df = error_rates_for_sam(filtered_sam)
     logger.info("%s\n%s", name, df.describe())
     dfs[name] = df
 

### 2017-05-15 mincall/tools/analyse_bc.py
Updated analysis_bc

@@ -18,6 +18,7 @@ args = argparse.ArgumentParser()
 args.add_argument("input_folder", help="Input fast5 folder to basecall")
 args.add_argument("out_folder", help="Output folder for all analysis")
 args.add_argument("--ref", help="Reference genome")
+args.add_argument("--name", help="Run name")
 args.add_argument("--min_length", type=int, default=500, help="Minimum read lenght for further analysis")
 args.add_argument("-c", "--circular", help="Is genome circular", action="store_true")
 args.add_argument("--coverage_threshold", help="Minimal coverage threshold for consensus", type=float, default=0.0)
@@ -26,8 +27,9 @@ args = args.parse_args()
 basecallers = {
     "mincall_m270": ["nvidia-docker", "run", "--rm", "-v", "%s:/data" % args.input_folder, "-u=%d" % os.getuid(), "nmiculinic/mincall:9947283"],
     "nanonet": ["nanonetcall", args.input_folder, "--chemistry", "r9", "--platforms", "nvidia:0:20", "--exc_opencl"],
-    "albacore": None
-    # "nanonet": ["nanonetcall", args.input_folder, "--chemistry", "r9", "--jobs", "8"]
+    "albacore": None,
+    # "nanonet": ["nanonetcall", args.input_folder, "--chemistry", "r9", "--jobs", "8"],
+    'metrichorn': ["poretools", "fastq", "--type fwd", args.input_folder]
 }
 
 os.makedirs(args.out_folder, exist_ok=True)
@@ -35,7 +37,8 @@ dfs = {}
 for name, cmd in basecallers.items():
     logger = logging.getLogger(name)
     fasta_path = os.path.join(args.out_folder, name + ".fa")
-    if os.path.isfile(fasta_path):
+    fastq_path = os.path.join(args.out_folder, name + ".fastq")
+    if os.path.isfile(fasta_path) or os.path.isfile(fastq_path):
         logger.info("%s exists, skipping", fasta_path)
     else:
         logger.info("Basecalling %s with %s", args.input_folder, name)
@@ -48,8 +51,11 @@ for name, cmd in basecallers.items():
     if os.path.isfile(sam_path):
         logger.info("%s exists, skipping", sam_path)
     else:
-        logger.info("Aligning %s to reference %s with graphmap", fasta_path, args.ref)
[…]

### 2017-05-15 mincall/tools/analyse_bc.py
Upodate

@@ -66,22 +66,20 @@ for name, cmd in basecallers.items():
         logger.info("Outputed filtered sam to %s\n%d kept, %d discarded",
                  filtered_sam, n_kept, n_discarded)
 
-    reads_csv = os.path.join(args.out_folder, name + "_read_data.csv")
-    if os.path.isfile(reads_csv):
-        logger.info("%s file exists, loading", reads_csv)
-        df = pd.read_pickle(os.path.join(args.out_folder, name + "_read_summary.pkl"))
+    reads_pkl = os.path.join(args.out_folder, name + "_read_data.pkl")
+    if os.path.isfile(reads_pkl):
+        logger.info("%s file exists, loading", reads_pkl)
+        df = pd.read_pickle(reads_pkl)
     else:
         df = error_rates_for_sam(filtered_sam)
-        df.to_csv(reads_csv)
-        df.to_pickle(os.path.join(args.out_folder, name + "_read_summary.pkl"))
-        desc = df.describe()
-        desc.to_latex(os.path.join(args.out_folder, name + "_read_summary.tex"))
-        desc.to_string(os.path.join(args.out_folder, name + "_read_summary.txt"))
-    df = error_rates_for_sam(filtered_sam)
-    logger.info("%s\n%s", name, df.describe())
+        df.to_pickle(reads_pkl)
+
+    desc = df.describe()
+    desc.to_latex(os.path.join(args.out_folder, name + "_read_summary.tex"))
+    logger.info("%s\n%s", name, desc)
     dfs[name] = df
 
-    consensus_report_path = os.path.join(args.out_folder, name + "consensus_report.csv")
+    consensus_report_path = os.path.join(args.out_folder, name + "_consensus_report.csv")
     if os.path.isfile(consensus_report_path):
         consensus_report = pd.read_csv(consensus_report_path)
         logger.info("%s exists, loading", consensus_report_path)

### 2017-05-15 mincall/tools/analyse_bc.py
Update analysis

@@ -32,6 +32,8 @@ basecallers = {
     'metrichorn': ["poretools", "fastq", "--type fwd", args.input_folder]
 }
 
+consensus_reports = []
+
 os.makedirs(args.out_folder, exist_ok=True)
 dfs = {}
 for name, cmd in basecallers.items():
@@ -79,17 +81,20 @@ for name, cmd in basecallers.items():
     logger.info("%s\n%s", name, desc)
     dfs[name] = df
 
-    consensus_report_path = os.path.join(args.out_folder, name + "_consensus_report.csv")
+    consensus_report_path = os.path.join(args.out_folder, name + "_consensus_report.pkl")
     if os.path.isfile(consensus_report_path):
-        consensus_report = pd.read_csv(consensus_report_path)
+        consensus_report = pd.read_pickle(consensus_report_path)
         logger.info("%s exists, loading", consensus_report_path)
     else:
-        consensus_report = get_consensus_report(filtered_sam, args.ref, args.coverage_threshold)
-        consensus_report.to_csv(consensus_report_path)
-        with open(os.path.splitext(consensus_report_path)[0] + ".txt", 'w') as f:
-            consensus_report.to_string(f)
+        consensus_report = get_consensus_report(name, filtered_sam, args.ref, args.coverage_threshold)
+        consensus_report.to_pickle(consensus_report_path)
+    consensus_reports.append(consensus_report)
     logger.info("%s consensus_report:\n%s", name, consensus_report)
 
+consensus_reports = pd.concat(consensus_reports)
+logger.info("Consensus Reports \n%s", consensus_reports)
+consensus_reports.to_csv(os.path.join(args.out_dir, "consensus.csv"))
+consensus_reports.to_latex(os.path.join(args.out_dir, "consensus.tex"))
 
 columns = list(iter(next(iter(dfs.values()))._get_numeric_data()))
 df_prep = []

### 2017-05-15 mincall/tools/analyse_bc.py
Fixed bug

@@ -93,8 +93,8 @@ for name, cmd in basecallers.items():
 
 consensus_reports = pd.concat(consensus_reports)
 logger.info("Consensus Reports \n%s", consensus_reports)
-consensus_reports.to_csv(os.path.join(args.out_dir, "consensus.csv"))
-consensus_reports.to_latex(os.path.join(args.out_dir, "consensus.tex"))
+consensus_reports.to_csv(os.path.join(args.out_folder, "consensus.csv"))
+consensus_reports.to_latex(os.path.join(args.out_folder, "consensus.tex"))
 
 columns = list(iter(next(iter(dfs.values()))._get_numeric_data()))
 df_prep = []

### 2017-06-27 mincall/tools/align_ref.cpp
Set default gap to 1000

@@ -51,7 +51,7 @@ std::ostream& operator<<(std::ostream& stream,
 
 const std::string DATASET_NAME("/Analyses/MinCall/Logits");
 const std::string ALIGNMENT_OUTPUT("/Analyses/MinCall/RefAlignment");
-const int GAP_WIDTH=100;
+const int GAP_WIDTH=1000;
 
 std::map<char, int> n2i {
     {'A', 0},

### 2017-06-27 mincall/models/test_model.py
Simplified test_model

@@ -7,6 +7,7 @@ from mincall.ops import central_cut
 from mincall.util import sigopt_double
 from mincall.model_utils import Model
 from mincall.controller import control
+from mincall import input_readers
 
 load_dotenv(find_dotenv())
 
@@ -23,22 +24,7 @@ def model_fn(net, X_len, max_reach, block_size, out_classes, batch_size, dtype, 
         with tf.variable_scope("block%d" % block):
             for layer in range(1, 1 + 1):
                 with tf.variable_scope('layer_%d' % layer):
-                    res = net
-                    for sublayer in [1, 2]:
-                        res = batch_normalization(
-                            res, scope='bn_%d' % sublayer)
-                        res = tf.nn.relu(res)
-                        res = conv_1d(
-                            res,
-                            64,
-                            3,
-                            scope="conv_1d_%d" % sublayer,
-                            weights_init=variance_scaling_initializer(
-                                dtype=dtype)
-                        )
-                    k = tf.get_variable(
-                        "k", initializer=tf.constant_initializer(1.0), shape=[])
-                    net = tf.nn.relu(k) * res + net
+                    net = conv_1d(net, 32, 3)
             net = max_pool_1d(net, 2)
         net = tf.nn.relu(net)
 
@@ -68,6 +54,7 @@ def create_train_model(hyper, **kwargs):
         shrink_factor=8,
         dtype=tf.float32,
         model_fn=model_fn,
+        in_data=input_readers.MinCallAlignedRaw(),
         lr_fn=lambda global_step: tf.train.exponential_decay(
             hyper['initial_lr'], global_step, 100000, hyper['decay_factor']),
         hyper=hyper,

### 2017-06-27 mincall/input_readers.py
Initial commit for mincall_input_reader

@@ -5,6 +5,7 @@ import numpy as np
 import h5py
 from glob import glob
 from collections import defaultdict
+import sys
 
 from mincall import util
 from mincall import errors
@@ -31,8 +32,12 @@ def find_ref(fast5_path):
 
     return ref_path
 
+
 class InputReader():
     def input_fn(self):
+        raise NotImplemented
+
+    def get_signal(self, fast5_path):
         raise NotImplemented
 
 
@@ -245,6 +250,80 @@ class AlbacoreAlignedRaw(AlignedRawAbstract):
         )
 
 
+class MinCallAlignedRaw(InputReader):
+    def input_fn(self, *args, **kwargs):
+        return AlignedRawAbstract.input_fn(self, *args, **kwargs)
+
+    def preprocessSignal(self, signal):
+        return (signal - 646.11133) / 75.673653
+
+    def get_signal(self, fast5_path):
+        with h5py.File(fast5_path, 'r') as h5:
+            reads = h5['Raw/Reads']
+            target_read = list(reads.keys())[0]
+            h5_logits = h5['Analyses/MinCall/Logits']
+            shrink_factor = h5_logits.attrs.get("shrink_factor", 8)  # Default 8..TODO
+            signal_len = h5_logits.shape[0] * shrink_factor
+            signal = h5['Raw/Reads/' + target_read]['Signal']
+            start_pad = h5_logits.attrs['start_pad']
+            signal = signal[start_pad:start_pad + signal_len]
+
+        return self.preprocessSignal(signal).reshape(1, -1, 1), start_pad
+
+    def read_fast5_raw_ref(self, fast5_path, ref_path, block_size_x, block_size_y, num_blocks, n_samples_per_ref=1,
+                           verify_file=True):
+
+        with h5py.File(fast5_path, 'r') as h5:
+            reads = h5['Raw/Reads']
+            target_read = list(reads.keys())[0]
+            h5_logits = h5['Analyses/MinCall/Logits']
+            h5_refalignment = h5['Analyses/MinCall/RefAlignment']
+
+            shrink_factor = h5_logits.attrs.get("shrink_factor", 8)  # Default 8..TODO
+
+            signal = h5['Raw/Reads/' + target_read]['Signal']
+            start_pad = h5_logits.attrs['start_pad']
+            signal_len = h5_logits.shape[0] * shrink_factor
+            signal = signal[start_pad:start_pad + signal_len]
+            signal = self.preprocessSignal(signal)
+
+            # -2 for first and last that are always skipped
[…]

### 2017-06-27 mincall/input_readers.py
Fixed small but important bug

@@ -318,6 +318,7 @@ class MinCallAlignedRaw(InputReader):
                         if h5_refalignment[b] != prev:
                             y[i * block_size_y + y_len[i]] = h5_refalignment[b]
                             y_len[i] += 1
+                            prev = h5_refalignment[b]
 
                 yield x, x_len, y, y_len
 

### 2017-06-27 mincall/input_readers.py
More  bugfixes

@@ -314,7 +314,7 @@ class MinCallAlignedRaw(InputReader):
                 for i in range(num_blocks):
                     prev = -1
                     mult = block_size_x // shrink_factor
-                    for b in range(start_block * mult, mult * (start_block + num_blocks)):
+                    for b in range(mult * (start_block + i), mult * (start_block + i + 1)):
                         if h5_refalignment[b] != prev:
                             y[i * block_size_y + y_len[i]] = h5_refalignment[b]
                             y_len[i] += 1

### 2017-06-28 mincall/align_utils.py
% used instead of rates,

@@ -351,15 +351,16 @@ def split_aligments_in_sam(in_sam_path, out_sam_path=None):
     with pysam.AlignmentFile(in_sam_path, "r") as samfile, \
             pysam.AlignmentFile(tmp_sam_out, "w", template=samfile) as out_sam:
 
-        for x in tqdm(samfile.fetch()):
+        for idx, x in enumerate(samfile.fetch()):
             name = x.query_name
+
             if x.is_unmapped:
                 logging.warning("%s unmapped" % name)
                 out_sam.write(x)
                 continue
             try:
                 target = x.get_reference_sequence()
-            except ValueError:
+            except (ValueError , AssertionError, Exception):
                 logging.error("%s Mapped but reference len equals 0, md tag: %s", name, x.has_tag('MD'))
                 out_sam.write(x)
                 continue

### 2017-06-28 mincall/input_readers.py
Updated

@@ -9,7 +9,7 @@ import sys
 
 from mincall import util
 from mincall import errors
-from mincall.errors import MissingRNN1DBasecall, InsufficientDataBlocks, MinIONBasecallerException
+from mincall.errors import *
 
 root_dir_map = {
     'karla': '/hgst8TB/fjurisic/ecoli',
@@ -193,8 +193,8 @@ class AlignedRawAbstract(InputReader):
                     model.logger.error('in filename %s \n' % fast5_path, exc_info=True)
 
                 if total % 10000 == 0 or total in [10, 100, 200, 1000]:
-                    model.logger.debug(
-                        "read %d datapoints: %s ",
+                    model.logger.info(
+                        "{%s} read %d datapoints: %s ", subdir,
                         total,
                         " ".join(["{}: {:.2f}%".format(key, 100 * errors[key] / total) for key in sorted(errors.keys())])
                     )
@@ -276,8 +276,15 @@ class MinCallAlignedRaw(InputReader):
         with h5py.File(fast5_path, 'r') as h5:
             reads = h5['Raw/Reads']
             target_read = list(reads.keys())[0]
-            h5_logits = h5['Analyses/MinCall/Logits']
-            h5_refalignment = h5['Analyses/MinCall/RefAlignment']
+            try:
+                h5_logits = h5['Analyses/MinCall/Logits']
+            except:
+                raise MissingMincallLogits()
+
+            try:
+                h5_refalignment = h5['Analyses/MinCall/RefAlignment']
+            except:
+                raise MissingMincallAlignedRef()
 
             shrink_factor = h5_logits.attrs.get("shrink_factor", 8)  # Default 8..TODO
 

### 2017-06-28 paper/main.tex
Updated paper; still have to resolve citations

@@
-\documentclass[times, utf8, seminar, numeric]{fer}
+\documentclass[runningheads,a4paper]{llncs}
+\usepackage{makeidx}
 \usepackage{booktabs}
-\usepackage{mathtools, amsmath,amsfonts,amssymb, amsthm}
+\usepackage{mathtools}
+\usepackage{amsmath}
+\usepackage{amsfonts}
+% \usepackage{amssymb}
+% \usepackage{amsthm}
 \usepackage{acronym}
@@
 \usepackage{graphicx}
+\usepackage{cite}
+
 \graphicspath{ {Figures/} }
+\begin{document}
 
-\usepackage{tikz}
-\usetikzlibrary{shapes,arrows}
-\begin{document}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
+\frontmatter
 
-% - razvio si novi base caller za 1D nanopore reads koji je temeljen na konvolucijskim neuronskim mrežama
-% - reci koliko slojeva ima model
-% - kakve si rezultate dobio te da li su bolji od postojećih.
-% - koju si razinu greške dobio
-% - daj neko ime base calleru
-
-\voditelj{izv.~prof.~dr.~sc.~Mile Sikić}
-\title{MinCall --- MinION basecaller model pogonjen dubokim učenjem}
-\engtitle{MinCall --- MinION end2end deep learning basecaller}
-\author{Neven Miculinić}
-
+\title{MinCall --- MinION end2end deep learning basecaller}
+\author{Neven Miculinić \and Marko Ratković \and Mile Sikić}
+\institute{Faculty of
+Electrical Engineering and Computing (FER), Zagreb, Croatia}
 \maketitle
-\tableofcontents
-
-\begin{sazetak}
-Uređaj MinION firme Oxford Nanopore Technology prvi je prenosivi sekvencer DNA.
-Sposoban je pročitati nizove značajno velike duljine, preko 100 kBp, no nažalost pati od značajno vise pogrešaka nego ostale metode sekvenciranja.
-
-U ovoj studiji kreirali smo MinCall, cjelokupni model očitanja nukleotida za MinION.
-Model je baziran na metodama dubokog učenja i koristi konvolucijske neuronske mreže u svojoj implementaciji.
-Za najbolje performanse koristi najnovije tehnike dubokog učenja i arhitekture, specifično rezidualne neuronske mreže, normalizaciju nad grupom i gubitak vezane vremenske klasifikacije.
-
-Najuspješniji model, u koji je uključeno 270 slojeva neurona, postiže najbolje dosad zabilježene performan- se od 90.5\% preciznosti na skupu za učenje iz E.
-coli koristeći R9-kemiju nanopore i 1D-čitanja.
-\kljucnerijeci{Basecaller, MinION, R9, CNN, CTC, sekvenciniranje sljedece generacije}
-\end{sazetak}
+% \tableofcontents
 
@@
 
-\chapter{Introduction}
+\section{Introduction}
 
@@
 
-\chapter{Background}
+\section{Background}
 
@@
 
-\chapter{Method}
+\section{Method}
 
@@
 
-\chapter{Results}
+\section{Results}
 
@@
 
+\begin{table}[h!]
+\centering
+\begin{tabular}{lr}
+\toprule
+{} &  Base pairs/second\\
+\midrule
+albacore      &      38000 \\
+nanonet       &       Slow \\
+mincall\_m270 &       3774 \\
+\bottomrule
+\end{tabular}
+\caption{Speed}
+\label{table:speed}
+\end{table}
+
 \begin{figure}[t]
@@
 
-\chapter{Conclusion and further work}
+\section{Conclusion and further work}
 
@@
 
-\chapter{Acknowledgments}
+\section{Acknowledgments}
 
@@
 
-\bibliographystyle{fer}
 \bibliography{refs}

### 2017-06-28 mincall/input_readers.py
Implemeted feedback
Replaced outer i with _
Replaced logits.shape[0] * shrink_factor with precomputed signal_len

@@ -295,7 +295,7 @@ class MinCallAlignedRaw(InputReader):
             signal = self.preprocessSignal(signal)
 
             # -2 for first and last that are always skipped
-            num_blocks_max = h5_logits.shape[0] * shrink_factor // block_size_x - 2
+            num_blocks_max = signal_len // block_size_x - 2
             if num_blocks_max < num_blocks:
                 raise InsufficientDataBlocks("Has only {} blocks, while requesting {} + first and last"
                                              .format(num_blocks_max, num_blocks))
@@ -304,7 +304,7 @@ class MinCallAlignedRaw(InputReader):
             n_different_non_overlap_blocks = n_different_blocks // num_blocks
             n_samples = min(n_samples_per_ref, n_different_non_overlap_blocks)
 
-            for i in range(n_samples):
+            for _ in range(n_samples):
                 x = np.zeros([block_size_x * num_blocks, 1], dtype=np.float32)
                 x_len = min(len(signal), block_size_x * num_blocks)
 

### 2017-06-28 mincall/input_readers.py
Added aditional safety check

@@ -326,6 +326,8 @@ class MinCallAlignedRaw(InputReader):
                             y[i * block_size_y + y_len[i]] = h5_refalignment[b]
                             y_len[i] += 1
                             prev = h5_refalignment[b]
+                            if y_len[i] > block_size_y:
+                                raise BlockSizeYTooSmall()
 
                 yield x, x_len, y, y_len
 

### 2017-06-28 mincall/input_readers.pyx
Added cython support

@@ -10,6 +10,8 @@ import sys
 from mincall import util
 from mincall import errors
 from mincall.errors import *
+cimport numpy as np
+
 
 root_dir_map = {
     'karla': '/hgst8TB/fjurisic/ecoli',
@@ -273,6 +275,7 @@ class MinCallAlignedRaw(InputReader):
     def read_fast5_raw_ref(self, fast5_path, ref_path, block_size_x, block_size_y, num_blocks, n_samples_per_ref=1,
                            verify_file=True):
 
+        cdef int i, b, prev, mult
         with h5py.File(fast5_path, 'r') as h5:
             reads = h5['Raw/Reads']
             target_read = list(reads.keys())[0]

### 2017-06-28 mincall/errors.py
Mincall input reader (#1)
Merge mincall_input_reader and seminar

@@ -24,3 +24,11 @@ class ZeroLenY(MinIONBasecallerException):
 
 class RefFileNotFound(MinIONBasecallerException):
     pass
+
+
+class MissingMincallLogits(MinIONBasecallerException):
+    pass
+
+
+class MissingMincallAlignedRef(MinIONBasecallerException):
+    pass

### 2017-06-28 mincall/input_readers.pyx

@@ -5,10 +5,13 @@ import numpy as np
 import h5py
 from glob import glob
 from collections import defaultdict
+import sys
 
 from mincall import util
 from mincall import errors
-from mincall.errors import MissingRNN1DBasecall, InsufficientDataBlocks, MinIONBasecallerException
+from mincall.errors import *
+cimport numpy as np
+
 
 root_dir_map = {
     'karla': '/hgst8TB/fjurisic/ecoli',
@@ -31,8 +34,12 @@ def find_ref(fast5_path):
 
     return ref_path
 
+
 class InputReader():
     def input_fn(self):
+        raise NotImplemented
+
+    def get_signal(self, fast5_path):
         raise NotImplemented
 
 
@@ -188,8 +195,8 @@ class AlignedRawAbstract(InputReader):
                     model.logger.error('in filename %s \n' % fast5_path, exc_info=True)
 
                 if total % 10000 == 0 or total in [10, 100, 200, 1000]:
-                    model.logger.debug(
-                        "read %d datapoints: %s ",
+                    model.logger.info(
+                        "{%s} read %d datapoints: %s ", subdir,
                         total,
                         " ".join(["{}: {:.2f}%".format(key, 100 * errors[key] / total) for key in sorted(errors.keys())])
                     )
@@ -245,6 +252,91 @@ class AlbacoreAlignedRaw(AlignedRawAbstract):
         )
 
 
+class MinCallAlignedRaw(InputReader):
+    def input_fn(self, *args, **kwargs):
+        return AlignedRawAbstract.input_fn(self, *args, **kwargs)
+
+    def preprocessSignal(self, signal):
+        return (signal - 646.11133) / 75.673653
+
+    def get_signal(self, fast5_path):
+        with h5py.File(fast5_path, 'r') as h5:
+            reads = h5['Raw/Reads']
+            target_read = list(reads.keys())[0]
+            h5_logits = h5['Analyses/MinCall/Logits']
+            shrink_factor = h5_logits.attrs.get("shrink_factor", 8)  # Default 8..TODO
+            signal_len = h5_logits.shape[0] * shrink_factor
+            signal = h5['Raw/Reads/' + target_read]['Signal']
+            start_pad = h5_logits.attrs['start_pad']
+            signal = signal[start_pad:start_pad + signal_len]
+
+        return self.preprocessSignal(signal).reshape(1, -1, 1), start_pad
+
[…]

### 2017-06-28 mincall/model_utils.py

@@ -593,6 +593,7 @@ class Model():
                 h5[h5_path].attrs['model_logdir'] = self.log_dir
                 h5[h5_path].attrs['run_id'] = self.run_id
                 h5[h5_path].attrs['in_data_classname'] = type(self.in_data).__name__
+                h5[h5_path].attrs['shrink_factor'] = self.shrink_factor
 
                 fname = os.path.join(self.log_dir, 'model_hyperparams.json')
                 with open(fname, 'r') as f:

### 2017-06-28 mincall/models/selu.py

@@ -20,7 +20,7 @@ def model_fn(net, X_len, max_reach, block_size, out_classes, batch_size, dtype, 
         logits -> Unscaled logits tensor in time_major form, (block_size, batch_size, out_classes)
     """
 
-    net = batch_normalization(net, decay=0.99, scope="bn")
+    net = batch_normalization(net, decay=0.999, scope="bn")
 
     for block in range(1, 3):
         with tf.variable_scope("block%d" % block):
@@ -45,7 +45,7 @@ def model_setup_params(hyper):
         g=tf.Graph(),
         block_size_x=4 * 600,
         block_size_y=630,
-        in_data=input_readers.HMMAlignedRaw(),
+        in_data=input_readers.MinCallAlignedRaw(),
         num_blocks=3,
         batch_size=16,
         max_reach=8 * 20,  # 240
@@ -72,7 +72,7 @@ sigopt_params = [
 default_params = {
     'initial_lr': 0.0022528366355169436,
     'decay_factor': 0.000005,
-    'num_layers': 30,
+    'num_layers': 5,
 }
 
 

### 2017-06-28 setup.py

@@ -1,0 +1,6 @@
+from distutils.core import setup
+from Cython.Build import cythonize
+
+setup(
+    ext_modules=cythonize("mincall/input_readers.pyx")
+)

### 2017-06-29 mincall/models/selu.py
Added cython_generated.c to gitignore

@@ -1,3 +1,5 @@
+import pyximport; pyximport.install()
+
 import tensorflow as tf
 from tflearn.layers.conv import max_pool_1d, conv_1d
 from tflearn.layers.normalization import batch_normalization
@@ -46,7 +48,7 @@ def model_setup_params(hyper):
         block_size_x=4 * 600,
         block_size_y=630,
         in_data=input_readers.MinCallAlignedRaw(),
-        num_blocks=3,
+        num_blocks=4,
         batch_size=16,
         max_reach=8 * 20,  # 240
         queue_cap=300,
@@ -55,7 +57,7 @@ def model_setup_params(hyper):
         shrink_factor=4,
         dtype=tf.float32,
         model_fn=model_fn,
-        n_samples_per_ref=3,
+        n_samples_per_ref=5,
         lr_fn=lambda global_step: tf.train.exponential_decay(
             hyper['initial_lr'], global_step, 100000, hyper['decay_factor']),
         hyper=hyper,

### 2017-06-30 paper/main.tex
Small paper changes

@@
 \section{Introduction}
+In recent years, deep learning methods significantly improved the state-of-the-art in multiple domains such as computer vision, speech recognition, and natural language processing \cite{LeCun:1998:CNI:303568.303704}\cite{NIPS2012_4824}.
+In this paper, we present application of deep learning in the field of  Bioinformatics for DNA basecalling problem.
 
@@
 In this paper, we show that this error rate can be reduced by replacing the default base caller provided by the manufacturer with a properly trained neural network model.
-In the future new R9.5 chemistry and 1D\^2 reads supersede current models.
+In the future new R9.5 chemistry and 1D\^2 reads should supersede current models.
 
@@
 As the DNA moves through the pore, the context shifts, and the electric current changes.
-
-A MinION device typically yields reads several thousand bases long; reads as long as 100,000 bp have been reported.
-1D read has usually error rate about   10\%, though it varies across different genomes, human one being lower accuracy than E.Coli one.
-(Check details!!, What is the error rate, how it's defined???) After parameter training, base calling can be performed by running the Viterbi algorithm, which will result in the sequence of states with the highest likelihood.
+The MinION device typically yields reads several thousands bases long, even couple hundred thousand bases long reads were repoted.
+However the cost in on accuracy, signicifantly lower than older, more reliable and expensive sequencing methods.
 
@@
 
+% set style - not sure which one is official for papers
+% spbasic (springer basic makes sense)
+\bibliographystyle{spbasic}
+% name your BibTeX data base
 \bibliography{refs}

### 2017-06-30 paper/main.tex
Small update

@@
 
-\graphicspath{ {Figures/} }
+\graphicspath{ {Figures/}{imgs/} }
 \begin{document}
@@
 Conceptually, the MinION sequencer works as follows.
-First, DNA is sheared into fragments of 8–20 kbp and adapters are ligated to either end of the fragments.
+First, DNA is sheared into smaller DNA fragments and adapters are ligated to either end of the fragments.
 The resulting DNA fragments pass through a protein embedded in a membrane via a nanometre-sized channel, a nanopore.
@@
 The nanopore is 6 nucleotides wide, and many models use this information in their approaches, while we're opted out of this technicality.
+
+\begin{figure}[!ht]
+	\begin{center}
+		\includegraphics[width=0.6\textwidth]{./imgs/sequencing.png}
+		\caption{Depiction of shotgun sequencing}
+		\label{fg:sequencing}
+	\end{center}
+\end{figure}
+
+\begin{figure}[!ht]
+	\begin{center}
+		\includegraphics[width=0.7\textwidth]{./imgs/nanopore.png}
+
+		\caption[DNA strain being pulled through a nanopore]{DNA strain being pulled through a nanopore \protect\footnotemark}
+		\label{fg:nanopore}
+	\end{center}
+\end{figure}
+\footnotetext{Figure adapted from https://nanoporetech.com/how-it-works}
 

### 2017-06-30 mincall/models/selu.py
C/p Marko's materials

@@ -23,12 +23,11 @@ def model_fn(net, X_len, max_reach, block_size, out_classes, batch_size, dtype, 
     """
 
     net = batch_normalization(net, decay=0.999, scope="bn")
-
     for block in range(1, 3):
         with tf.variable_scope("block%d" % block):
             for layer in range(kwargs['num_layers']):
                 with tf.variable_scope('layer_%d' % layer):
-                    net = conv_1d(net, 64, 3)
+                    net = conv_1d(net, 64, 5)
                     net = selu(net)
             net = max_pool_1d(net, 2)
 
@@ -48,7 +47,7 @@ def model_setup_params(hyper):
         block_size_x=4 * 600,
         block_size_y=630,
         in_data=input_readers.MinCallAlignedRaw(),
-        num_blocks=3,
+        num_blocks=1,
         batch_size=16,
         max_reach=8 * 20,  # 240
         queue_cap=300,
@@ -57,7 +56,7 @@ def model_setup_params(hyper):
         shrink_factor=4,
         dtype=tf.float32,
         model_fn=model_fn,
-        n_samples_per_ref=5,
+        n_samples_per_ref=7,
         lr_fn=lambda global_step: tf.train.exponential_decay(
             hyper['initial_lr'], global_step, 100000, hyper['decay_factor']),
         hyper=hyper,
@@ -74,7 +73,7 @@ sigopt_params = [
 default_params = {
     'initial_lr': 0.0022528366355169436,
     'decay_factor': 0.000005,
-    'num_layers': 5,
+    'num_layers': 30,
 }
 
 

### 2017-06-30 paper/main.tex
Some extra citations

@@
 
-Although MinION is able to produce long reads, they have a high sequencing error rate.
+Although MinION is able to produce long reads~\cite{loman1-100k,loman2-800k}, they have a high sequencing error rate.
 This has been somewhat alleviated with new R9 pore model, replacing older R7 ones.
@@
 As the DNA moves through the pore, the context shifts, and the electric current changes.
-The MinION device typically yields reads several thousands bases long, even couple hundred thousand bases long reads were repoted.
+
+The MinION device typically yields reads several thousands bases long, even couple hundred thousand bases long reads were repoted~\cite{loman1-100k,loman2-800k}.
 However the cost in on accuracy, signicifantly lower than older, more reliable and expensive sequencing methods.

### 2017-06-30 paper/main.tex
Some changes

@@
 		\includegraphics[width=0.7\textwidth]{./imgs/nanopore.png}
-
 		\caption[DNA strain being pulled through a nanopore]{DNA strain being pulled through a nanopore \protect\footnotemark}
@@
 
-The core of the decoding process is the basecalling step. Official basecaller is Metrichor, previously performed in the cloud before being open-sourced under name Nanonet. (Citation, whom?)
+The core of the decoding process is the basecalling step. Nowdays there's multiple basecalling options, what official and unoffical ones.
 
-Earlier models were (Hidden Markov model) HMM-based where hidden state modeled DNA sequence of length 6 (6-mer) in the nanopore.
+Earlier models were Hidden Markov model(HMM)-based where hidden state modeled DNA sequence of length 6 (6-mer) in the nanopore.
 Pore models were used in computing emission probabilities.~\cite{loman2015complete,schreiber2015analysis,szalay2015novo,timp2012dna} and the recent open source HMM-based basecaller Nanocall~\cite{david2016nanocall}.
 
-Recent models opted using (Recurrent neural network) RNN, notably, DeepNano~\cite{deepnano} and recently open sourced official Nanonet from ONT.
+\textit{Metrichor} is an Oxford Nanopore company that offers cloud-based platform \textit{EPI2ME} for analysis of nanopore data.
+Initially, base calling was only available by uploading data to the platform - that being the reason why this basecaller is often called Metrichor even though it is a name of the company.
+
+With the release of R9 chemistry, this model was replaced by a more accurate recurrent neural network (RNN) implementation.
+Currently, Oxford Nanopore offers several RNN-based local basecaller versions under different names: Albacore, Nanonet and basecaller integrated into MinKNOW \cite{ont-basecallers}.
+
+\textit{Albacore} is basecaller by Oxford Nanopore Technologies ready for production and actively supported.
+It is available to the Nanopore Community served as a binary.
+The source code of Albacore was not provided and is only available through the ONT Developer Channel.
+Tool supports only R9.4 and future R9.5 version of the chemistry.
+For R9 tests in this paper we used R9.4 chemisty setting as instructed on ONT forums.
+
+\textit{Nanonet}\footnote{\url{https://github.com/nanoporetech/nanonet/}} uses the same neural network that is used in Albacore but it is continually under development and does contain features such as error handling or logging needed for production use.
+It uses \textit{CURRENNT} library for running neural networks.
+It supportes basecalling of both R9 and R9.4 chemistry versions.
+However in our experiments it was painfully slow, which was as expected due to it's classification as not production ready.
+
+\textit{Scrappie}\footnote{\url{https://github.com/nanoporetech/scrappie}} is another basecaller by Oxford Nanopore Technologies.
+Similar to Nanonet, it is the platform for ongoing development.
+Scrappie is reported to be the first basecaller  that specifically address homopolymer base calling.
+It became publicly available just recently in June, 2017 and supports R9.4 and future R9.5 data.
+
+\textit{DeepNano}~\cite{deepnano} is open source RNN basecalled whose model are trained with older R7 chemisty, and thus are excluded from later analysis, yet it's mentioned here for completeness (or attempt thereof).
 
@@
 
-Instead of opting for the traditional path using HMM or RNN we tried using CNN~(Convolutional neural networks)~\cite{lecun-98}, that is their residual version~\cite{he2016deep}.
+Instead of opting for the traditional path using HMM or newly adopted RNN we tried using CNN~(Convolutional neural networks)~\cite{lecun-98}, that is their residual version~\cite{he2016deep}.
 We opted out for gated residual network variant~\cite{savarese2016learning}.
@@
 \bibliographystyle{spbasic}
-% name your BibTeX data base
 \bibliography{refs}

### 2017-06-30 paper/main.tex
Moved to appendix

@@
 Pore models were used in computing emission probabilities.~\cite{loman2015complete,schreiber2015analysis,szalay2015novo,timp2012dna} and the recent open source HMM-based basecaller Nanocall~\cite{david2016nanocall}.
+Modern basecallers use RNN base models, and we opted out using CNN instead with beam search.
 
-\textit{Metrichor} is an Oxford Nanopore company that offers cloud-based platform \textit{EPI2ME} for analysis of nanopore data.
-Initially, base calling was only available by uploading data to the platform - that being the reason why this basecaller is often called Metrichor even though it is a name of the company.
-
-With the release of R9 chemistry, this model was replaced by a more accurate recurrent neural network (RNN) implementation.
-Currently, Oxford Nanopore offers several RNN-based local basecaller versions under different names: Albacore, Nanonet and basecaller integrated into MinKNOW \cite{ont-basecallers}.
-
-\textit{Albacore} is basecaller by Oxford Nanopore Technologies ready for production and actively supported.
-It is available to the Nanopore Community served as a binary.
-The source code of Albacore was not provided and is only available through the ONT Developer Channel.
-Tool supports only R9.4 and future R9.5 version of the chemistry.
-For R9 tests in this paper we used R9.4 chemisty setting as instructed on ONT forums.
-
-\textit{Nanonet}\footnote{\url{https://github.com/nanoporetech/nanonet/}} uses the same neural network that is used in Albacore but it is continually under development and does contain features such as error handling or logging needed for production use.
-It uses \textit{CURRENNT} library for running neural networks.
-It supportes basecalling of both R9 and R9.4 chemistry versions.
-However in our experiments it was painfully slow, which was as expected due to it's classification as not production ready.
-
-\textit{Scrappie}\footnote{\url{https://github.com/nanoporetech/scrappie}} is another basecaller by Oxford Nanopore Technologies.
-Similar to Nanonet, it is the platform for ongoing development.
-Scrappie is reported to be the first basecaller  that specifically address homopolymer base calling.
-It became publicly available just recently in June, 2017 and supports R9.4 and future R9.5 data.
-
-\textit{DeepNano}~\cite{deepnano} is open source RNN basecalled whose model are trained with older R7 chemisty, and thus are excluded from later analysis, yet it's mentioned here for completeness (or attempt thereof).
+We compared our model on R9 chemistry with Metrichorn (HMM based approach) and Albacore (RNN based approach). For detailed basecaller overview  see the appendix~\ref{app:basecallers}.
 
@@
 \bibliography{refs}
+\appendix
+\section{Basecallers}\label{app:basecallers}
+
+\textit{Metrichor} is an Oxford Nanopore company that offers cloud-based platform \textit{EPI2ME} for analysis of nanopore data.
+Initially, base calling was only available by uploading data to the platform - that being the reason why this basecaller is often called Metrichor even though it is a name of the company.
+
+With the release of R9 chemistry, this model was replaced by a more accurate recurrent neural network (RNN) implementation.
+Currently, Oxford Nanopore offers several RNN-based local basecaller versions under different names: Albacore, Nanonet and basecaller integrated into MinKNOW \cite{ont-basecallers}.
+
+\textit{Albacore} is basecaller by Oxford Nanopore Technologies ready for production and actively supported.
+It is available to the Nanopore Community served as a binary.
+The source code of Albacore was not provided and is only available through the ONT Developer Channel.
+Tool supports only R9.4 and future R9.5 version of the chemistry.
+For R9 tests in this paper we used R9.4 chemisty setting as instructed on ONT forums.
+
+\textit{Nanonet}\footnote{\url{https://github.com/nanoporetech/nanonet/}} uses the same neural network that is used in Albacore but it is continually under development and does contain features such as error handling or logging needed for production use.
+It uses \textit{CURRENNT} library for running neural networks.
+It supportes basecalling of both R9 and R9.4 chemistry versions.
+However in our experiments it was painfully slow, which was as expected due to it's classification as not production ready.
+
+\textit{Scrappie}\footnote{\url{https://github.com/nanoporetech/scrappie}} is another basecaller by Oxford Nanopore Technologies.
+Similar to Nanonet, it is the platform for ongoing development.
+Scrappie is reported to be the first basecaller  that specifically address homopolymer base calling.
+It became publicly available just recently in June, 2017 and supports R9.4 and future R9.5 data.
+
+\textit{Nanocall} \cite{David046086} was the first third-party open source basecaller for nanopore data.
+It uses HMM approach like the original R7 Metrichor.
+Nanocall does not support newer chemistries after R7.3.
+
+\textit{DeepNano} \cite{Boza2017}  was the first open-source basecaller based on neural networks.
+It uses bidirectional recurrent neural networks implemented in Python, using the Theano library.
+When released, originally only supported R7 chemistry, but support for R9 and R9.4 was added recently.
+
 \end{document}

### 2017-06-30 paper/main.tex
Small update
g

@@
 The implementation used is open source warp-ctc~\cite{warpctc}.
-Main computation framework is tensorflow~\cite{tensorflow2015-whitepaper}
+Main computation framework is tensorflow~\cite{tensorflow2015-whitepaper}.
+
+\subsection{CTC}
+In this subsection we go into more detail outlining CTC loss and how we applied it.
 
@@
 \section{Basecallers}\label{app:basecallers}
+Here is currated basecaller list:
 
+\subsection{Official}
 \textit{Metrichor} is an Oxford Nanopore company that offers cloud-based platform \textit{EPI2ME} for analysis of nanopore data.
@@
 
-\textit{Nanocall} \cite{David046086} was the first third-party open source basecaller for nanopore data.
+\subsection{Third-party basecallers}
+\textit{Nanocall}~\cite{David046086} was the first third-party open source basecaller for nanopore data.
 It uses HMM approach like the original R7 Metrichor.
@@
 
-\textit{DeepNano} \cite{Boza2017}  was the first open-source basecaller based on neural networks.
+\textit{DeepNano}~\cite{Boza2017}  was the first open-source basecaller based on neural networks.
 It uses bidirectional recurrent neural networks implemented in Python, using the Theano library.

### 2017-06-30 mincall/input_readers.pyx
Processing changes to MinCallAlignedRaw

@@ -257,7 +257,7 @@ class MinCallAlignedRaw(InputReader):
         return AlignedRawAbstract.input_fn(self, *args, **kwargs)
 
     def preprocessSignal(self, signal):
-        return (signal - 646.11133) / 75.673653
+        return (signal - 652.56447114751563) / 189.0927367080821
 
     def get_signal(self, fast5_path):
         with h5py.File(fast5_path, 'r') as h5:

### 2017-06-30 paper/main.tex
Added names
Removed gated residual

@@
 % \usepackage{amsthm}
+\usepackage[utf8]{inputenc}
 \usepackage{acronym}
@@
 \title{MinCall --- MinION end2end deep learning basecaller}
-\author{Neven Miculinić \and Marko Ratković \and Mile Sikić}
+\author{Neven Miculinić \and Marko Ratković \and Mile Sikić\\
+\texttt{\{neven.miculinic, marko.ratkovic, mile.sikic\}@fer.hr}}
+
 \institute{Faculty of
@@
 The model is based on deep learning and uses convolutional neural networks (CNN) in its implementation.
-For extra performances is uses cutting edge deep learning techniques and architectures, notably gated residual network, batch normalization and Connectionist Temporal Classification (CTC) loss.
+For extra performances is uses cutting edge deep learning techniques and architectures, batch normalization and Connectionist Temporal Classification (CTC) loss.
 
@@
 Instead of opting for the traditional path using HMM or newly adopted RNN we tried using CNN~(Convolutional neural networks)~\cite{lecun-98}, that is their residual version~\cite{he2016deep}.
-We opted out for gated residual network variant~\cite{savarese2016learning}.
 For loss, we used CTC (Connectionist temporal classification)~\cite{graves2006connectionist} between basecalled and the target sequence.

### 2017-06-30 paper/main.tex

@@
 \usepackage{cite}
+\usepackage{hyperref}
+% \hypersetup{pdftex,colorlinks=true,allcolors=blue
 
@@
 To evaluate error rate metric basecalled read is aligned to the reference genome using aligners with their own errors/biases, mostly commonly used BWA-MEM~\cite{li2013aligning} and Graphmap~\cite{sovic2016fast}.
-
 \section{Background}
@@
 \section{Method}
-
 Instead of opting for the traditional path using HMM or newly adopted RNN we tried using CNN~(Convolutional neural networks)~\cite{lecun-98}, that is their residual version~\cite{he2016deep}.
@@
 
-\subsection{CTC}
-In this subsection we go into more detail outlining CTC loss and how we applied it.
+\section{Data}
 
-\section{Data preprocessing}
-
-Dataset was obtained from (Ask somebody, Marko doesn't know!).
+Dataset used were E.Coli K-12 strands from~\cite{loman1-100k} and Lamba basecalling\footnote{Acquired from doc.
+dr.
+sc.
+Petra Korać i dr.sc.
+Paula Dobrinić}.
 In this research, models were trained on the E.
@@
 Adjacent matching bases are separated with surrogate one, for example, AAA -> AA'A for reasons described in the following section.
-
-\section{Residual arhitecture}
-
-We use gated residual network, that is layer function is like this: $f(x) = k g(x) + x$ Where $x$ is input layer and $k$ constant learnt during training.
-Specifically, $g$ used is multiple, 1--3 times in our experiments, Relu-BatchNorm-CNN layers.
-
-\section{CTC}
-
-This section gives a brief overview over CTC, for further detail and in depth explanation we recommend original paper~\cite{graves2006connectionist} or searching for contemporary blog posts.
-
-In CTC we have target sequecne $\mathbf{l}$ consisting from symbols of alphabet $\Sigma$. Our models outputs $\mathbf{x}$, and CTC loss maximizes $p(\mathbf{l}|\mathbf{x})$.
-
-$\mathbf{x}$ is consistend of $n$ independent discrete random variables, $X_i$  over domain $\Sigma \cup \text{Blank}$.
-Path $p$ through $\mathbf{x}$ is assigning for each $X_i$ single value.
-It's log probabity is $\log P(p) = \sum_i {\log P(X_i=x_i)}$ where $x_i$s define the path.
-
-We further have merging operator on path, $\text{merge}(x_1, x_2, \ldots, x_n)$ which merges together adjacent equivalent elements. For example $\text{merge}(AAAGC) = AGC$.
-
-Then $p(\mathbf{l}|\mathbf{x}) = \sum_{p\in\text{paths}(\mathbf{x}, \text{merged}(p) = \mathbf{l}}{P(p)}$ or less formally, $P(\mathbf{l}|\mathbf{x})$ is sum of all path probabilites which when merged give $\mathbf{l}$
-
-In out concrete $x$ application $\Sigma = \{\text{A G T C A' G' T' A'}\}$
-
-For decoding, we use beam search decoder, with beam width 100.
 
@@
 
+Also some data were obtained in cooperation with Biologists doc. dr. sc. Petra Korać and dr.sc. Paula Dobrinić. Other were procured publicly shared from Loman labs~\cite{loman1-100k}
+
 % set style - not sure which one is official for papers

### 2017-06-30 paper/main.tex
Bibliography style from official template added
Added updated initial dataset description

@@
 \usepackage{makeidx}
+\usepackage{makecell}
 \usepackage{booktabs}
 \usepackage{mathtools}
+\usepackage{footnote}
+\usepackage{nccmath}
 \usepackage{amsmath}
@@
 \usepackage{geometry}
-\usepackage{hyperref}
+\usepackage[hidelinks]{hyperref}
+% \hypersetup{pdftex,colorlinks=true,allcolors=blue
 \usepackage{float}
@@
 \usepackage{cite}
-\usepackage{hyperref}
-% \hypersetup{pdftex,colorlinks=true,allcolors=blue
+\usepackage{verbatimbox}
 
@@
 
-\section{Data}
+\section{Dataset}
 
@@
 Paula Dobrinić}.
-In this research, models were trained on the E.
-Coli K-12 strand.
-(Should I cite someone for strand).
-Data was split into train and test subset, such that aligned reads map to different reference genome parts.
-For initial model data bootstrapping metrichorn was used, version (Ask Someone?).
+Both used datasets show in table \ref{tbl:datasets} have been previously have passed through MinKNOW and had been basecalled by Metrichor.
+As 1D read analysis was the focus of this paper, only those reads were used.
 
-The fast5 input training files were further split into smaller training blocks, consisting of fixed block size on raw signal. For each block target sequence, basecalled data is used in following way.
+\begin{savenotes}
+	\begin{table}[htb]
+		\caption{Used datasets}
+		\label{tbl:datasets}
+		\centering
 
-We're using basecalled knowledge which tells us on each raw read part which 6-mer were currently in the nanopore.
-Using this data, we get the intermediate target sequence for each block.
-To correct for model errors, we use aligned information, that is reference genome and alignment data (cigar string) to correct that information, yielding finalized target sequence.
-For further performance, we skip first and last training block, since most error aggregate on edges.
+		\begin{tabular}{lcc| c}
+			\toprule
+			{} &  \thead{Number of reads} &   \thead{Total bases \lbrack bp\rbrack\footnote{Total number of bases calle by Metrichor}} &    \thead{Whole genome size \lbrack bp\rbrack} \\
+			\midrule
+			\textit{{E.
+Coli}}\footnote{R9 sequencing data from \url{http://lab.loman.net/2016/07/30/nanopore-r9-data-release/}, reference taken from \url{https://www.ncbi.nlm.nih.gov/nuccore/48994873}} & 164471 & 1 481 687 490 & 4 639 675\\
+			\textit{lambda}\footnote{Internal dataset acquired from doc.~dr.sc.~Petra Korać and dr.sc.~Paula Dobrinić, reference taken from \url{https://www.ncbi.nlm.nih.gov/nuccore/NC_001416.1}}   & 86 &  466 465 & 48 502  \\
+			\bottomrule
+		\end{tabular}
+	\end{table}
+\end{savenotes}
 
-Adjacent matching bases are separated with surrogate one, for example, AAA -> AA'A for reasons described in the following section.
+To help training process, the raw signal is split into smaller blocks that are used as inputs.
+For each Metrichor basecalled event is easy to determine the block it falls into using \textit{start} field.
+Using this information output given by Metrichor can be determined for each block.
+To correct errors produced by Metrichor and possibly increase the quality of data, each read is aligned to the reference.
+This is done using aligner GraphMap~\cite{sovic2016fast} that returns the best position in the genome, hopefully, the part of the genome from which read came from.
+Alignment part in the genome is used as a target.
+Using CIGAR string returned by aligner we can correct Metrichor data and get target output for each block.
+This process is shown in figure~\ref{fg:data_correction}.
+
+\begin{figure}[!ht]
+	\begin{center}
+		\includegraphics[width=1\textwidth]{./imgs/train_data_correction.png}
+		\caption{Dataset preparation}
+		\label{fg:data_correction}
+	\end{center}
+\end{figure}
+
+To eliminate the possibility of overfitting to the known reference, the model is trained and tested on reads from different organisms.
+Due to limited amount of public available raw nanopore sequence data, ecoli was \emph{divided} into two regions.
+Reads were split into train and test portions, depending on which region of ecoli they align.
+If read aligns inside first 70\% of the ecoli, it is placed into train set, and if it aligns to the second portion, it is placed into test set.
+Reads whose alignment overlaps train and test region are not used.
+Important to note that ecoli genome, and genomes of the majority of other bacteria, is cyclical, so reads with alignments that wrap over edges are also discarded.
+Total train set consist of over 110 thousand reads.
+Overview of the entire learning pipeline is shown in figure \ref{fg:train_pipe}.
+
+\begin{figure}[!ht]
+	\begin{center}
+		\includegraphics[width=0.7\textwidth]{./imgs/train_pipeline.png}
+		\caption{Training pipeline overview}
+		\label{fg:train_pipe}
+	\end{center}
+\end{figure}
+
+\begin{verbbox}
+    Target      :  A  G  A  A  A
+    Preprocessed:  A  G  A  A' A
+\end{verbbox}
+
+\begin{figure}[!h]
+    \centering
+    \theverbbox
+    \caption{Target nucleotide sequence preprocessing. A' is surrogate nucleotide during training. After basecalling surrogate nucleotied are converted to original form.}
+    \label{fig:data_preprocessing}
+\end{figure}
+
+Due to CTC merged nature during decoding, that is in best matched path adjacent duplicates are merged into one, we preprocess the target nucleotide sequence with surrogate nucleotides, such that each second repeated nucleotide is its surrogate.
+Example provided in figure~\ref{fig:data_preprocessing}.
 

### 2017-06-30 paper/main.tex
Adapted bib stuff
Add zero-mean and unit variance preporcessing step
Update

@@
 \usepackage{verbatimbox}
-
+% \usepackage{natbib}
 \graphicspath{ {Figures/}{imgs/} }
@@
 To evaluate error rate metric basecalled read is aligned to the reference genome using aligners with their own errors/biases, mostly commonly used BWA-MEM~\cite{li2013aligning} and Graphmap~\cite{sovic2016fast}.
-\section{Background}
 
@@
 Total train set consist of over 110 thousand reads.
-Overview of the entire learning pipeline is shown in figure \ref{fg:train_pipe}.
+Overview of the entire learning pipeline is shown in figure~\ref{fg:train_pipe}.
 
@@
     \theverbbox
-    \caption{Target nucleotide sequence preprocessing. A' is surrogate nucleotide during training. After basecalling surrogate nucleotied are converted to original form.}
+    \caption{Target nucleotide sequence preprocessing}
     \label{fig:data_preprocessing}
@@
 Example provided in figure~\ref{fig:data_preprocessing}.
+All raw input data were normalized to zero mean and unit variance as it yield superiour perfomance with neural networks.
 
@@
 % spbasic (springer basic makes sense)
-\bibliographystyle{spbasic}
+% \bibliographystyle{spbasic}
+\bibliographystyle{splncs03}
 \bibliography{refs}

### 2017-06-30 paper/main.tex
Finished rough model section draft

@@
 
-\section{Method}
-Instead of opting for the traditional path using HMM or newly adopted RNN we tried using CNN~(Convolutional neural networks)~\cite{lecun-98}, that is their residual version~\cite{he2016deep}.
-For loss, we used CTC (Connectionist temporal classification)~\cite{graves2006connectionist} between basecalled and the target sequence.
-The implementation used is open source warp-ctc~\cite{warpctc}.
-Main computation framework is tensorflow~\cite{tensorflow2015-whitepaper}.
-
 \section{Dataset}
@@
 \end{savenotes}
+
+\subsection{Data preprocessing}
 
@@
 
+\section{Method}
+Instead of opting for the traditional path using HMM or newly adopted RNN we tried using CNN~(Convolutional neural networks)~\cite{lecun-98}, that is their residual version~\cite{he2016deep}.
+For loss, we used CTC (Connectionist temporal classification)~\cite{graves2006connectionist} between basecalled and the target sequence.
+Other building blocks used are Batch normalization(BN)~\cite{BNORM} and pooling layers.
+No dropout~\cite{srivastava2014dropout} were used.
+We used open source warp-ctc~\cite{warpctc} GPU CTC loss implementaion and tensorflow~\cite{tensorflow2015-whitepaper}.
+
+The final model is a residual neural network consisting of 72 residual blocks BN\footnote{Batch normalization}-ELU\footnote{Exponential Linear Unit
+
+\begin{equation*}
+\text{ELU}(x)=
+    \begin{cases}
+        x, & \text{if}\ x>0 \\
+        \alpha (exp(x) - 1), & \text{otherwise}
+        \end{cases}    \\
+\end{equation*}
+
+}-CONV\footnote{1D convolutional layer}-BN-ELU-CONV, to grand total of 2 million parameters.
+The used model is a variant of architecture proposed in paper~\cite{identitet} with the difference of ELU being used as activation instead of ReLU as it is reported \cite{resnet-elu} to speeds up the learning process and improve accuracy as the depth increase.
+
+Each convolutional layer in this models uses 64 chanells with receptive field of 3.
+Because sequenced read is always shorter than the raw signal, pooling with kernel size two is used every 24 layers resulting in a reduction of dimensionality by factor 8.
+This is beneficial in faster learning, better generalization and increased basecalling speed.
+
+Training the model is the minimization of previously described CTC loss.
+It was done using Adam \cite{adam} with default parameters, and minibatch size 8.
+To prevent gradients exploding on bad inputs, they were clipped to a range [-2, 2].
+We observed no overfitting due to large dataset size.
+
 \section{Results}
-
-\subsection{Final model}
-
-Best performing model used has 270 total layers, divided into 3 90-layer blocks.
-Between each 90 layers blocks, MaxPool layer is inserted with the receptive width of 2 and stride 2, to ease computation effort and add precision.
-
-Each block consists of 30 gated residual layers, each residual layer composed of 3 sequential Relu-Batch Normalization-Conv1D layers.
-Each convolutional layer uses the receptive width of 3 with 64 channels as output throughout the model.
-
-\subsection{Performance tables}
-
-As described in the previous section, the model was tested on E.Coli test set and compared to open source Nanonet and Albacore.
-Albacore doesn't have specific R9 chemistry mode, thus R9.4 was used instead which explains its lower performance on this task.
-Mean CIGAR operation are in table\ref{table:1} and KDE Match rate plot is in figure~\ref{fig:ecoli_match}.
-
-\begin{table}[h!]
-\centering
-\begin{tabular}{lrrrrrrr}
-\toprule
-{} &  Deletion rate &  Error rate &  Insertion rate &  Match rate &  Mismatch rate &  Read length\\
-\midrule
-albacore     &       0.060 &    0.194 &        0.070 &    0.867 &       0.063 &  9843 \\
-nanonet      &       0.088 &    0.190 &        0.040 &    0.897 &       0.062 &  5029 \\
-mincall\_m270 &       0.077 &    0.172 &        0.040 &   0.905 &       0.056 &  9378 \\
-\bottomrule
-\end{tabular}
-
-\caption{Mean performance metrics on E.Coli dataset, 5k sample}
-\label{table:1}
-\end{table}
-
-\begin{table}[h!]
-\centering
-\begin{tabular}{lr}
-\toprule
-{} &  Base pairs/second\\
-\midrule
-albacore      &      38000 \\
-nanonet       &       Slow \\
-mincall\_m270 &       3774 \\
-\bottomrule
-\end{tabular}
-\caption{Speed}
-\label{table:speed}
-\end{table}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=\textwidth]{ecoli_match}
-\label{fig:ecoli_match}
-\end{figure}
-
-% We also tested and compared models on Human genome~\cite{jain2017nanopore} chromosome Y with R9.4 chemistry.
-Our model was only trained on R9 chemistry which explain the drop in performance compared to Albacore.
-In our tests, Nanonet produces significantly shorter read then data length and generally unusable performance on this dataset despite supporting R9.4 chemistry used in the dataset.
 

### 2017-06-30 paper/main.tex
Update

@@
 Training the model is the minimization of previously described CTC loss.
-It was done using Adam \cite{adam} with default parameters, and minibatch size 8.
+It was done using Adam~\cite{adam} with default parameters, and exponentially decaying learning rate starting from 1e-3 and decay rate of 5e-6 over 100k steps\footnote{We use tf.train.exponential\_decay where current learning rate, lr is $lr=\text{initial\_lr} \cdot \text{decay\_rate}^\frac{\text{global\_step}}{\text{decay\_step}}$} and minibatch size 8.
 To prevent gradients exploding on bad inputs, they were clipped to a range [-2, 2].

### 2017-06-30 paper/main.tex
Update

@@
 
+During testing we tried ReLu and PrELU~\cite{prelu} with no significat result difference.
+We also tried different channel numbers, receptive field width and various other hyperparameters during hyperparameter optimization and conclude after enough complexity, that is sufficient layers, all choices were performing similarly.
+
 \section{Results}

### 2017-06-30 paper/main.tex
Added initial results CIGAR section

@@
 
+Developed tool was compared with other available basecallers that support R9 chemistry.
+This includes third-party basically DeepNano and official basecallers by Oxford Nanopore (cloud-based Metrichor and Nanonet).
+The fact that ground truth is not known makes evaluation difficult.
+Different methods for evaluation were used to get clearer information about each basecaller.
+
+A portion of the read length that aligns as correctly is called match\_rate. Same goes for mismatches and insertions.
+Sum of all matches, mismatches, and insertions is equal to the reads length \ref{eq:read_len}.
+For specific details see appendix~\ref{app:evaluation:CIGAR}.
+Results on E.Coli test set with Graphmap aligner are shown in table~\ref{tbl:ecoli_rates}.
+Evaluation with BWM-MEM and on lambda can be found in appendix~\ref{app:evaluation:CIGAR} under tables~\ref{tbl:ecoli_rates_bwa},~\ref{tbl:lambda_rates} and~\ref{tbl:lambda_rates_bwa}.
+Furthermore we plot Kernel Density estimation(KDE) plots for each mentioned statistic on E.Coli dataset in figure~\ref{fg:ecoli_kde}.
+
+\begin{table}[]
+	\caption{Alignment specifications of Ecoli R9 basecalled reads using GraphMap}
+	\label{tbl:ecoli_rates}
+	\centering
+	\begin{tabular}{lcccc}
+		\toprule
+		{} &  \thead{Match \% \\(median)} &  \thead{Mismatch \% \\(median)} &  \thead{Insertion \% \\(median)} &  \thead{Deletion \% \\(median)} \\
+		\midrule
+		DeepNano   &                  90.254762 &                      6.452852 &                       \textbf{3.274420} &                     11.829965 \\
+		Metrichor  &                  90.560455 &                      5.688105 &                       3.660381 &                      8.328271 \\
+		Nanonet    &                  90.607674 &                      5.608912 &                       3.652791 &                      8.299046 \\
+		MinCall    &                  \textbf{91.408591} &                     \textbf{ 5.019141} &                       3.477739 &                      \textbf{7.471608 }\\
+		\bottomrule
+	\end{tabular}
+\end{table}
+
+\begin{figure}[!htb]
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{./imgs/results/ecoli/kde_cigar.png}
+		\caption{KDE plot for distribution of percentage of alimnment operations for ecoli}
+		\label{fg:ecoli_kde}
+	\end{center}
+\end{figure}
+
 \section{Conclusion and further work}
@@
 
+\section{Evaluation metrics}
+
+\subsection{CIGAR derived metrics}
+\label{app:evaluation:CIGAR}
+
+A portion of the read length that aligns as correctly is called match\_rate. Same goes for mismatches and insertions.
+Sum of all matches, mismatches, and insertions is equal to the reads length \ref{eq:read_len}.
+Here is the specific equation list:
+
+\begin{equation}
+\begin{gathered}
+\label{eq:read_len}
+read\_len =  n\_matches + n\_mismatches + n\_insertions
+\end{gathered}
+\end{equation}
+\begin{equation}
+\begin{gathered}
+match\_rate = \dfrac{n\_matches}{read\_length}
+\end{gathered}
+\end{equation}
+\begin{equation}
+\begin{gathered}
+missmatch\_rate = \dfrac{n\_mismatches}{read\_length}
+\end{gathered}
+\end{equation}
+\begin{equation}
+\begin{gathered}
+insertion\_rate = \dfrac{n\_insertions}{read\_length}
+\end{gathered}
+\end{equation}
+
+\begin{equation}
+\begin{gathered}
+match\_rate  + snp\_rate + insertion\_rate = 1
+\end{gathered}
+\end{equation}
+
+Deletion rate is defined as a total number of deletions in the alignment over the length of the aligned read.
+
+\begin{equation}
+\begin{gathered}
+deletion\_rate = \dfrac{n\_deletion}{read\_length}
+\end{gathered}
+\end{equation}
+
+\begin{table}[]
+	\caption{Alignment specifications of Ecoli R9 basecalled reads using BWA mem}
+	\label{tbl:ecoli_rates_bwa}
+	\centering
+	\begin{tabular}{lcccc}
+		\toprule
+		{} &  \thead{Match \% \\(median)} &  \thead{Mismatch \% \\(median)} &  \thead{Insertion \% \\(median)} &  \thead{Deletion \% \\(median)} \\
+		\midrule
+		DeepNano   &                  90.254762 &                      6.452852 &                       3.274420 &                     11.829965 \\
+		Metrichor  &                  90.595441 &                      6.869543 &                       2.531646 &                      7.567381 \\
+		Nanonet    &                  90.988989 &                      6.674760 &                     \textbf{  2.348552} &                      7.698530 \\
+		MinCall    &                 \textbf{ 91.470588} &                      \textbf{5.929204 }&                       2.477283 &                     \textbf{ 6.970362} \\
+		\bottomrule
+	\end{tabular}
+
+\end{table}
+\begin{table}[]
+	\caption{Alignment specifications of Lambda basecalled reads using GraphMap}
+	\label{tbl:lambda_rates}
+	\centering
+	\begin{tabular}{lcccc}
+		\toprule
+		{} &  \thead{Match \% \\(median)} &  \thead{Mismatch \% \\(median)} &  \thead{Insertion \% \\(median)} &  \thead{Deletion \% \\(median)} \\
+		\midrule
+		DeepNano   &                  86.997687 &                      9.623494 &                       3.442490 &                     16.052830 \\
+		Metrichor  &                  87.714988 &                      7.835052 &                       4.093851 &                     \textbf{10.757491} \\
+		Nanonet    &                  88.415611 &                      8.178372 &                       3.629653 &                     11.793022 \\
+		MinCall    &                 \textbf{ 89.694482} &                     \textbf{ 7.238095} &                       \textbf{3.078796} &                     13.450292 \\
+		\bottomrule
+	\end{tabular}
+
+\end{table}
+\begin{table}[]
+	\caption{Alignment specifications of lambda R9 basecalled reads using BWA mem}
+	\label{tbl:lambda_rates_bwa}
+	\centering
+	\begin{tabular}{lcccc}
+		\toprule
+		{} &  \thead{Match \% \\(median)} &  \thead{Mismatch \% \\(median)} &  \thead{Insertion \% \\(median)} &  \thead{Deletion \% \\(median)} \\
+		\midrule
+
+		DeepNano   &                  86.625973 &                     11.288361 &                       2.098225 &                     14.648308 \\
+		Metrichor  &                  87.294093 &                     10.109186 &                       2.376476 &                      \textbf{9.645323 }\\
+		Nanonet    &                  87.767037 &                     10.017598 &                       2.354248 &                     10.597232 \\
+		MinCall    &                 \textbf{ 89.049870} &                     \textbf{ 9.480883} &                       \textbf{1.615188 }&                     12.962441 \\
+		\bottomrule
+	\end{tabular}
+\end{table}
+
 \end{document}

### 2017-06-30 paper/main.tex

@@
 
+In appendix~\ref{app:evaluation:CIGAR} the histogram figure~\ref{fg:cigar_op_dist} shows how matches, mismatches, insertions, and deletions are distributed across the reads.
+It is shown that mismatches and insertion occur more frequently at the beginnings and the ends of the reads.
+This is not only the case for the developed basecaller, but all other show the same property.
+This could be due to lack of context information from both sides when edges are base called.
+
 \begin{table}[]
@@
 
+\begin{figure}[!ht]
+	\begin{center}
+		\includegraphics[width=1\textwidth]{./imgs/operation_distributed_r9.png}
+		\caption{Histogram of alignment operations over relative position inside of read}
+		\label{fg:cigar_op_dist}
+	\end{center}
+\end{figure}
+
 \end{document}

### 2017-06-30 paper/main.tex
Update

@@
 			\midrule
-			\textit{{E.
+			\emph{{E.
 Coli}}\footnote{R9 sequencing data from \url{http://lab.loman.net/2016/07/30/nanopore-r9-data-release/}, reference taken from \url{https://www.ncbi.nlm.nih.gov/nuccore/48994873}} & 164471 & 1 481 687 490 & 4 639 675\\
-			\textit{lambda}\footnote{Internal dataset acquired from doc.~dr.sc.~Petra Korać and dr.sc.~Paula Dobrinić, reference taken from \url{https://www.ncbi.nlm.nih.gov/nuccore/NC_001416.1}}   & 86 &  466 465 & 48 502  \\
+			\emph{lambda}\footnote{Internal dataset acquired from doc.~dr.sc.~Petra Korać and dr.sc.~Paula Dobrinić, reference taken from \url{https://www.ncbi.nlm.nih.gov/nuccore/NC_001416.1}}   & 86 &  466 465 & 48 502  \\
 			\bottomrule
@@
 To help training process, the raw signal is split into smaller blocks that are used as inputs.
-For each Metrichor basecalled event is easy to determine the block it falls into using \textit{start} field.
+For each Metrichor basecalled event is easy to determine the block it falls into using \emph{start} field.
 Using this information output given by Metrichor can be determined for each block.
@@
 
+\subsection{Per read metrics}
+\label{subs:read_metrics}
 A portion of the read length that aligns as correctly is called match\_rate. Same goes for mismatches and insertions.
 Sum of all matches, mismatches, and insertions is equal to the reads length \ref{eq:read_len}.
-For specific details see appendix~\ref{app:evaluation:CIGAR}.
+For specific details see Appendix~\ref{app:evaluation:CIGAR}.
 Results on E.Coli test set with Graphmap aligner are shown in table~\ref{tbl:ecoli_rates}.
@@
 
-In appendix~\ref{app:evaluation:CIGAR} the histogram figure~\ref{fg:cigar_op_dist} shows how matches, mismatches, insertions, and deletions are distributed across the reads.
+In Appendix~\ref{app:evaluation:CIGAR} the histogram figure~\ref{fg:cigar_op_dist} shows how matches, mismatches, insertions, and deletions are distributed across the reads.
 It is shown that mismatches and insertion occur more frequently at the beginnings and the ends of the reads.
@@
 
-\begin{figure}[!htb]
+\begin{figure}[]
 	\begin{center}
@@
 \end{figure}
+
+\subsection{Consensus metrics}
+
+In Subsection~\ref{subs:read_metrics} we showed metrics after the read is aligned of reference genome.
+However, if basecalled simply yield A irrespective of ground truth, it shall have 100\% match rate since it will definitely align somewhere.
+Therefore we reconstruct original genome from basecalled reads.
+Specific metrics definitions analysed are in Appendix~\ref{app:evaluation:consensus}.
+
+\subsubsection{Consensus from pileup}
+
+Instead of going through the whole assembly process, as we know the reference genome of the data used in these tests, we simply align all the reads to the genome, stack them on top of each other forming pileup of read bases.
+Using majority vote, dominant bases are called on each position.
+The resulting sequence is called consensus.
+When calling consensus for deletions,  there has to be a majority of deletions of the same length.
+Calling insertions has the additional condition, the majority has to agree on both length and the bases of insertion.
+Figure~\ref{fg:consensus} shows how consensus is called from pileup created from aligned reads.
+Pileup is stored in mpileup format.
+
+All models show a slight bias towards deletions than insertions, but this may be the limitation of technology as it has been reported that deletion and mismatch rates for nanopore data are ordinarily higher than insertion rates~\cite{sovic2016fast}.
+Results are shown in table~\ref{tbl:spec_ecoli}.
+Lambda table is deferred to appendix in table~\ref{tbl:spec_lambda}.
+
+\begin{figure}[]
+   \begin{center}
+       \includegraphics[width=0.8\textwidth]{./imgs/consnesus.png}
+       \caption{Consensus from pileup}
+       \label{fg:consensus}
+   \end{center}
+\end{figure}
+
+\begin{table}[]
+	\caption{Consensus specifications of Ecoli R9 basecalled reads}
+	\label{tbl:spec_ecoli}
+	\centering
+	\begin{tabular}{lcccccc}
+		\toprule
+		{} &  \thead{Total called\\\lbrack bp\rbrack} &  \thead{Correctly called\\\lbrack bp\rbrack} &  \thead{Match\\\%} &  \thead{Snp\\\%} &  \thead{Insertion\\\%} &  \thead{Deletion\\\%} \\
+		\midrule
+		DeepNano  &                  1510244.0 &                      1493242.0 &          98.8742 &         1.0044 &               0.1214 &              0.9041 \\
+		Metrichor &                  1515893.0 &                      1502588.0 &          99.1223 &         0.7464 &               0.1313 &              0.6300 \\
+		Nanonet   &                  1414237.0 &                      1385515.0 &          97.9691 &         1.5700 &               0.4609 &              1.5158 \\
+		resdeep   &                  1517828.0 &                      1506233.0 &          \textbf{99.2361} &         \textbf{0.6474} &               \textbf{0.1165} &             \textbf{ 0.5510 }\\
+		\bottomrule
+	\end{tabular}
+\end{table}
+
+\subsubsection{Consensus from de novo assembly}
+In this evaluation method, consensus sequence is not calculated from pileup, but by de novo genome assembly.
+For this task, fast and accurate de novo genome assembler \emph{ra}\footnote{\url{https://github.com/rvaser/ra}}~\cite{vaser} was used and obtained consensus sequence is compared to the reference using dnadiff present in the Mumer\footnote{\url{https://github.com/garviz/MUMmer}}.
+The length of the reference, consensus sequence, number of contigs and percentages of aligned bases from the reference to the query and vice versa are shown in the table \ref{tbl:assembly}.
+Average identity summarizes how closely does the assembled sequence match the reference.
+This is run on full ecoli sequence run for 1D template reads (\textasciitilde 160k reads), for both developed tools, Nanonet and Metrichor.
+Developed tool has shown a small increase in quality of the assembled sequence over Metrichor by offering longer consensus, higher identity percentage, and overall smaller edit distance\footnote{Calculated using \url{https://github.com/isovic/racon/blob/master/scripts/edcontigs.py}}.
+
+\begin{table}[htb]
+	\caption{Assembly and consensus results for ecoli}
+	\label{tbl:assembly}
+	\centering
+\begin{tabular}{lccc}
+\toprule
+&         Metrichor &           MinCall &          Nanonet \\
+\midrule
+\thead{Ref. genome size (bp)} &           4639675 &           4639675 &            4639675 \\
+\thead{Total bases (bp)}      &           4604806 &           \textbf{4614354} &          4600056 \\
+\thead{Contigs [\#]}           &                 1 &                 1 &                1 \\
+\thead{Aln. bases ref. (bp)}  &  4639641(100.00\%) &  4639612(100.00\%) &  4639031(99.99\%) \\
+\thead{Aln. bases query (bp)} &  4604787(100.00\%) &  4614351(100.00\%) &  4599745(99.99\%) \\
+\thead{Avg. Identity}         &             98.76 &             \textbf{99.06} &            98.47 \\
+\thead{Edit distance}         &             60418 &             \textbf{46686 }&            74341 \\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 \subsection{Official}
-\textit{Metrichor} is an Oxford Nanopore company that offers cloud-based platform \textit{EPI2ME} for analysis of nanopore data.
+\emph{Metrichor} is an Oxford Nanopore company that offers cloud-based platform \emph{EPI2ME} for analysis of nanopore data.
 Initially, base calling was only available by uploading data to the platform - that being the reason why this basecaller is often called Metrichor even though it is a name of the company.
@@
 
-\textit{Albacore} is basecaller by Oxford Nanopore Technologies ready for production and actively supported.
+\emph{Albacore} is basecaller by Oxford Nanopore Technologies ready for production and actively supported.
 It is available to the Nanopore Community served as a binary.
@@
 
-\textit{Nanonet}\footnote{\url{https://github.com/nanoporetech/nanonet/}} uses the same neural network that is used in Albacore but it is continually under development and does contain features such as error handling or logging needed for production use.
-It uses \textit{CURRENNT} library for running neural networks.
+\emph{Nanonet}\footnote{\url{https://github.com/nanoporetech/nanonet/}} uses the same neural network that is used in Albacore but it is continually under development and does contain features such as error handling or logging needed for production use.
+It uses \emph{CURRENNT} library for running neural networks.
 It supportes basecalling of both R9 and R9.4 chemistry versions.
@@
 
-\textit{Scrappie}\footnote{\url{https://github.com/nanoporetech/scrappie}} is another basecaller by Oxford Nanopore Technologies.
+\emph{Scrappie}\footnote{\url{https://github.com/nanoporetech/scrappie}} is another basecaller by Oxford Nanopore Technologies.
 Similar to Nanonet, it is the platform for ongoing development.
@@
 \subsection{Third-party basecallers}
-\textit{Nanocall}~\cite{David046086} was the first third-party open source basecaller for nanopore data.
+\emph{Nanocall}~\cite{David046086} was the first third-party open source basecaller for nanopore data.
 It uses HMM approach like the original R7 Metrichor.
@@
 
-\textit{DeepNano}~\cite{Boza2017}  was the first open-source basecaller based on neural networks.
+\emph{DeepNano}~\cite{Boza2017}  was the first open-source basecaller based on neural networks.
 It uses bidirectional recurrent neural networks implemented in Python, using the Theano library.
@@
 
+\subsection{Consensus metrics}
+\label{app:evaluation:consensus}
+
+Similarly, as before, match rate, mismatch rate, insertion and deletion rates are calculated but this time for whole consensus sequence.
+In this context, mismatches are called \emph{single nucleotide polymorphisms} (snp).
+
+\begin{equation}
+\begin{gathered}
+match\_rate = \dfrac{n\_correct\_bases}{consensus\_length}
+\end{gathered}
+\end{equation}
+\begin{equation}
+\begin{gathered}
+snp\_rate = \dfrac{n\_snp}{consensus\_length}
+\end{gathered}
+\end{equation}
+\begin{equation}
+\begin{gathered}
+insertion\_rate = \dfrac{n\_insertions}{consensus\_length}
+\end{gathered}
+\end{equation}
+
+\begin{equation}
+\begin{gathered}
+match\_rate  + snp\_rate + insertion\_rate = 1
+\end{gathered}
+\end{equation}
+
+\begin{equation}
+\begin{gathered}
+deletion\_rate = \dfrac{n\_deletion}{consensus\_length}
+\end{gathered}
+\end{equation}
+
+\begin{table}[]
+	\caption{Consensus specifications of lambda R9 basecalled reads}
+	\label{tbl:spec_lambda}
+	\centering
+	\begin{tabular}{lcccccc}
+		\toprule
+		{} &  \thead{Total called\\\lbrack bp\rbrack} &  \thead{Correctly called\\\lbrack bp\rbrack} &  \thead{Match\\\%} &  \thead{Snp\\\%} &  \thead{Insertion\\\%} &  \thead{Deletion\\\%} \\
+		\midrule
+		DeepNano  &                    48342.0 &                        48025.0 &          99.3443 &         0.6433 &               \textbf{0.0124} &              0.2648 \\
+		Metrichor &                    48469.0 &                        48257.0 &          \textbf{99.5626} &         \textbf{0.4188} &               0.0186 &              \textbf{0.1465} \\
+		Nanonet   &                    48438.0 &                        48168.0 &          99.4426 &         0.5409 &               0.0165 &              0.1961 \\
+		resdeep   &                    48385.0 &                        48163.0 &          99.5412 &         0.4402 &               0.0186 &              0.1976 \\
+		\bottomrule
+	\end{tabular}
+\end{table}
+
 \end{document}

### 2017-06-30 paper/main.tex
Removed shotgun sequecing figure
Update

@@
 
-    The best performing 270 layers deep model achieves state-of-the-art 90.5\% match rate on E.Coli dataset using R9 pore chemistry and 1D reads.
+    The best performing deep learning model achieves state-of-the-art median 91.4\% match rate on E.Coli dataset using R9 pore chemistry and 1D reads.
 \keywords{Basecaller, MinION, R9, CNN, CTC, Next generation sequecing}
@@
 Conceptually, the MinION sequencer works as follows.
+It's variation on now standard shotgun sequecing approach.
 First, DNA is sheared into smaller DNA fragments and adapters are ligated to either end of the fragments.
@@
 
-\begin{figure}[!ht]
-	\begin{center}
-		\includegraphics[width=0.6\textwidth]{./imgs/sequencing.png}
-		\caption{Depiction of shotgun sequencing}
-		\label{fg:sequencing}
-	\end{center}
-\end{figure}
+% \begin{figure}[!ht]
+% 	\begin{center}
+% 		\includegraphics[width=0.6\textwidth]{./imgs/sequencing.png}
+% 		\caption{Depiction of shotgun sequencing}
+% 		\label{fg:sequencing}
+% 	\end{center}
+% \end{figure}
 
@@
 
-This model used advance state-of-the-art gated residual convolutional neural network, with top models having 270 layers and over 3M parameters, yet improvements over Metrichorn baseline are marginal.
-As the conclusion, it might be that we've reached Bayesian error rate for R9 chemistry.
-Furthermore, R9.5 and 1D\^2 reads are under development which shall yield this paper's result obsolete quite soon, yet underlying code developed could easily be adjusted and trained on new data.
+In this paper we used CNN instead of already tried RNN or HMM approaches, which yielded higher accuracy than previously tried models.
+Unlike HMM and RNN, there's no explicit dependency on previous hidden state, therefore this model is massively parallizible and more sensible given data nature --- that is we're dealing with signal processing, not heavily context depended language modeling.
 
-Unlike Nanonet which uses custom OpenCL kernels or Albacore --- a novel ONT basecaller as of May 2017 lacking GPU support, this work used world-class computational framework tensorflow with highly optimized kernels and large development community.
-Therefore resulting paper's effect is showcasing Residual CNN approach or pure CNN approach with CTC loss is marginally better than already established basecaller and providing code in the contemporary framework.
+All test are done on data for R9 chemistry, but the developed open source code could easily be adjusted and trained on R9.4 and newest R9.5 data when it becomes publicly available.
+
+Currently, without support for newer sequencing data, this model has limited application.
+It can be used as a demonstration of a different approach to base calling which yields promising results.
+As newer versions of basecallers by Oxford Nanopore do not offer any support for data sequenced with previous version of chemistries, this tool can be used to re-basecall that data and improvement of the quality of reads retrospectively.
+
+Future work includes experiments with recently proposed Scaled exponential linear units (SELU)~\cite{selu} that eliminates the need for normalization techniques such as used batch normalization.
+Possible improvements of the model include the combination of convolutions and attention mechanism proposed just recently in the paper~\cite{facebook} showing excellent results in both speed and accuracy, for tasks of language translation.
+Last approach could use stacked simpled models, such as logistic regression and SVM to predict each nucleotide given raw signal context, similar to our deep learning model pre-CTC layer, and use linear chain CRF for full sequence basecalling.
+Approaches are many, and the time is limited.
 
@@
 
-Also some data were obtained in cooperation with Biologists doc. dr. sc. Petra Korać and dr.sc. Paula Dobrinić. Other were procured publicly shared from Loman labs~\cite{loman1-100k}
+Also some data were obtained in cooperation with Biologists doc.~dr.~sc. Petra Korać and dr.sc. Paula Dobrinić. Other were procured publicly shared from Loman labs~\cite{loman1-100k}
 

### 2017-06-30 paper/main.tex
Edited introduction/sequecing overview

@@
 
-Although MinION is able to produce long reads~\cite{loman1-100k,loman2-800k}, they have a high sequencing error rate.
+Although MinION is able to produce long reads, even up to 882 kb~\cite{loman1-100k,loman2-800k}, they have a high sequencing error rate.
 This has been somewhat alleviated with new R9 pore model, replacing older R7 ones.
-In this paper, we show that this error rate can be reduced by replacing the default base caller provided by the manufacturer with a properly trained neural network model.
-In the future new R9.5 chemistry and 1D\^2 reads should supersede current models.
-
-In the MinION device, single-stranded DNA fragments move through nanopores, which causes drops in the electric current.
-The electric current is measured at each pore several thousand times per second, 4000 times exactly in our dataset.
-The electric current depends mostly on the context of several DNA bases passing through the pore at the time of measurement.
-As the DNA moves through the pore, the context shifts, and the electric current changes.
-
-The MinION device typically yields reads several thousands bases long, even couple hundred thousand bases long reads were repoted~\cite{loman1-100k,loman2-800k}.
-However the cost in on accuracy, signicifantly lower than older, more reliable and expensive sequencing methods.
+In this paper, we show that this error rate can be reduced by our approach with properly trained neural network model.
 
@@
 
-Electrical current runs through the nanopore and exact nucleotides context within influences the nanopore's resistance.
-This resistive effect is our sensor data, that is the current fluctuations as DNA passes though the pore.
-The nanopore is 6 nucleotides wide, and many models use this information in their approaches, while we're opted out of this technicality.
+As DNA strand pass through the nanopore, they are propelled by the current.
+However this current varies depending on specific DNA nucletide context within the nanopore, changing its resistance.
+We sample the curreny multiple times per second, 4000Hz in our dataset, and from this data we deduce passing DNA fragement.
+By design, nanopore is 6 nucleotides wide, and many models use this information internally, yet we opted our of it in hopes of simpler model, less feature engineering and better homopolymer detection.
 
@@
 \section{Basecalling}
-
-The core of the decoding process is the basecalling step. Nowdays there's multiple basecalling options, what official and unoffical ones.
+The core of the decoding process is the basecalling step, that is translating the current sample to passed nucleotided sequence.
+Nowdays there's multiple basecalling options, what official and unoffical ones.
 

### 2017-07-01 paper/main.tex
Typo fixes, replaced ecoli to E. Coli in figures

@@
     The Oxford Nanopore Technologies's MinION is the first portable DNA sequencing device.
-It's capable of producing long reads, over 100 kBp were reported, however, is has significantly higher error rate than other methods.
+It's capable of producing long reads, over 100 kBp were reported, however, it has significantly higher error rate than other methods.
 
@@
 First the sample is prepared, hopefully, uncontaminated and matching reference genome as close as possible, then sequenced using the MinION device obtaining raw data.
-Next, our model (or other groups ones) come along, basecall the sequence.
-To evaluate error rate metric basecalled read is aligned to the reference genome using aligners with their own errors/biases, mostly commonly used BWA-MEM~\cite{li2013aligning} and Graphmap~\cite{sovic2016fast}.
+Next, our model (or any other) comes along and basecalls the sequences.
+To evaluate error rate metric basecalled read is aligned to the reference genome using aligners with their own errors/biases, most commonly used BWA-MEM~\cite{li2013aligning} and Graphmap~\cite{sovic2016fast}.
 
@@
 Conceptually, the MinION sequencer works as follows.
-It's variation on now standard shotgun sequecing approach.
+It's a variation on now standard shotgun sequencing approach.
 First, DNA is sheared into smaller DNA fragments and adapters are ligated to either end of the fragments.
@@
 
-As DNA strand pass through the nanopore, they are propelled by the current.
-However this current varies depending on specific DNA nucletide context within the nanopore, changing its resistance.
-We sample the curreny multiple times per second, 4000Hz in our dataset, and from this data we deduce passing DNA fragement.
-By design, nanopore is 6 nucleotides wide, and many models use this information internally, yet we opted our of it in hopes of simpler model, less feature engineering and better homopolymer detection.
+As DNA strand passes through the nanopore, they are propelled by the current.
+However, this current varies depending on specific nucleotide context within the nanopore, changing its resistance.
+We sample the currency multiple times per second, 4000Hz in our dataset, and from this data, we deduce passing DNA fragment.
+By design, the nanopore is 6 nucleotides wide, and many models use this information internally, yet we opted ours of it in hopes of a simpler model, less feature engineering, and better homopolymer detection.
 
@@
 \section{Basecalling}
-The core of the decoding process is the basecalling step, that is translating the current sample to passed nucleotided sequence.
-Nowdays there's multiple basecalling options, what official and unoffical ones.
+The core of the decoding process is the basecalling step, that is translating the current samples to the nucleotide sequence.
+Nowadays there are multiple basecalling options, what official and unofficial ones.
 
@@
 
-We compared our model on R9 chemistry with Metrichorn (HMM based approach) and Albacore (RNN based approach). For detailed basecaller overview see the appendix~\ref{app:basecallers}.
+We compared our model on R9 chemistry with Metrichor (HMM based approach), Nanonet and DeepNano (RNN based approaches). For detailed basecaller overview see the appendix~\ref{app:basecallers}.
 
@@
 Paula Dobrinić}.
-Both used datasets show in table \ref{tbl:datasets} have been previously have passed through MinKNOW and had been basecalled by Metrichor.
+Both used datasets show in table \ref{tbl:datasets} have been previously passed through MinKNOW and had been basecalled by Metrichor.
 As 1D read analysis was the focus of this paper, only those reads were used.
@@
 To eliminate the possibility of overfitting to the known reference, the model is trained and tested on reads from different organisms.
-Due to limited amount of public available raw nanopore sequence data, ecoli was \emph{divided} into two regions.
-Reads were split into train and test portions, depending on which region of ecoli they align.
-If read aligns inside first 70\% of the ecoli, it is placed into train set, and if it aligns to the second portion, it is placed into test set.
+Due to the limited amount of public available raw nanopore sequence data, E.
+Coli was \emph{divided} into two regions.
+Reads were split into train and test portions, depending on which region of E. Coli they align.
+If read aligns inside first 70\% of the E.
+Coli, it is placed into train set, and if it aligns to the second portion, it is placed into test set.
 Reads whose alignment overlaps train and test region are not used.
-Important to note that ecoli genome, and genomes of the majority of other bacteria, is cyclical, so reads with alignments that wrap over edges are also discarded.
+Important to note that E.
+Coli genome, and genomes of the majority of other bacteria, is cyclical, so reads with alignments that wrap over edges are also discarded.
 Total train set consist of over 110 thousand reads.
@@
 
-Due to CTC merged nature during decoding, that is in best matched path adjacent duplicates are merged into one, we preprocess the target nucleotide sequence with surrogate nucleotides, such that each second repeated nucleotide is its surrogate.
+Due to CTC merged nature during decoding, adjacent duplicates are merged into one, we preprocess the target nucleotide sequence with surrogate nucleotides, such that each second repeated nucleotide is replaced with its surrogate.
 Example provided in figure~\ref{fig:data_preprocessing}.
-All raw input data were normalized to zero mean and unit variance as it yield superiour perfomance with neural networks.
+All raw input data were normalized to zero mean and unit variance as it yields superior performance with neural networks.
 
@@
 Other building blocks used are Batch normalization(BN)~\cite{BNORM} and pooling layers.
-No dropout~\cite{srivastava2014dropout} were used.
-We used open source warp-ctc~\cite{warpctc} GPU CTC loss implementaion and tensorflow~\cite{tensorflow2015-whitepaper}.
+Dropout~\cite{srivastava2014dropout} was not used.
+Described model is implemented using tensorflow~\cite{tensorflow2015-whitepaper} and source warp-ctc~\cite{warpctc} GPU CTC loss implementation.
 
@@
 
-}-CONV\footnote{1D convolutional layer}-BN-ELU-CONV, to grand total of 2 million parameters.
+}-CONV\footnote{1D convolutional layer}-BN-ELU-CONV, to a grand total of 2 million parameters.
 The used model is a variant of architecture proposed in paper~\cite{identitet} with the difference of ELU being used as activation instead of ReLU as it is reported \cite{resnet-elu} to speeds up the learning process and improve accuracy as the depth increase.
 
-Each convolutional layer in this models uses 64 chanells with receptive field of 3.
+Each convolutional layer in this models uses 64 channels with kernel size 3.
 Because sequenced read is always shorter than the raw signal, pooling with kernel size two is used every 24 layers resulting in a reduction of dimensionality by factor 8.
@@
 
-During testing we tried ReLu and PrELU~\cite{prelu} with no significat result difference.
-We also tried different channel numbers, receptive field width and various other hyperparameters during hyperparameter optimization and conclude after enough complexity, that is sufficient layers, all choices were performing similarly.
+During testing, we tried ReLu and PrELU~\cite{prelu} with no significant result difference.
+We also tried different channel numbers, receptive field width, and various other hyperparameters during hyperparameter optimization and conclude after enough complexity, that is sufficient layers, all choices were performing similarly.
 
@@
 
-In Subsection~\ref{subs:read_metrics} we showed metrics after the read is aligned of reference genome.
-However, if basecalled simply yield A irrespective of ground truth, it shall have 100\% match rate since it will definitely align somewhere.
-Therefore we reconstruct original genome from basecalled reads.
+In Subsection~\ref{subs:read_metrics} we showed metrics after the read is aligned to reference genome.
+However, if basecalled simply yield nucleotide sequence \emph{A} irrespective of ground truth, it shall have 100\% match rate since it will definitely align somewhere.
+Therefore we reconstruct the original genome from basecalled reads.
 Specific metrics definitions analysed are in Appendix~\ref{app:evaluation:consensus}.
@@
 \subsubsection{Consensus from de novo assembly}
-In this evaluation method, consensus sequence is not calculated from pileup, but by de novo genome assembly.
+
+In this evaluation method, the consensus sequence is not calculated from the pileup, but by de novo genome assembly.
 For this task, fast and accurate de novo genome assembler \emph{ra}\footnote{\url{https://github.com/rvaser/ra}}~\cite{vaser} was used and obtained consensus sequence is compared to the reference using dnadiff present in the Mumer\footnote{\url{https://github.com/garviz/MUMmer}}.
@@
 
-In this paper we used CNN instead of already tried RNN or HMM approaches, which yielded higher accuracy than previously tried models.
-Unlike HMM and RNN, there's no explicit dependency on previous hidden state, therefore this model is massively parallizible and more sensible given data nature --- that is we're dealing with signal processing, not heavily context depended language modeling.
+In this paper, we used CNN instead of already tried RNN or HMM approaches, which resulted in in higher accuracy compared to other existing basecallers.
+Unlike HMM and RNN, there's no explicit dependency on previous hidden state, therefore this model is massively parallelizable and more sensible given data nature --- that is we're dealing with signal processing, not heavily context-depended language modeling.
 
@@
 It can be used as a demonstration of a different approach to base calling which yields promising results.
-As newer versions of basecallers by Oxford Nanopore do not offer any support for data sequenced with previous version of chemistries, this tool can be used to re-basecall that data and improvement of the quality of reads retrospectively.
+As newer versions of basecallers by Oxford Nanopore do not offer any support for data sequenced with the previous version of chemistries, this tool can be used to re-basecall that data and improvement of the quality of reads retrospectively.
 
-Future work includes experiments with recently proposed Scaled exponential linear units (SELU)~\cite{selu} that eliminates the need for normalization techniques such as used batch normalization.
+Future work includes experiments with recently proposed Scaled exponential linear units (SELU)~\cite{selu} that eliminate the need for normalization techniques such as used batch normalization.
 Possible improvements of the model include the combination of convolutions and attention mechanism proposed just recently in the paper~\cite{facebook} showing excellent results in both speed and accuracy, for tasks of language translation.
-Last approach could use stacked simpled models, such as logistic regression and SVM to predict each nucleotide given raw signal context, similar to our deep learning model pre-CTC layer, and use linear chain CRF for full sequence basecalling.
-Approaches are many, and the time is limited.
+The other option could usage of stacked simple models, such as logistic regression and SVM to predict each nucleotide given raw signal context, similar to our deep learning model pre-CTC layer, and use linear chain CRF for full sequence basecalling.
 
 \section{Acknowledgments}
+First and foremost I'd like to thank my mentor izv.~prof.~dr.~sc.~Mile Šikić for setting up the problem, guiding us through its beginnings and providing helpful advice for its completion.
+Next Marko Ratković for his not only insightful conversations but also meaningful code contributions.
 
-First and foremost I'd like to thank my mentor izv.~prof.~dr.~sc.~Mile Sikić for setting up the problem, guiding us through its beginnings and providing helpful advice for its completion.
-Next Marko Ratkovic for his not only insightful conversations, but also meaningful code contributions.
+Finally, I'm thanking various other people whose code, tools and advice I've used in completing this paper: Fran Jurišić, Ana Marija Selak, Ivan Sović, Robert Vaser and Martin Šošić.
 
-Finally, I'm thanking various other people whose code, tools and advice I've used in completing this paper: Fran Jurišić, Ana Marija Selak, Ivan Sović and Martin Šošić.
-
-Also some data were obtained in cooperation with Biologists doc.~dr.~sc. Petra Korać and dr.sc. Paula Dobrinić. Other were procured publicly shared from Loman labs~\cite{loman1-100k}
+Also, some data were obtained in cooperation with Biologists doc.~dr.~sc. Petra Korać and dr.sc. Paula Dobrinić. Other were procured publicly shared from Loman labs~\cite{loman1-100k}
 

### 2017-07-01 paper/main.tex
Replaced "base call" with basecall

@@
 This is not only the case for the developed basecaller, but all other show the same property.
-This could be due to lack of context information from both sides when edges are base called.
+This could be due to lack of context information from both sides when edges are basecalled.
 
@@
 Currently, without support for newer sequencing data, this model has limited application.
-It can be used as a demonstration of a different approach to base calling which yields promising results.
+It can be used as a demonstration of a different approach to basecalling which yields promising results.
 As newer versions of basecallers by Oxford Nanopore do not offer any support for data sequenced with the previous version of chemistries, this tool can be used to re-basecall that data and improvement of the quality of reads retrospectively.
@@
 \emph{Metrichor} is an Oxford Nanopore company that offers cloud-based platform \emph{EPI2ME} for analysis of nanopore data.
-Initially, base calling was only available by uploading data to the platform - that being the reason why this basecaller is often called Metrichor even though it is a name of the company.
+Initially, basecalling was only available by uploading data to the platform - that being the reason why this basecaller is often called Metrichor even though it is a name of the company.
 
@@
 Similar to Nanonet, it is the platform for ongoing development.
-Scrappie is reported to be the first basecaller  that specifically address homopolymer base calling.
+Scrappie is reported to be the first basecaller  that specifically address homopolymer basecalling.
 It became publicly available just recently in June, 2017 and supports R9.4 and future R9.5 data.

### 2017-07-01 paper/main.tex
Normalized E. Coli

@@
 \begin{table}[]
-	\caption{Alignment specifications of Ecoli R9 basecalled reads using GraphMap}
+	\caption{Alignment specifications of E. Coli R9 basecalled reads using GraphMap}
 	\label{tbl:ecoli_rates}
@@
 		\includegraphics[width=0.8\textwidth]{./imgs/results/ecoli/kde_cigar.png}
-		\caption{KDE plot for distribution of percentage of alimnment operations for ecoli}
+		\caption{KDE plot for distribution of percentage of alimnment operations for E. Coli}
 		\label{fg:ecoli_kde}
@@
 \begin{table}[]
-	\caption{Consensus specifications of Ecoli R9 basecalled reads}
+	\caption{Consensus specifications of E. Coli R9 basecalled reads}
 	\label{tbl:spec_ecoli}
@@
 Average identity summarizes how closely does the assembled sequence match the reference.
-This is run on full ecoli sequence run for 1D template reads (\textasciitilde 160k reads), for both developed tools, Nanonet and Metrichor.
+This is run on full E.
+Coli sequence run for 1D template reads (\textasciitilde 160k reads), for both developed tools, Nanonet and Metrichor.
 Developed tool has shown a small increase in quality of the assembled sequence over Metrichor by offering longer consensus, higher identity percentage, and overall smaller edit distance\footnote{Calculated using \url{https://github.com/isovic/racon/blob/master/scripts/edcontigs.py}}.
@@
 \begin{table}[htb]
-	\caption{Assembly and consensus results for ecoli}
+	\caption{Assembly and consensus results for E. Coli}
 	\label{tbl:assembly}
@@
 \begin{table}[]
-	\caption{Alignment specifications of Ecoli R9 basecalled reads using BWA mem}
+	\caption{Alignment specifications of E. Coli R9 basecalled reads using BWA mem}
 	\label{tbl:ecoli_rates_bwa}

### 2017-07-01 paper/main.tex
Rewritten Acknowledgments
* Added Sigopt
* Added NVIDIA Corporation for GPU donations
* Removed co-authors from acknowledgment section

@@
 We also tried different channel numbers, receptive field width, and various other hyperparameters during hyperparameter optimization and conclude after enough complexity, that is sufficient layers, all choices were performing similarly.
+We used SigOpt~\cite{dewancker2016strategy}\footnote{\url{https://sigopt.com/}} bayesian hyperoptimization library.
 
@@
 \section{Acknowledgments}
-First and foremost I'd like to thank my mentor izv.~prof.~dr.~sc.~Mile Šikić for setting up the problem, guiding us through its beginnings and providing helpful advice for its completion.
-Next Marko Ratković for his not only insightful conversations but also meaningful code contributions.
 
-Finally, I'm thanking various other people whose code, tools and advice I've used in completing this paper: Fran Jurišić, Ana Marija Selak, Ivan Sović, Robert Vaser and Martin Šošić.
+During this paper creation we used Sigopt academic licence for hyperparameter optimization.
+We gratefully acknowledge the support of NVIDIA Corporation with the donation of the Titan Xp GPU used for this research.
 
-Also, some data were obtained in cooperation with Biologists doc.~dr.~sc. Petra Korać and dr.sc. Paula Dobrinić. Other were procured publicly shared from Loman labs~\cite{loman1-100k}
+We're thankful to various other people whose code, tools and advice we've used in completing this project: Fran Jurišić, Ana Marija Selak, Ivan Sović, Robert Vaser and Martin Šošić.
+
+For data, we're thankful for cooperation of Biologists doc.~dr.~sc. Petra Korać and dr.sc. Paula Dobrinić and Loman labs~\cite{loman1-100k} for publicly posting their dataset.
 

### 2017-07-01 paper/main.tex
Passed through grammar check

@@
 
-    The best performing deep learning model achieves state-of-the-art median 91.4\% match rate on E.Coli dataset using R9 pore chemistry and 1D reads.
+    The best performing deep learning model achieves state-of-the-art  91.4\% median match rate on E.Coli dataset using R9 pore chemistry and 1D reads.
 \keywords{Basecaller, MinION, R9, CNN, CTC, Next generation sequecing}
@@
 This has been somewhat alleviated with new R9 pore model, replacing older R7 ones.
-In this paper, we show that this error rate can be reduced by our approach with properly trained neural network model.
+In this paper, we show that this error rate can be reduced by our approach with the properly trained neural network model.
 
 The exact error rate metric is unreliable since multiple pipeline tools could be the issue.
-First the sample is prepared, hopefully, uncontaminated and matching reference genome as close as possible, then sequenced using the MinION device obtaining raw data.
+First the sample is prepared, hopefully, uncontaminated and matching reference genome as close as possible then sequenced using the MinION device obtaining raw data.
 Next, our model (or any other) comes along and basecalls the sequences.
@@
 
-We compared our model on R9 chemistry with Metrichor (HMM based approach), Nanonet and DeepNano (RNN based approaches). For detailed basecaller overview see the appendix~\ref{app:basecallers}.
+We compared our model on R9 chemistry with Metrichor (HMM-based approach), Nanonet and DeepNano (RNN based approaches). For the detailed basecaller overview see the appendix~\ref{app:basecallers}.
 
@@
 sc.
-Petra Korać i dr.sc.
+Petra Korać and dr.sc.
 Paula Dobrinić}.
@@
 Instead of opting for the traditional path using HMM or newly adopted RNN we tried using CNN~(Convolutional neural networks)~\cite{lecun-98}, that is their residual version~\cite{he2016deep}.
-For loss, we used CTC (Connectionist temporal classification)~\cite{graves2006connectionist} between basecalled and the target sequence.
+For loss, we used CTC (Connectionist Temporal Classification)~\cite{graves2006connectionist} between basecalled and the target sequence.
 Other building blocks used are Batch normalization(BN)~\cite{BNORM} and pooling layers.
@@
 We also tried different channel numbers, receptive field width, and various other hyperparameters during hyperparameter optimization and conclude after enough complexity, that is sufficient layers, all choices were performing similarly.
-We used SigOpt~\cite{dewancker2016strategy}\footnote{\url{https://sigopt.com/}} bayesian hyperoptimization library.
+We used SigOpt~\cite{dewancker2016strategy}\footnote{\url{https://sigopt.com/}} bayesian hyperparameter optimization library.
 
@@
 It is shown that mismatches and insertion occur more frequently at the beginnings and the ends of the reads.
-This is not only the case for the developed basecaller, but all other show the same property.
+This is not only the case for the developed basecaller, but all other shows the same property.
 This could be due to lack of context information from both sides when edges are basecalled.
@@
 Results are shown in table~\ref{tbl:spec_ecoli}.
-Lambda table is deferred to appendix in table~\ref{tbl:spec_lambda}.
+Lambda table is deferred to Appendix in table~\ref{tbl:spec_lambda}.
 
@@
 
-During this paper creation we used Sigopt academic licence for hyperparameter optimization.
+During this paper creation, we used Sigopt academic license for hyperparameter optimization.
 We gratefully acknowledge the support of NVIDIA Corporation with the donation of the Titan Xp GPU used for this research.
@@
 
-For data, we're thankful for cooperation of Biologists doc.~dr.~sc. Petra Korać and dr.sc. Paula Dobrinić and Loman labs~\cite{loman1-100k} for publicly posting their dataset.
+For data, we're thankful for the cooperation of Biologists doc.~dr.~sc. Petra Korać and dr.sc. Paula Dobrinić and Loman labs~\cite{loman1-100k} for publicly posting their dataset.
 
@@
 With the release of R9 chemistry, this model was replaced by a more accurate recurrent neural network (RNN) implementation.
-Currently, Oxford Nanopore offers several RNN-based local basecaller versions under different names: Albacore, Nanonet and basecaller integrated into MinKNOW \cite{ont-basecallers}.
+Currently, Oxford Nanopore offers several RNN-based local basecaller versions under different names: Albacore, Nanonet, and basecaller integrated into MinKNOW \cite{ont-basecallers}.
 
@@
 The source code of Albacore was not provided and is only available through the ONT Developer Channel.
-Tool supports only R9.4 and future R9.5 version of the chemistry.
-For R9 tests in this paper we used R9.4 chemisty setting as instructed on ONT forums.
+The tool supports only R9.4 and future R9.5 version of the chemistry.
+For R9 tests in this paper, we used R9.4 chemistry setting as instructed on ONT forums.
 
 \emph{Nanonet}\footnote{\url{https://github.com/nanoporetech/nanonet/}} uses the same neural network that is used in Albacore but it is continually under development and does contain features such as error handling or logging needed for production use.
-It uses \emph{CURRENNT} library for running neural networks.
-It supportes basecalling of both R9 and R9.4 chemistry versions.
-However in our experiments it was painfully slow, which was as expected due to it's classification as not production ready.
+It uses the \emph{CURRENNT} library for running neural networks.
+It supports basecalling of both R9 and R9.4 chemistry versions.
+However, in our experiments it was painfully slow, which was as expected due to it's classification as not production ready.
 
@@
 Scrappie is reported to be the first basecaller  that specifically address homopolymer basecalling.
-It became publicly available just recently in June, 2017 and supports R9.4 and future R9.5 data.
+It became publicly available just recently in June 2017 and supports R9.4 and future R9.5 data.
 

### 2017-07-01 paper/main.tex
Changed title
Minor fixes

@@
 
-\title{MinCall --- MinION end2end deep learning basecaller}
+\title{MinCall --- MinION end2end convolutional deep learning basecaller}
 \author{Neven Miculinić \and Marko Ratković \and Mile Sikić\\
@@
 
-\section{Sequencing overview}
+\subsection{Sequencing overview}
 Conceptually, the MinION sequencer works as follows.
@@
 
-\section{Basecalling}
+\section{Related work}
 The core of the decoding process is the basecalling step, that is translating the current samples to the nucleotide sequence.
@@
 
-Earlier models were Hidden Markov model(HMM)-based where hidden state modeled DNA sequence of length 6 (6-mer) in the nanopore.
+Earlier models were Hidden Markov model (HMM)-based where hidden state modeled DNA sequence of length 6 (6-mer) in the nanopore.
 Pore models were used in computing emission probabilities.~\cite{loman2015complete,schreiber2015analysis,szalay2015novo,timp2012dna} and the recent open source HMM-based basecaller Nanocall~\cite{david2016nanocall}.
@@
 
-We compared our model on R9 chemistry with Metrichor (HMM-based approach), Nanonet and DeepNano (RNN based approaches). For the detailed basecaller overview see the appendix~\ref{app:basecallers}.
+We compared our model on R9 chemistry with Metrichor (HMM-based approach), Nanonet\footnote{\url{https://github.com/nanoporetech/nanonet/}} and DeepNano~\cite{deepnano} (RNN based approaches).
+For the detailed basecaller overview see the appendix~\ref{app:basecallers}.
 
@@
 
-\emph{DeepNano}~\cite{Boza2017}  was the first open-source basecaller based on neural networks.
+\emph{DeepNano}~\cite{deepnano}  was the first open-source basecaller based on neural networks.
 It uses bidirectional recurrent neural networks implemented in Python, using the Theano library.

### 2017-07-01 paper/main.tex
Reorganised few stuff

@@
 Oxford Nanopore Technology's MinION nanopore sequencing platform~\cite{mikheyev2014first} is the first portable DNA sequencing device.
-It's small weight, of only 90 grams, low capital cost, and long read length combined with high-throughput, real-time data analysis, and decent accuracy yield promising results in various applications.
+It's light, only 90 grams.
+It's has low capital cost.
+It's featuring long read length combined with high-throughput, real-time data analysis, and decent accuracy.
+Therefore it yields promising results in various applications.
 From clinical application such as monitoring infectious disease outbreaks~\cite{judge2015early}\cite{quick2016real}, characterizing structural variants in cancer\cite{norris2016nanopore} and even full human genome assembly~\cite{jain2017nanopore}.
@@
 In this paper, we show that this error rate can be reduced by our approach with the properly trained neural network model.
-
-The exact error rate metric is unreliable since multiple pipeline tools could be the issue.
-First the sample is prepared, hopefully, uncontaminated and matching reference genome as close as possible then sequenced using the MinION device obtaining raw data.
-Next, our model (or any other) comes along and basecalls the sequences.
-To evaluate error rate metric basecalled read is aligned to the reference genome using aligners with their own errors/biases, most commonly used BWA-MEM~\cite{li2013aligning} and Graphmap~\cite{sovic2016fast}.
 
@@
 
-\section{Related work}
+\subsection{Related work}
 The core of the decoding process is the basecalling step, that is translating the current samples to the nucleotide sequence.
@@
 Using this information output given by Metrichor can be determined for each block.
+
 To correct errors produced by Metrichor and possibly increase the quality of data, each read is aligned to the reference.
 This is done using aligner GraphMap~\cite{sovic2016fast} that returns the best position in the genome, hopefully, the part of the genome from which read came from.
+
 Alignment part in the genome is used as a target.
@@
 This includes third-party basically DeepNano and official basecallers by Oxford Nanopore (cloud-based Metrichor and Nanonet).
-The fact that ground truth is not known makes evaluation difficult.
-Different methods for evaluation were used to get clearer information about each basecaller.
+
+The exact error rate metric is unreliable since multiple pipeline tools could be the issue.
+First the sample is prepared, hopefully, uncontaminated and matching reference genome as close as possible then sequenced using the MinION device obtaining raw data.
+Next, our model (or any other) comes along and basecalls the sequences.
+To evaluate error rate metric basecalled read is aligned to the reference genome using aligners with their own errors/biases.
+We used standard BWA-MEM~\cite{li2013aligning} and Graphmap~\cite{sovic2016fast}.
+
+The fact that ground truth is not known makes evaluation difficult. Different methods for evaluation were used to get clearer information about each basecaller.
 

### 2017-07-01 paper/main.tex
Added basecall speeds and removed input pipeline figure

@@
 We compared our model on R9 chemistry with Metrichor (HMM-based approach), Nanonet\footnote{\url{https://github.com/nanoporetech/nanonet/}} and DeepNano~\cite{deepnano} (RNN based approaches).
-For the detailed basecaller overview see the appendix~\ref{app:basecallers}.
+For the detailed basecaller overview see the Appendix~\ref{app:basecallers}.
 
@@
 
-Dataset used were E.Coli K-12 strands from~\cite{loman1-100k} and Lamba basecalling\footnote{Acquired from doc.
-dr.
-sc.
+Dataset used were E.Coli K-12 strands from~\cite{loman1-100k} and Lamba basecalling\footnote{Acquired from doc.~dr.~sc.
 Petra Korać and dr.sc.
 Paula Dobrinić}.
-Both used datasets show in table \ref{tbl:datasets} have been previously passed through MinKNOW and had been basecalled by Metrichor.
+Both used datasets show in table~\ref{tbl:datasets} have been previously passed through MinKNOW and had been basecalled by Metrichor.
 As 1D read analysis was the focus of this paper, only those reads were used.
@@
 Total train set consist of over 110 thousand reads.
-Overview of the entire learning pipeline is shown in figure~\ref{fg:train_pipe}.
 
-\begin{figure}[!ht]
-    \begin{center}
-        \includegraphics[width=0.7\textwidth]{./imgs/train_pipeline.png}
-        \caption{Training pipeline overview}
-        \label{fg:train_pipe}
-    \end{center}
-\end{figure}
+% Overview of the entire learning pipeline is shown in figure~\ref{fg:train_pipe}.
+% \begin{figure}[!ht]
+%     \begin{center}
+%         \includegraphics[width=0.7\textwidth]{./imgs/train_pipeline.png}
+%         \caption{Training pipeline overview}
+%         \label{fg:train_pipe}
+%     \end{center}
+% \end{figure}
 
@@
 
-The fact that ground truth is not known makes evaluation difficult. Different methods for evaluation were used to get clearer information about each basecaller.
+The fact that ground truth is not known makes evaluation difficult.
+Different methods for evaluation were used to get clearer information about each basecaller.
+In the following subsections are different evaluation metrics described and analysed.
+For completeness speed measurements are given in Table~\ref{tbl:speeds}
+
+\begin{table}[]
+	\caption{Base calling speeds measured in \textit{base pairs per second}. CPU is Intel Xeon E5-2640 v2 @ 2Ghz. GPU is NVIDIA Titan X Black}
+	\label{tbl:speeds}
+	\centering
+
+	\begin{tabular}{lcc}
+		\toprule
+		{} &  \thead{ecoli (bp/s)} &  \thead{lambda (bp/s)} \\
+		\midrule
+		resdeep (CPU)  &               \textbf{1174.28} &              \textbf{ 1363.340 }\\
+		Nanonet (CPU)  &                856.01 &                897.499 \\
+		DeepNano (CPU) &                626.99 &                692.370 \\
+		\midrule
+		resdeep (GPU)  &              \textbf{ 6571.76 }&              \textbf{ 6140.300} \\
+		Nanonet (GPU)  &               3828.39 &               3787.510 \\
+		\bottomrule
+	\end{tabular}
+
+\end{table}
 
